---
title: "Getting started with anovakit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with anovakit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 3.5, dpi = 96)
library(anovakit)
```

One dataset, run through the whole package. Every function names its columns as
character strings and returns the same class, so once you have read one section
you have read them all. Six take `data, response, groups`; `anova_manova()`
takes `responses` (plural) and `anova_rm()` takes `subject`, `within` and
`between`, because those designs need more than a grouping variable to describe
them.

```{r data}
set.seed(2024)
n <- 240

trial <- data.frame(
  arm      = factor(rep(c("placebo", "low", "high"), each = n / 3)),
  site     = factor(rep(c("north", "south"), times = n / 2)),
  baseline = rnorm(n, mean = 100, sd = 12)
)

effect <- c(placebo = 0, low = 3, high = 7)[as.character(trial$arm)]
trial$score    <- 0.6 * trial$baseline + effect + rnorm(n, 0, 6)
trial$improved <- rbinom(n, 1, plogis(-1.2 + 0.25 * effect))
trial$visits   <- rpois(n, exp(1.1 + 0.06 * effect))
trial$weeks    <- runif(n, 4, 12)

str(trial)
```

## Comparing means

`anova_welch()` is the default choice for a one-way comparison. It does not
assume equal variances, which costs almost nothing when they are equal and
matters a great deal when they are not.

```{r welch}
fit <- anova_welch(trial, "score", "arm")
fit
```

`print()` shows the omnibus test and anything the function wants you to know.
`summary()` adds the rest.

```{r welch-summary}
fit$emmeans
fit$posthoc[, c("group1", "group2", "difference", "conf_low", "conf_high",
                "p_adjusted")]
fit$effect_sizes
```

Assumption checks are chosen to suit the method. Because Welch's test permits
unequal variances, normality is assessed within each group rather than on pooled
residuals:

```{r welch-assumptions}
fit$assumptions$normality
fit$assumptions$variance_ratio
```

Plots are built and returned, never drawn:

```{r welch-plot}
plot(fit, "means")
```

## Adjusting for a covariate

Baseline score is a strong predictor here, so an ANCOVA is the better analysis.

```{r ancova}
anc <- anova_ancova(trial, "score", "arm", "baseline", plots = FALSE)
anc$anova
anc$emmeans
```

Note what `$notes` says:

```{r ancova-notes}
anc$notes
```

Covariates are mean-centred by default. That matters as soon as the model
contains a group-by-covariate term: under sum-to-zero contrasts the Type III row
for `arm` then tests the group difference *at covariate = 0*, and `baseline` is
distributed around 100, so an uncentred fit would be comparing the arms at a
point nowhere near the data.

```{r ancova-centring}
p_of <- function(...) {
  f <- anova_ancova(trial, "score", "arm", "baseline", force_interaction = TRUE,
                    plots = FALSE, ...)
  f$anova$p_value[f$anova$term == "arm"]
}
c(centred = p_of(), uncentred = p_of(center_covariates = FALSE))
```

The additive model fitted above is not affected — there the group row is the
same either way, and what centring buys is that `$emmeans` and the intercept are
read at the covariate mean rather than at zero. Use `force_interaction` to fix
the model in advance rather than letting a test on the same data choose it.

## A binary outcome

```{r bin}
bin <- anova_bin(trial, "improved", "arm", plots = FALSE)
bin$anova
bin$effect_sizes
bin$emmeans
```

Odds-ratio intervals are profile-likelihood by default; `ci_method = "wald"`
switches them. Separation, which makes Wald intervals meaningless while `glm()`
still reports convergence, is detected and named in `$notes`.

```{r bin-separated}
sep <- trial
sep$improved[sep$arm == "high"] <- 1L
anova_bin(sep, "improved", "arm", plots = FALSE)$notes
```

## A count outcome, with exposure

`weeks` is time at risk, so it belongs in the model as a log offset. It enters
the formula rather than `glm()`'s `offset` argument, which is what lets
**emmeans** see it: the marginal means below are rates per week.

```{r count}
cnt <- anova_count(trial, "visits", "arm", offset = "weeks", plots = FALSE)
cnt$model_type
cnt$dispersion        # Pearson dispersion of the Poisson fit
cnt$model_dispersion  # ... and of the model actually returned
cnt$emmeans
```

If the Pearson dispersion had exceeded the threshold, the function would have
moved to a negative binomial model and said so in `$notes`. Set `model` to take
that decision yourself.

## Two grouping variables

Every function accepts several grouping variables. They are combined into a
single cell factor, and level combinations that never occur are dropped, so the
number of groups reported is the number that contain data.

```{r twoway}
tw <- anova_glm(trial, "score", c("arm", "site"), interaction = TRUE,
                type = "III", plots = FALSE)
tw$anova
```

When you ask for `type = "III"` the model is *fitted* under sum-to-zero
contrasts. Setting `options(contrasts = )` after a model exists does not change
the contrasts stored on it, and the resulting table would be simple effects at
the reference level wearing a Type III label.

```{r contrasts}
unlist(tw$model$contrasts)
```

## Several outcomes at once

```{r manova}
mv <- anova_manova(trial, c("score", "visits"), "arm",
                   covariates = "baseline", plots = FALSE)
mv$multivariate
mv$assumptions$mardia
mv$assumptions$box_m
mv$canonical
mv$canonical_term
```

Mardia's tests, Box's M and the canonical discriminant analysis are computed
directly, so none of them adds a dependency. With covariates present, all three
are computed on the responses with the covariate effect removed — the space the
multivariate test actually operates in.

The follow-up univariate analyses are kept individually as well as stacked, so
each response has its own model and its own emmeans grid:

```{r manova-univariate}
mv$univariate$score$anova
```

## When the data are not normal

```{r kw}
kw <- anova_kw(trial, "score", "arm", plots = FALSE)
kw$anova
kw$posthoc
```

Dunn's test is computed from the rank sums with the standard tie correction, so
any `p.adjust()` method may be used and nothing is printed to the console.
Kruskal-Wallis assumes no distribution, so no assumption check runs unless you
ask for one with `diagnostics = TRUE`.

## Repeated measures

```{r rm, eval = requireNamespace("afex", quietly = TRUE)}
set.seed(11)
long <- expand.grid(id = factor(1:40), week = factor(c("w0", "w4", "w8")))
long$arm <- factor(rep(rep(c("placebo", "active"), each = 20), 3))
long$score <- 50 + 3 * as.numeric(long$week) +
  2 * (long$arm == "active") + rnorm(nrow(long), 0, 4)

rm_fit <- anova_rm(long, "score", subject = "id", within = "week",
                   between = "arm", plots = FALSE)
rm_fit$anova
rm_fit$sphericity
```

Sphericity comes from the fitted model, along with the Greenhouse-Geisser and
Huynh-Feldt corrections. Subjects missing a within-subject cell are removed
before fitting and listed in `$subjects_dropped` — empty above, because that
design is complete:

```{r rm-dropped, eval = requireNamespace("afex", quietly = TRUE)}
gappy <- long[!(long$id %in% c("1", "2") & long$week == "w4"), ]
gap_fit <- anova_rm(gappy, "score", subject = "id", within = "week",
                    between = "arm", plots = FALSE)
gap_fit$subjects_dropped
c(used = nrow(gap_fit$data_used), removed = gap_fit$n_removed,
  supplied = nrow(gappy))
```

## Two arguments worth knowing about

`vcov_type` gives heteroskedasticity-consistent standard errors, and they flow
through to the marginal means and the comparisons rather than stopping at the
coefficient table:

```{r robust}
anova_glm(trial, "score", "arm", vcov_type = "HC3", plots = FALSE)$emmeans$se
anova_glm(trial, "score", "arm", plots = FALSE)$emmeans$se
```

`weights` is a column name rather than a vector, so prior weights are subsetted
along with the data and cannot misalign when a row is dropped:

```{r weights}
trial$wt <- runif(n, 0.5, 2)
anova_glm(trial, "score", "arm", weights = "wt", plots = FALSE)$anova
```

## The habit worth forming

Read `$notes` on every fit. It is where the package records what it decided on
your behalf, what it could not compute, and what it thinks you should know
before you quote a number.

```{r notes}
cnt$notes
```

## Where to go next

`?anovakit_fit` documents every component of the returned object, including
the ones each function adds for itself. `names(fit)` lists them for any fit you
have in hand.

```{r next}
names(mv)
```

Each function's own help page covers the arguments this vignette did not reach —
`reference` and `success` on `anova_bin()`, `model` and
`overdispersion_threshold` on `anova_count()`, `correction` and `emm_specs` on
`anova_rm()`, `test` and `assumptions` on `anova_manova()`. All of them are
optional, and all of them have defaults the functions record in `$notes` when
they matter.
