---
title: "ANCOVA: comparing groups at the same value of a covariate"
description: >
  A walkthrough of anova_ancova() on the anorexia treatment data: covariate-adjusted
  means, the homogeneity-of-slopes test, simple slopes, centring and robust tests.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

`anova_ancova()` compares the mean of a numeric response across groups while
adjusting for one or more numeric covariates. Use it when something you measured
before the groups could differ -- a baseline score, a starting weight -- also
predicts the outcome: adjusting for it removes that source of noise and compares
the groups at the same value of the covariate. With no covariate, use
`anova_welch()` or `anova_glm()`; with several responses, `anova_manova()`
accepts covariates too. The [decision guide](choosing.html) covers the choice in
more detail.

## The data

`MASS::anorexia` records the weight (in pounds) of 72 young women with anorexia
before and after a period of treatment: cognitive behavioural treatment (`CBT`),
family treatment (`FT`) or a control condition (`Cont`).

```{r data}
an <- MASS::anorexia
head(an)
table(an$Treat)
aggregate(cbind(Prewt, Postwt) ~ Treat, data = an, FUN = mean)
```

```{r data-cor, include = FALSE}
r_pre_post <- cor(an$Prewt, an$Postwt)
```

The groups start at slightly different weights, and a patient's weight
afterwards depends partly on where she started (the correlation between the
two is `r round(r_pre_post, 2)`). The question an ANCOVA answers is: *for patients of
the same starting weight*, does the treatment make a difference to the weight
afterwards? `Postwt` is the response, `Treat` the grouping variable and `Prewt`
the covariate.

## Fitting the model

Columns are named as character strings: the response, then the grouping
variable(s), then the covariate(s).

```{r fit}
fit <- anova_ancova(an, "Postwt", "Treat", "Prewt")
fit
```

Reading the printout from the top:

- The method line says the tests use Type III sums of squares (the default
  `type = "III"`, fitted under sum-to-zero contrasts).
- `Observations used` is the number of rows in `$data_used`. Rows with a missing
  value in any analysed column are dropped and counted in `$n_removed`; there
  are none here.
- The omnibus table has a `Prewt:Treat` row because the function fitted two
  candidate models and a homogeneity-of-slopes test chose the one with a
  separate slope for each group (see the next section).
- The notes say what the function decided and why. Read them on every fit; they
  are discussed [below](#what-the-notes-say).
- `Plots available` lists the ggplot2 objects in `$plots`.

`summary(fit)` prints all of this and then the assumption checks, effect
sizes, marginal means, pairwise comparisons, the slopes test and the per-group
slopes. The rest of this article takes those components one at a time.

## The homogeneity-of-slopes test

Classical ANCOVA assumes that the covariate has the same slope in every group,
so that the groups differ by a constant amount whatever the covariate value.
`anova_ancova()` checks this before anything else. It fits two candidate
models -- the covariate entered additively, and the same model plus the
product of the covariate with the grouping terms -- and compares them with one
nested F test:

```{r slopes}
fit$slopes_test
```

Here the test rejects (p = `r signif(fit$slopes_test$p_value, 3)`), so
`homogeneous` is `FALSE` and the interaction model was fitted. Both candidates
are kept, so the one not chosen is still available, and the comparison can be
reproduced from them:

```{r candidate-models}
deparse(formula(fit$model_additive))
deparse(formula(fit$model_interaction))
anova(fit$model_additive, fit$model_interaction)
```

When the slopes differ, the thing to read is the slope in each group,
`$simple_slopes` (from `emmeans::emtrends()`):

```{r simple-slopes}
print(fit$simple_slopes[, c("Treat", "slope", "se", "conf_low", "conf_high",
                            "p_value")], digits = 3)
```

```{r slopes-inline, include = FALSE}
ss <- setNames(fit$simple_slopes$slope, fit$simple_slopes$Treat)
```

In both active treatments a heavier start goes with a heavier finish (about
`r round(ss[["CBT"]], 2)` and `r round(ss[["FT"]], 2)` lb per lb), but in the
control group there is essentially no relation (`r round(ss[["Cont"]], 2)`, with
an interval that spans zero). The treatment effect is therefore not one number:
it depends on the starting weight. `$simple_slopes` is `NULL` whenever the
additive model is used, because there is then only one slope.

### Why `force_interaction` exists

Letting a test on the same data choose the model has a cost in both directions,
and the notes say so. When the test retains the interaction, as here, the
p-values that follow do not allow for that choice and can be too small. When
the slopes really differ but the test misses it, the additive model's adjusted
comparisons are biased by the slope difference times the difference in
covariate means -- which averages out when groups are randomised, but not
otherwise. The help page (`?anova_ancova`) reports simulations of how often
that second case rejects a true null hypothesis.

`force_interaction` removes the selection step. Decide in advance:

- `force_interaction = TRUE` always fits the separate-slopes model. With centred
  covariates and Type III tests, the group row is then a valid comparison of
  the groups at the covariate mean whether or not the slopes differ. This is
  the safer choice when the groups were not randomised.
- `force_interaction = FALSE` always fits the additive model.
- `NULL` (the default) lets the slopes test decide, at level
  `homogeneity_alpha` (0.05 by default).

```{r force}
fixed <- anova_ancova(an, "Postwt", "Treat", "Prewt", force_interaction = TRUE,
                      plots = FALSE)
all.equal(fixed$anova, fit$anova)
fixed$notes[2]
```

The model and the table are the same as before -- the test happened to choose
the same model -- but the notes now say the choice was yours, and the warning
about selection is gone.

## The omnibus test

```{r anova}
fit$anova
```

Each row is a term of the model:

- `sum_sq` is its Type III sum of squares: what the term adds when every other
  term is already in the model. `Residuals` is the error sum of squares.
- `df` is its degrees of freedom (a three-level factor has 2).
- `statistic` is the F ratio, the term's mean square over the residual mean
  square, and `p_value` its upper-tail probability on `df` and the residual
  degrees of freedom.

The `Treat` row is the one most people want, and in this model it needs care.
Because the covariate is centred and the slopes are separate, it compares the
groups *at the mean starting weight* (`r round(fit$covariate_means[["Prewt"]], 1)` lb).
It is not a constant treatment effect, since there is none when the slopes
differ. `Prewt` tests the covariate's slope averaged over the groups, and
`Prewt:Treat` is the slope difference the homogeneity test found (with the same
F and p-value).

## Effect sizes

```{r effect-sizes}
fit$effect_sizes
```

`partial_eta_sq` is the term's sum of squares over itself plus the residual sum
of squares: the share of the variation not explained by the other terms that
this term explains. `partial_omega_sq` corrects that for the bias of a sample
estimate and is the better estimate of the population value; it is floored at
0 (with a note) when a term explains less than chance would. Both are
*partial* measures, each with its own denominator, and Type III sums of squares
do not add up to the total, so do not add the rows together. Neither comes with
an interval.

## Covariate centring

Every covariate is mean-centred before fitting (`center_covariates = TRUE`). The
means subtracted are in `$covariate_means`, and `$data_used` holds the centred
values:

```{r centring}
fit$covariate_means
head(fit$data_used, 3)
```

What centring changes depends on the model. It always moves the intercept. When
the covariate-by-group terms are in the model and `type = "III"`, it also moves
the group row, because that row then tests the groups *at covariate = 0*.
Without centring, zero pounds is a starting weight none of these patients had:

```{r uncentred}
unc <- anova_ancova(an, "Postwt", "Treat", "Prewt", center_covariates = FALSE,
                    plots = FALSE)
data.frame(term = fit$anova$term,
           p_centred = fit$anova$p_value,
           p_uncentred = unc$anova$p_value)
all.equal(fit$emmeans, unc$emmeans)
```

Only the `Treat` row changed. The covariate row, the interaction row and the
adjusted means are identical, because `$emmeans`, `$posthoc` and
`$simple_slopes` are always evaluated with every covariate at its mean. In the
additive model, and under Type II, centring does not change the group row at
all. Set `center_covariates = FALSE` only if the covariate's own zero is the
point you want the intercept to describe; the notes of `unc` explain what it
did.

## Marginal means and comparisons

```{r emmeans}
fit$emmeans
```

These are covariate-adjusted means: the predicted post-treatment weight in each
group for a patient starting at the mean weight of
`r round(fit$covariate_means[["Prewt"]], 2)` lb. They differ from the raw means
above because the groups did not start at exactly that weight.

```{r posthoc}
print(fit$posthoc[, c("contrast", "estimate", "conf_low", "conf_high",
                      "p_value", "p_adjusted", "adjustment")], digits = 3)
```

`p_value` is the unadjusted p-value of each comparison; `p_adjusted` is
adjusted for the three comparisons by the method in `adjustment` (Tukey by
default), and the intervals are adjusted to match. Report `p_adjusted` when
you are looking at all pairs. `adjust` takes any of the methods listed on the
help page, for example `adjust = "holm"`; `"none"` makes the two columns equal.

```{r holm}
holm <- anova_ancova(an, "Postwt", "Treat", "Prewt", adjust = "holm",
                     plots = FALSE)
print(holm$posthoc[, c("contrast", "p_value", "p_adjusted")], digits = 3)
```

The emmeans grid itself is in `$emmeans_object`, for any contrast the wrapper
does not compute (give `level =` when you summarise it; see
[Working with results](results.html)).

### What the adjustment buys

Compare this with two analyses that ignore the starting weight in different
ways. `anova_welch()` compares the raw post-treatment weights; an ANOVA on the
change scores (`Postwt - Prewt`) assumes that each pound of starting weight
carries over one for one, a slope of exactly 1.

```{r unadjusted}
welch <- anova_welch(an, "Postwt", "Treat", plots = FALSE)
print(welch$posthoc[, c("group1", "group2", "difference", "p_adjusted")],
      digits = 3)

an$change <- an$Postwt - an$Prewt
chg <- anova_glm(an, "change", "Treat", plots = FALSE)
print(chg$posthoc[, c("contrast", "estimate", "p_adjusted")], digits = 3)
```

```{r unadjusted-inline, include = FALSE}
p_anc <- fit$posthoc$p_adjusted[fit$posthoc$contrast == "CBT - Cont"]
p_chg <- chg$posthoc$p_adjusted[chg$posthoc$contrast == "CBT - Cont"]
pre_means <- tapply(an$Prewt, an$Treat, mean)
pre_gap <- pre_means[["CBT"]] - pre_means[["Cont"]]
```

All three find that family treatment beats the control, but they answer
different questions. Welch's test ignores the starting weight altogether: its
CBT-control difference of `r round(welch$posthoc$difference[1], 2)` lb makes no
allowance for the `r round(pre_gap, 2)` lb by which the CBT group started
heavier, and the variation the starting weight explains stays in the error. The change-score
analysis imposes a slope of 1, which the simple slopes above show no group has,
and with it the CBT-control comparison weakens to adjusted p =
`r signif(p_chg, 2)`, against `r signif(p_anc, 2)` in the ANCOVA. ANCOVA
estimates the slope from the data instead, which is why it is usually the
better choice for a pre-post design.

## Checking assumptions

```{r assumptions}
names(fit$assumptions)
fit$assumptions$normality
fit$assumptions$levene
```

- `normality` is a Shapiro-Wilk test on the model residuals. It tests the
  assumption that the errors are normal; a small p-value in a small sample
  deserves a look at the Q-Q plot, while in a large sample the test flags
  departures too small to matter. Here W =
  `r round(unname(fit$assumptions$normality$statistic), 3)` and p =
  `r round(fit$assumptions$normality$p.value, 2)`.
- `levene` is Levene's test (on absolute deviations from the medians) of the
  residuals across the groups (the cells, with several grouping variables). A
  rejection means the pooled residual variance is not a fair summary; the note
  it triggers suggests `vcov_type = "HC3"`, shown [below](#robust-tests-with-vcov_type).
  Here p = `r round(fit$assumptions$levene$p_value, 2)`.
- `slopes` is the homogeneity-of-slopes test, the same table as
  `$slopes_test`.

A covariate that is constant, constant within every group, or an exact linear
combination of the groups and the other covariates is refused with an error
rather than fitted, since there is nothing to adjust for. The
[diagnostics article](diagnostics.html) covers the checks across the package.

## Plots

Every plot is a ggplot2 object in `$plots`; `plot(fit, "name")` returns one, and
you can add layers or themes to it with `+` (see [Plots and visual
customisation](visuals.html)).

### Residuals against fitted values

```{r plot-residuals}
plot(fit, "residuals")
```

What to look for: an even band around zero. A funnel (spread growing with the
fitted value) points to unequal variances; a curve points to a covariate
relation that is not linear. Here there is neither. The tight column of
points at fitted values of about 80 to 82 lb is the control group: with a
nearly flat slope, its fitted values hardly vary.

### Normal Q-Q plot

```{r plot-qq}
plot(fit, "qq")
```

What to look for: points close to the dashed line. Systematic bends at the ends
mean heavy or light tails; a single point far off it is an outlier worth
checking in the data. These residuals follow the line with a mild S-shape,
which is in keeping with the Shapiro-Wilk test above.

### The covariate plot

```{r plot-covariate}
plot(fit, "covariate")
```

This is the plot to look at first in any ANCOVA. It shows the response against
the covariate *on its original scale*, with a separate least-squares line and
confidence band for each group (the bands use `conf_level`). The lines are each
group's own fit, drawn whichever model was chosen, so they show you directly
whether parallel lines are a fair description. Here they are not: the control
group's line is almost flat while the two treatment lines rise. Look too at
whether the groups cover the same range of the covariate -- adjusted means for
a group observed only at one end are extrapolations, and `$notes` names such a
group. With several covariates there is one plot per covariate, named
`covariate_<name>`.

### Adjusted means

```{r plot-emmeans}
plot(fit, "emmeans")
```

The covariate-adjusted means from `$emmeans` with their confidence intervals.
With a separate-slopes model, remember that these are the group means at one
starting weight; the covariate plot shows how the gaps change away from it.

## Options worth knowing

### Robust tests with `vcov_type`

`vcov_type = "HC3"` (or `"HC0"` to `"HC4"`, which need the sandwich package)
replaces the model-based covariance with a heteroscedasticity-consistent one.

```{r robust}
rob <- anova_ancova(an, "Postwt", "Treat", "Prewt", force_interaction = TRUE,
                    vcov_type = "HC3", plots = FALSE)
rob$anova
attr(rob$anova, "statistic")
data.frame(Treat = fit$emmeans$Treat, se_model = fit$emmeans$se,
           se_HC3 = rob$emmeans$se)
```

The F tests in `$anova` are now Wald F tests built from the robust covariance
(through `car::Anova(vcov. = )`), on the model's residual degrees of freedom,
so the table has no sums of squares, and its `statistic` attribute (which
`print()` shows in the heading) says so. The same covariance reaches `$emmeans`,
`$posthoc` and `$simple_slopes`: the control group's standard error shrinks
and the family-treatment group's grows, reflecting their different residual
spreads. Two things do not change: `$effect_sizes` still come from the
model-based sums of squares, and the homogeneity-of-slopes test stays the
model-based F test. The notes of `rob` say both.

### Several grouping variables: `interaction`

With two or more grouping variables, the default `interaction = TRUE` fits
their full factorial, which is what a factorial ANCOVA is. `MASS::birthwt`
has birth weights (grams) by the mother's race and smoking status, with the
mother's weight (pounds) as a covariate.

```{r birthwt}
bw <- MASS::birthwt
bw$race <- factor(bw$race, labels = c("white", "black", "other"))
bw$smoke <- factor(bw$smoke, labels = c("no", "yes"))

two <- anova_ancova(bw, "bwt", c("race", "smoke"), "lwt", plots = FALSE)
two$anova
two$slopes_test
```

The slopes test now allows a separate slope in every cell (5 extra degrees of
freedom) and finds no evidence against a common one, so the additive-covariate
model is used. `$emmeans` and `$posthoc` compare the six cells:

```{r birthwt-emmeans}
print(two$emmeans, digits = 4)
```

With `interaction = FALSE` the grouping variables enter additively, and the
means and comparisons are reported for each factor separately, averaged over
the other, with a `term` column; each factor is its own multiplicity family.

```{r birthwt-additive}
add <- anova_ancova(bw, "bwt", c("race", "smoke"), "lwt", interaction = FALSE,
                    plots = FALSE)
print(add$posthoc[, c("term", "contrast", "estimate", "p_adjusted")],
      digits = 3)
```

A whole number keeps interactions up to that order. The notes of both fits
carry one more message worth reading:

```{r birthwt-notes}
two$notes[3]
```

```{r birthwt-gap, include = FALSE}
cell <- interaction(bw$race, bw$smoke)
sd_within <- sqrt(sum((bw$lwt - ave(bw$lwt, cell))^2) /
                    (nrow(bw) - nlevels(cell)))
imb_gap <- diff(range(tapply(bw$lwt, cell, mean))) / sd_within
```

The mothers' mean weights differ between the cells by up to
`r sprintf("%.2f", imb_gap)` pooled within-group SDs, so a slope difference
the test missed would bias the comparisons. Since smoking and race were not randomised,
this is the situation where `force_interaction = TRUE` is the safer choice.

### The rest

- `type = "II"` gives Type II sums of squares, where each term is adjusted for
  every other term that does not contain it. Under Type II the group row is
  the same whether or not the covariate is centred.
- `conf_level` sets the level of every interval: the marginal means, the
  comparisons, the simple slopes and the bands of the covariate plot.
- `homogeneity_alpha` is the level of the slopes test (default 0.05).
- `posthoc = FALSE` skips the pairwise comparisons, which is worth doing when
  there are many cells; a note records it.
- `plots = FALSE` skips building the plots, which saves time in a loop.

## What the notes say

```{r notes}
fit$notes
```

1. The covariate was centred at its mean, and so the Type III group row
   compares the groups at `Prewt` = `r round(fit$covariate_means[["Prewt"]], 2)`
   rather than at 0. The marginal means are at the mean either way.
2. The slopes test rejected, so the interaction model was kept, and the group
   row is a comparison at one point on the covariate: read `$simple_slopes`
   and `$emmeans`.
3. The model was chosen by a test on the same data, so the p-values do not
   account for that choice. `force_interaction = TRUE`, decided in advance,
   avoids it.

Other notes appear when they apply: an imbalance warning when the additive
model is used and the covariate means differ between groups by more than half a
pooled within-group SD (as in the birth-weight example), an extrapolation
warning when the covariate mean lies outside a group's observed range, a
Levene rejection, dropped rows, and anything `lm()`, car or emmeans said
while the model was being fitted.

## Reporting the result

```{r report-values, include = FALSE}
a <- fit$anova
row <- function(t) a[a$term == t, ]
tr <- row("Treat"); it <- row("Prewt:Treat")
df_res <- a$df[a$term == "Residuals"]
em <- setNames(fit$emmeans$estimate, fit$emmeans$Treat)
ph <- fit$posthoc
pc <- function(k) ph[ph$contrast == k, ]
fmt <- function(p) {
  if (p < 0.001) "p < 0.001" else paste("p =", format(signif(p, 2)))
}
f1 <- function(x) sprintf("%.1f", x)
es <- fit$effect_sizes
```

> Post-treatment weight was analysed by ANCOVA with pre-treatment weight
> (centred at its mean, `r round(fit$covariate_means[["Prewt"]], 1)` lb) as the
> covariate. The slope of pre-treatment weight differed between treatments,
> F(`r it$df`, `r df_res`) = `r round(it$statistic, 2)`, `r fmt(it$p_value)`:
> it was `r round(ss[["CBT"]], 2)` for CBT, `r round(ss[["FT"]], 2)` for family
> treatment and `r round(ss[["Cont"]], 2)` for the control group, so the model
> kept separate slopes. At the mean pre-treatment weight, post-treatment weight
> differed between treatments, F(`r tr$df`, `r df_res`) =
> `r round(tr$statistic, 2)`, `r fmt(tr$p_value)`, partial
> &eta;^2^ = `r round(es$partial_eta_sq[es$term == "Treat"], 2)`. Adjusted means
> were `r f1(em[["FT"]])` lb (family treatment), `r f1(em[["CBT"]])` lb
> (CBT) and `r f1(em[["Cont"]])` lb (control). Family treatment exceeded the
> control by `r f1(-pc("Cont - FT")$estimate)` lb (Tukey-adjusted
> `r fmt(pc("Cont - FT")$p_adjusted)`) and CBT exceeded it by
> `r f1(pc("CBT - Cont")$estimate)` lb
> (`r fmt(pc("CBT - Cont")$p_adjusted)`); the two treatments did not differ
> clearly (`r fmt(pc("CBT - FT")$p_adjusted)`). Because the model was
> chosen by a test on the same data, these p-values may be somewhat too small.

## See also

- `?anova_ancova` for every argument and the simulations behind the selection
  caveat.
- [Working with results](results.html) for the `anovakit_fit` object, the notes
  and `$emmeans_object`.
- `anova_welch()` ([article](welch.html)) and `anova_glm()`
  ([article](glm.html)) for comparisons without a covariate.
- `anova_manova()` ([article](manova.html)) for several responses, with or
  without covariates.
- `anova_rm()` ([article](repeated-measures.html)) when every subject is
  measured under several conditions.
- [Diagnostics](diagnostics.html) and [Plots and visual
  customisation](visuals.html).
