---
title: "Binary outcomes: logistic regression with anova_bin()"
description: >
  A walkthrough of anova_bin() on the Berkeley admissions data: analysis of
  deviance, odds ratios, marginal probabilities, frequency weights and
  separation.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

`anova_bin()` fits a logistic regression of a two-level outcome (admitted or
not, healed or not) on one or more grouping variables, and reports an analysis
of deviance table, odds ratios, marginal probabilities and pairwise odds
ratios. Use it when each observation is a yes/no outcome, or when you have a
table of such outcomes with a count in each row. For counts of events use
`anova_count()` ([walkthrough](counts.html)); for other response distributions
use `anova_glm()` ([walkthrough](glm.html)). The [decision guide](choosing.html)
covers the choice in more detail.

## The data

`UCBAdmissions` is a three-way table of the 4,526 applications to the six
largest graduate departments at Berkeley in 1973, by outcome, gender and
department. `as.data.frame()` turns it into one row per combination, with the
number of applicants in `Freq`.

```{r data}
ucb <- as.data.frame(UCBAdmissions)
head(ucb)
sum(ucb$Freq)
```

The famous feature of these data is that the overall admission rates and the
department-level rates point in different directions:

```{r data-rates}
# Overall admission rate by gender
round(prop.table(xtabs(Freq ~ Gender + Admit, data = ucb), 1), 3)

# Admission rate by department and gender, and where each gender applied
admitted <- xtabs(Freq ~ Dept + Gender, data = ucb[ucb$Admit == "Admitted", ])
applied  <- xtabs(Freq ~ Dept + Gender, data = ucb)
round(admitted / applied, 3)
applied
```

```{r data-inline, include = FALSE}
rate <- prop.table(xtabs(Freq ~ Gender + Admit, data = ucb), 1)
```

Overall, `r round(100 * rate["Male", "Admitted"], 1)`% of men and
`r round(100 * rate["Female", "Admitted"], 1)`% of women were admitted. Within
departments the gap mostly disappears, and in department A it reverses. The
explanation is in the last table: most women applied to departments C to F,
which admitted a third or fewer of their applicants, while just over half of the
men applied to A or B, which admitted most of theirs. Department is associated
with both gender and admission, so a comparison of genders that ignores it is
confounded. This is Simpson's paradox, and a model with both variables is the
way to separate the two.

Each row of `ucb` stands for `Freq` identical applicants, so the counts go in
as frequency weights. For later comparison, here is the same data with one row
per applicant:

```{r data-expand}
applicants <- ucb[rep(seq_len(nrow(ucb)), ucb$Freq),
                  c("Admit", "Gender", "Dept")]
nrow(applicants)
```

## Fitting the model

### Which outcome is the success?

The response can be a factor with two observed levels, a logical, a numeric
0/1 column, or a character column with two values. The function models the
probability of one level, the *success*, and by default that is the second
level of the factor, which is what `glm()` does. For `Admit` the second level
is `"Rejected"`:

```{r success-default}
levels(ucb$Admit)
default <- anova_bin(ucb, "Admit", "Gender", weights = "Freq", plots = FALSE)
default$notes[1]
```

That model is not wrong, but every odds ratio would be an odds ratio of
*rejection*. Name the level you want with `success`. (A logical response
models `TRUE`, a 0/1 response models `1`, and a character response is sorted
in the C locale, so the default does not depend on your session.)

### The fit

Admission modelled on gender and department, additively:

```{r fit}
fit <- anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
                 success = "Admitted")
fit
```

Reading the printout from the top:

- The method and the call.
- `Observations used: 24`. This counts *rows* of `$data_used`, not applicants:
  each row carries its weight, and the weights add up to
  `r format(sum(fit$data_used$Freq), big.mark = ",")`.
- The omnibus table, with a heading that says what it holds: Type II tests,
  likelihood-ratio chi-square statistics.
- The notes: which level is modelled, how the marginal means are laid out for
  an additive model, and why there is no dispersion statistic. They are
  discussed [below](#what-the-notes-say).
- The plots that were built (they are not drawn until you ask).

`summary(fit)` prints all of this plus the assumption tables, the odds ratios,
the marginal probabilities and the pairwise comparisons. The sections below
take the components one at a time.

## The omnibus test

```{r anova}
fit$anova
```

Each row is a likelihood-ratio test: `statistic` is the increase in deviance
when the term is dropped, referred to a chi-square distribution on `df`
degrees of freedom. Type II means each factor is tested after the other, so
the `Gender` row asks whether gender predicts admission *among applicants to
the same department*. It does not (p = `r signif(fit$anova$p_value[1], 3)`).
Department matters a great deal.

Compare the model with gender alone:

```{r anova-alone}
alone <- anova_bin(ucb, "Admit", "Gender", weights = "Freq",
                   success = "Admitted", plots = FALSE)
alone$anova
```

Without department the gender statistic is
`r round(alone$anova$statistic, 1)` on 1 degree of freedom. The difference
between the two tables is the paradox in numbers.

### The model as a whole

`$model_stats` holds statistics for the whole model:

```{r model-stats}
fit$model_stats
```

`lr_vs_null` is the likelihood-ratio test of this model against one with no
predictors (the null deviance minus the deviance, on
`r fit$model_stats$lr_vs_null$df` degrees of freedom). `mcfadden_r2` is
McFadden's pseudo R squared, `1 - deviance / null deviance`: the model removes
`r round(100 * fit$model_stats$mcfadden_r2, 1)`% of the deviance. It is not a
proportion of variance and runs lower than the R squared of a linear model, so
judge it against other logistic models rather than against that.
`success_level` repeats which level was modelled, so a saved result says what
its odds ratios are odds of. `AIC`, `BIC` and `logLik` use the prior weights,
and are there to compare models fitted to the same data.

## Effect sizes

`$effect_sizes` holds odds ratios, one per coefficient, each a level against
its factor's reference level:

```{r effect-sizes}
es_cols <- c("factor", "comparison", "odds_ratio", "conf_low", "conf_high",
             "p_value", "ci_method")
print(fit$effect_sizes[, es_cols], digits = 3)
```

`factor` names the grouping variable and `comparison` says what the ratio
compares. The first row is the odds of admission for women divided by the odds
for men, in the same department: `r round(fit$effect_sizes$odds_ratio[1], 3)`,
with an interval that comfortably includes 1. The department rows compare each
department with A; in department F the odds of admission are
`r round(fit$effect_sizes$odds_ratio[6], 3)` times those in A, whatever the
applicant's gender. (`term`, `se_log_or` and `statistic`, left out above, hold
the coefficient's name, its standard error on the log-odds scale and its Wald
z statistic.)

Put the adjusted odds ratio beside the one from the gender-only model:

```{r effect-simpson}
rbind(alone = alone$effect_sizes[, es_cols],
      adjusted = fit$effect_sizes[1, es_cols])
```

Ignoring department, women's odds of admission were about half of men's;
comparing applicants to the same department, they were slightly higher.

The odds ratios always come from a reference-coded fit of the same model, so
each one is a level against the reference level whatever `type`, the global
`contrasts` option or the class of the factor. `$anova` still comes from the
Type II or Type III fit, which describes the same fitted probabilities.

### Profile or Wald intervals

`ci_method = "profile"` (the default) gives profile-likelihood intervals, which
follow the shape of the likelihood rather than assuming it is quadratic. With
thousands of observations the two kinds agree closely:

```{r ci-wald}
wald <- anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
                  success = "Admitted", ci_method = "wald", plots = FALSE)
data.frame(comparison = fit$effect_sizes$comparison,
           profile_low = fit$effect_sizes$conf_low,
           profile_high = fit$effect_sizes$conf_high,
           wald_low = wald$effect_sizes$conf_low,
           wald_high = wald$effect_sizes$conf_high)
```

They part company in small samples and, above all, under separation, where
only the profile interval means anything (see [Separation](#separation)).

### Choosing the reference level

`reference` takes a named list, one level per factor:

```{r reference}
flipped <- anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
                     success = "Admitted",
                     reference = list(Gender = "Female", Dept = "F"),
                     plots = FALSE)
print(flipped$effect_sizes[, es_cols], digits = 3)
all.equal(flipped$anova, fit$anova)
```

The ratios are now men against women and each department against F. The
omnibus table is unchanged: the reference level changes what the coefficients
compare, not the fitted model.

## Marginal means and comparisons

`$emmeans` holds estimated marginal probabilities of the success level, with
intervals built on the logit scale and back-transformed, so they stay between
0 and 1:

```{r emmeans}
print(fit$emmeans, digits = 3)
```

Because the model is additive, each factor is reported on its own (the `term`
column says which), averaged over the levels of the other. The averaging gives
each department equal weight on the log-odds scale before transforming back,
so the gender rows are not the raw admission rates of
`r round(rate["Male", "Admitted"], 3)` and
`r round(rate["Female", "Admitted"], 3)`: they are the probabilities for a
hypothetical applicant spread evenly over the six departments. That is the
comparison the adjusted model makes.

`$posthoc` holds the pairwise comparisons, as odds ratios:

```{r posthoc}
ph_cols <- c("term", "contrast", "ratio", "conf_low", "conf_high", "p_value",
             "p_adjusted")
print(fit$posthoc[, ph_cols], digits = 3)
```

Each `ratio` is the odds for the first group in `contrast` divided by the odds
for the second, so the gender row, `Male / Female`, is the reciprocal of the
`Female vs Male` odds ratio in `$effect_sizes`. `p_value` is the unadjusted
p-value and `p_adjusted` the Tukey-adjusted one (the method is in the
`adjustment` column). Each factor is its own family: gender has one comparison,
so its two p-values are equal; department has 15, and the adjustment raises
them. The adjusted p-values printed as 0 are below the precision of the Tukey
calculation (about 1e-10); `summary(fit)` prints them as `< 1e-10`. The
intervals are simultaneous for each family.

`adjust` changes the method: `"tukey"` (the default), `"sidak"`,
`"scheffe"`, `"dunnettx"`, `"bonferroni"`, `"holm"`, `"hochberg"`,
`"hommel"`, `"BH"`, `"BY"`, `"fdr"` or `"none"`. A step-down method such as
Holm adjusts p-values but cannot be turned into intervals, and `$notes` says
what the intervals use instead:

```{r adjust-holm}
holm <- anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
                  success = "Admitted", adjust = "holm", plots = FALSE)
grep("adjustment", holm$notes, value = TRUE)
```

The department intervals are Bonferroni intervals, while the p-values are
Holm's. The gender family has a single comparison, which needs no adjustment of
either kind, so there is no note about it.

## Checking assumptions

A logistic regression makes few distributional assumptions that can be checked
from grouped data like these: the outcome is binary, the observations are
independent, and the log-odds are additive in the factors (which the
`interaction` option below tests). `$assumptions` has two elements.

```{r assumptions-dispersion}
fit$assumptions$dispersion
```

`dispersion` is always `NA`, and a note says why. For a 0/1 response the
Pearson dispersion carries no information about overdispersion: in a model
saturated in the cells it equals N / (N - k) whatever the data, and it changes
when the same trials are grouped into weighted rows. You can see the second
point by computing it on the two versions of this data set, which describe the
same fit:

```{r assumptions-pearson}
pearson <- function(m) sum(residuals(m, type = "pearson")^2) / df.residual(m)
one_row <- anova_bin(applicants, "Admit", c("Gender", "Dept"),
                     success = "Admitted", plots = FALSE)
c(weighted_table = pearson(fit$model),
  one_row_per_applicant = pearson(one_row$model))
```

`proportions` is the observed share of each response level in each cell. With
weights, `n` and `group_total` are weighted counts, and `rows` and
`group_rows` give the number of data rows behind them:

```{r assumptions-proportions}
props <- fit$assumptions$proportions
props[props$Admit == "Admitted", ]
```

This is the table to look at before trusting a model: it shows the raw
material, and it shows cells with few observations. Department B had only
`r props$group_total[props$.cell == "Female : B" & props$Admit == "Admitted"]`
women applicants. When fewer than about 5 events or non-events are expected in
a cell, `$notes` warns that the likelihood-ratio test is liberal; no cell here
comes close, but the next example shows the note.

More on the checks across the package is in the
[diagnostics article](diagnostics.html).

### Separation

When every observation in a group has the same outcome, the maximum-likelihood
odds ratio for that group is infinite. `glm()` still reports convergence, with
an enormous coefficient and an even larger standard error. Here is a small
constructed table in which every patient in the `high` arm healed:

```{r separation}
healing <- data.frame(
  arm    = factor(rep(c("control", "low", "high"), each = 2),
                  levels = c("control", "low", "high")),
  healed = rep(c("no", "yes"), times = 3),
  n      = c(9, 6, 5, 10, 0, 15)
)
healing
sep_fit <- anova_bin(healing, "healed", "arm", weights = "n")
sep_fit
```

The notes say what happened. The row with a count of zero was dropped (a
zero-weight row contributes nothing, and it is counted in `$n_removed`, which
is why the printout says 5 observations were used). Separation was detected in
`arm = high`, and the profile interval for that odds ratio is open on one side.
And the data are sparse: with 15 patients an arm, fewer than 5 non-events are
expected in each arm if the arms do not differ, so the likelihood-ratio test in
the table is liberal. The note names two tests that hold their level better,
and both are one line of code:

```{r separation-exact}
anova(sep_fit$model, test = "Rao")
fisher.test(xtabs(n ~ arm + healed, data = healing))
```

Both p-values are larger than the likelihood-ratio test's
`r format(signif(sep_fit$anova$p_value, 2), scientific = FALSE)`, though the
conclusion stands.

```{r separation-es}
print(sep_fit$effect_sizes[, es_cols], digits = 3)
sep_wald <- anova_bin(healing, "healed", "arm", weights = "n",
                      ci_method = "wald", plots = FALSE)
print(sep_wald$effect_sizes[, es_cols], digits = 3)
```

The point estimate for `high vs control` is meaningless, and so is its Wald
p-value of `r round(sep_fit$effect_sizes$p_value[2], 2)`. The profile interval
is still informative: it runs from
`r round(sep_fit$effect_sizes$conf_low[2], 1)` to infinity, so the data rule
out an odds ratio below about `r round(sep_fit$effect_sizes$conf_low[2])`. The
Wald interval, from 0 to infinity, says nothing at all. This is why
`ci_method = "profile"` is the default.

The separation note also warns that the separated arm's marginal probability
and the comparisons involving it are affected. They are Wald-based, and as
uninformative as the Wald odds ratio:

```{r separation-emm}
print(sep_fit$emmeans, digits = 3)
print(sep_fit$posthoc[, c("contrast", "ratio", "conf_low", "conf_high",
                          "p_adjusted")], digits = 3)
```

For a real analysis, follow the note: a penalised fit such as
`logistf::logistf()`, or collapsing the level with another, gives finite
estimates.

## Plots

`fit$plots` holds two **ggplot2** objects. Neither is drawn until you ask for
it, with `plot(fit, "name")` or `fit$plots$name`.

```{r plot-proportions}
plot(fit, "proportions")
```

The `proportions` plot stacks the observed share of each outcome in each cell
(gender by department here), labelled with percentages. What to look for:
cells whose proportions differ from their neighbours, cells near 0% or 100%
(the warning sign of separation), and whether a pattern in one factor holds
across the levels of the other. Here the within-department gender differences
are small next to the differences between departments, and department A is
the one where women did clearly better.

```{r plot-emmeans}
plot(fit, "emmeans")
```

The `emmeans` plot shows the estimated marginal probabilities with their
intervals, one panel per factor because the model is additive. What to look
for: intervals that do not overlap are clearly different; overlapping ones
need the comparisons in `$posthoc`. Each is a ggplot, so it can be modified
with `+`:

```{r plot-custom}
plot(fit, "emmeans") +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(y = "Probability of admission")
```

The [plots article](visuals.html) covers customisation in more detail.

## Options worth knowing

### `interaction` and `type`

The model above is additive: it assumes the gender odds ratio is the same in
every department. `interaction = TRUE` fits the full factorial and tests that
assumption. The first grouping variable goes on the x axis of the `emmeans`
plot, so department comes first here:

```{r interaction}
fit_int <- anova_bin(ucb, "Admit", c("Dept", "Gender"), weights = "Freq",
                     success = "Admitted", interaction = TRUE)
fit_int$anova
```

The `Dept:Gender` row is a likelihood-ratio test of whether the gender odds
ratio differs between departments (p =
`r signif(fit_int$anova$p_value[3], 2)`). With an interaction, each lower-order
odds ratio is taken at the reference level of the factors it interacts with,
and the interaction rows are ratios of odds ratios; the `comparison` column
says which:

```{r interaction-es}
print(fit_int$effect_sizes[, es_cols], digits = 3)
```

The `Gender` row is the gender odds ratio in department A, and each
`Dept:Gender` row is the gender odds ratio in another department divided by
the one in A. All of them are below 1: department A is the odd one out.

`$emmeans` now has one row per cell, and `$posthoc` compares every pair of
cells (`r nrow(fit_int$posthoc)` comparisons, most of them not interesting).
For the comparison that matters here, women against men within each
department, pass `$emmeans_object` to **emmeans** directly:

```{r interaction-contrast}
emmeans::contrast(fit_int$emmeans_object, method = "revpairwise", by = "Dept")
```

The grid carries **emmeans**' own default confidence level rather than
`conf_level`, so give `level =` when you summarise it.

```{r interaction-plot}
plot(fit_int, "emmeans")
```

With department on the x axis and one line per gender, the interaction is easy
to see: the two lines lie almost on top of each other everywhere except in
department A.

With an interaction in the model, `type` matters. Type II tests each main
effect assuming the interaction is absent; Type III (fitted under sum-to-zero
contrasts) tests it averaged over the other factor's levels:

```{r interaction-type3}
fit_int3 <- anova_bin(ucb, "Admit", c("Dept", "Gender"), weights = "Freq",
                      success = "Admitted", interaction = TRUE, type = "III",
                      plots = FALSE)
fit_int3$anova
all.equal(fit_int3$effect_sizes, fit_int$effect_sizes)
```

Only the omnibus table changes: the odds ratios come from a reference-coded
fit whatever `type` is. With an interaction this clear, neither main-effect row
tells the whole story; report the interaction and the within-department odds
ratios.

### `test_statistic`

The omnibus table uses likelihood-ratio tests by default.
`test_statistic = "Wald"` gives Wald chi-square tests instead:

```{r wald-test}
anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
          success = "Admitted", test_statistic = "Wald", plots = FALSE)$anova
```

The two agree closely for gender and differ for department, whose effects are
large. The likelihood-ratio test is the better default; the Wald test is there
mainly because it is the one that can use a robust covariance (below).

### Frequency weights and `vcov_type`

Whole-number weights, some above 1, are frequency weights: a row with weight
`w` counts as `w` identical observations. The weighted table therefore gives
the same answers as one row per applicant. `one_row` was fitted above on the
`r format(nrow(applicants), big.mark = ",")`-row version:

```{r weights-equal}
all.equal(fit$anova, one_row$anova)
data.frame(comparison = fit$effect_sizes$comparison,
           or_table = fit$effect_sizes$odds_ratio,
           or_applicants = one_row$effect_sizes$odds_ratio,
           se_table = fit$effect_sizes$se_log_or,
           se_applicants = one_row$effect_sizes$se_log_or)
```

The standard errors agree to five or more significant figures (the rest is
`glm()`'s convergence tolerance). The one statistic that differs is BIC,
because R bases it on the number of rows:

```{r weights-bic}
c(table = fit$model_stats$BIC, applicants = one_row$model_stats$BIC)
```

`vcov_type` requests heteroskedasticity-consistent (sandwich) standard errors,
`"HC0"` to `"HC4"`, which need the **sandwich** package. With frequency
weights they are computed as they would be on the expanded data. A sandwich
computed row by row on the table would treat each row as one unit, and
overstate the standard errors badly:

```{r weights-hc3}
hc_table <- anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
                      success = "Admitted", vcov_type = "HC3", plots = FALSE)
hc_rows <- anova_bin(applicants, "Admit", c("Gender", "Dept"),
                     success = "Admitted", vcov_type = "HC3", plots = FALSE)
naive <- sqrt(diag(sandwich::vcovHC(fit$model, type = "HC3")))[-1]
data.frame(comparison = hc_table$effect_sizes$comparison,
           model_based = fit$effect_sizes$se_log_or,
           hc3_table = hc_table$effect_sizes$se_log_or,
           hc3_applicants = hc_rows$effect_sizes$se_log_or,
           hc3_row_by_row = unname(naive))
```

The HC3 standard errors from the table and from one row per applicant are the
same, and here both are close to the model-based ones. The last column,
`sandwich::vcovHC()` applied to the table directly, is wrong by a factor of 10
or more. The notes record what the robust covariance was used for:

```{r weights-hc3-notes}
hc_table$notes
```

Robust standard errors reach the odds ratios, the marginal means and the
comparisons, but not a likelihood-ratio omnibus test, which compares
deviances; set `test_statistic = "Wald"` for an omnibus test that uses them.
The odds-ratio intervals also become Wald intervals, since a profile
likelihood cannot use a sandwich covariance.

### `conf_level` and `posthoc`

`conf_level` sets the level of every interval in the result: odds ratios,
marginal probabilities and comparisons. `posthoc = FALSE` skips the pairwise
comparisons, which is worth doing when there are many cells; `$notes` records
how many were skipped, and `$effect_sizes` is still computed.

```{r conf-level}
narrow <- anova_bin(ucb, "Admit", c("Gender", "Dept"), weights = "Freq",
                    success = "Admitted", conf_level = 0.90, posthoc = FALSE,
                    plots = FALSE)
narrow$effect_sizes[1, c("comparison", "odds_ratio", "conf_low", "conf_high")]
narrow$emmeans[1:2, c("Gender", "estimate", "conf_low", "conf_high")]
grep("posthoc|still reports", narrow$notes, value = TRUE)
```

## What the notes say

```{r notes}
fit$notes
```

- The first note names the level being modelled. Check it on every fit: an
  odds ratio of admission and an odds ratio of rejection are reciprocals, and
  confusing them reverses every conclusion.
- The second says that, in an additive model, the marginal means and
  comparisons are reported for each factor separately, averaged over the
  other, with each factor's comparisons adjusted as a separate family.
- The third explains why `$assumptions$dispersion` is `NA`.

Other fits in this article showed the notes you may meet elsewhere: a
zero-weight row dropped, separation detected and an open-ended profile
interval, and robust standard errors with frequency weights. The
[results article](results.html) explains how the notes fit with the rest of
the object.

## Reporting the result

```{r report-inline, include = FALSE}
a <- fit$anova
es <- fit$effect_sizes
fmt_p <- function(p) if (p < 0.001) "< 0.001" else paste("=", signif(p, 2))
```

A write-up of the adjusted analysis might read:

> Admission was modelled by logistic regression on gender and department for
> the `r format(sum(ucb$Freq), big.mark = ",")` applicants. Department was
> strongly associated with admission (likelihood-ratio
> χ²(`r a$df[2]`) = `r round(a$statistic[2], 1)`, p `r fmt_p(a$p_value[2])`),
> but gender was not once department was taken into account
> (χ²(`r a$df[1]`) = `r round(a$statistic[1], 2)`, p `r fmt_p(a$p_value[1])`;
> odds ratio for women against men `r sprintf("%.2f", es$odds_ratio[1])`, 95%
> profile-likelihood CI `r sprintf("%.2f", es$conf_low[1])` to
> `r sprintf("%.2f", es$conf_high[1])`). The unadjusted odds ratio of
> `r sprintf("%.2f", alone$effect_sizes$odds_ratio)` is accounted for by
> the departments women applied to. The gender odds ratio varied between
> departments (χ²(`r fit_int$anova$df[3]`) =
> `r round(fit_int$anova$statistic[3], 1)`, p
> `r fmt_p(fit_int$anova$p_value[3])`), being largest in department A.

## See also

- `?anova_bin` for every argument, and `?anovakit_fit` for the returned
  object.
- [Counts and rates with `anova_count()`](counts.html), the companion function
  for count responses.
- [Generalised linear models with `anova_glm()`](glm.html) for other families.
- [Working with results](results.html) and [Diagnostics](diagnostics.html).
- [Plots and visual customisation](visuals.html).
