---
title: "Which function? Choosing an analysis"
description: >
  A decision guide: four questions about your data that lead to one of the
  eight functions, the situations each one is built for, and the mistakes the
  choice is meant to avoid.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

anovakit has eight analysis functions. They share one calling convention and
return the same kind of object, so moving between them costs nothing. The
choice between them is a choice about your *data*: what the outcome is, how
the observations are related, and what you need to adjust for. This guide
walks through that choice and links to a full walkthrough of each function.

## Four questions

Answer these in order. The first answer that settles it wins.

**1. Is the outcome a yes/no, a count, or something other than a continuous
measurement?**

| Outcome | Function |
|---|---|
| Two categories (yes/no, success/failure, 0/1) | `anova_bin()` |
| A count of events (0, 1, 2, ...), possibly over differing exposure time | `anova_count()` |
| A proportion out of a known number of trials, a positive right-skewed measurement (costs, times), or anything else with a GLM family | `anova_glm()` |
| A continuous measurement | go to question 2 |

**2. Are the same subjects measured more than once (before and after, at
several time points, under several conditions)?**

If yes, use `anova_rm()`. Treating repeated measurements as independent
observations overstates the sample size and makes every p-value too small.

**3. Are there several continuous outcomes that should be analysed together?**

If yes, use `anova_manova()`. It tests the groups on all outcomes jointly and
then follows up one outcome at a time.

**4. Is there a numeric covariate to adjust for (a baseline score, age, a
pre-treatment measurement)?**

If yes, use `anova_ancova()`. If no, use `anova_welch()`, or `anova_kw()`
when the outcome is ordinal or so skewed that a mean is not a useful summary.

That covers most analyses. The rest of this guide explains each branch, with
the reasoning behind it.

## One continuous outcome, independent groups

### `anova_welch()`: the default for comparing means

Welch's test compares group means without assuming the groups have equal
variances. When the variances are equal it gives almost the same answer as the
classical F test; when they are not, the classical test can be badly wrong,
especially when the smallest group has the largest variance. There is little to
lose and a lot to gain, so it is the default choice.

```{r welch}
fit <- anova_welch(PlantGrowth, "weight", "group", plots = FALSE)
fit$anova
```

The [Welch walkthrough](welch.html) covers the standardised mean differences,
the pairwise comparisons and the per-group diagnostics.

### `anova_kw()`: when a mean is the wrong summary

The Kruskal-Wallis test compares the groups' rank distributions. Use it for
ordinal outcomes (convert an ordered factor with `as.integer()` first), for
heavy-tailed or very skewed data in small samples, or when outliers are real
values you cannot drop. It answers a different question from Welch's test —
whether values in one group tend to be larger than in another — so it is not a
drop-in replacement when you care about means.

```{r kw}
anova_kw(airquality, "Ozone", "Month", plots = FALSE)$anova
```

See the [Kruskal-Wallis walkthrough](kruskal-wallis.html).

### `anova_ancova()`: when a covariate explains part of the outcome

If a numeric variable measured before treatment (a baseline score, age,
weight) predicts the outcome, adjusting for it removes that part of the
variation and makes the group comparison more precise. In a randomised study,
it is also the right way to use a baseline measurement: better than analysing
change scores, and better than ignoring the baseline.

```{r ancova}
anc <- anova_ancova(MASS::anorexia, "Postwt", "Treat", "Prewt", plots = FALSE)
anc$anova
```

The [ANCOVA walkthrough](ancova.html) explains covariate centring, the
homogeneity-of-slopes test and when to fix the model in advance.

### Several grouping variables

Every function accepts several grouping columns. `anova_welch()` and
`anova_kw()` combine them into one factor of the cells that contain data. The
model-based functions (`anova_ancova()`, `anova_manova()`, `anova_bin()`,
`anova_count()` and `anova_glm()`) fit them as factors, with or without their
interaction (`interaction`), and with Type II or Type III sums of squares
(`type`). For a classical two-way ANOVA of a continuous outcome, use
`anova_glm()` with its default Gaussian family:

```{r twoway}
tg <- transform(ToothGrowth, dose = factor(dose))
anova_glm(tg, "len", c("supp", "dose"), interaction = TRUE, plots = FALSE)$anova
```

See the [GLM walkthrough](glm.html).

## Repeated measurements: `anova_rm()`

When each subject contributes several measurements, the measurements are
correlated. `anova_rm()` fits the repeated-measures ANOVA through **afex**, with
within-subject factors (`within`), between-subject factors (`between`), the
Greenhouse-Geisser or Huynh-Feldt correction for sphericity, and generalised eta
squared. Subjects missing a cell are removed and listed. The
[repeated-measures walkthrough](repeated-measures.html) shows a mixed design.

A pre/post design with one follow-up measurement is a special case worth
knowing: if the groups were randomised, an ANCOVA of the post score adjusting
for the pre score usually has more power than the repeated-measures analysis
of the same data, and it answers the question most studies ask (did the groups
differ after treatment, given where they started?).

## Several outcomes: `anova_manova()`

Use a MANOVA when the outcomes are conceptually one construct measured several
ways, or when you want one test that controls the error rate across them before
looking at each. It uses the correlations between the outcomes, which separate
ANOVAs ignore. With covariates it becomes a MANCOVA.

```{r manova}
anova_manova(iris, c("Sepal.Length", "Sepal.Width", "Petal.Length",
                     "Petal.Width"), "Species", plots = FALSE)$multivariate
```

The [MANOVA walkthrough](manova.html) covers the four test statistics, the
canonical discriminant analysis, and what to do when Box's M rejects.

## Outcomes that are not continuous

### `anova_bin()`: binary outcomes

Logistic regression, reported as an analysis of deviance, with odds ratios,
marginal probabilities and pairwise comparisons. It detects separation (a group
in which every outcome is the same), which makes ordinary intervals meaningless.
Aggregated data (one row per group and outcome, with a count) go in through
`weights`. See the [binary-outcomes walkthrough](binary.html).

### `anova_count()`: counts and rates

Poisson regression when the counts are no more variable than a Poisson
distribution allows, and a negative binomial model when they are overdispersed,
chosen automatically from the Pearson dispersion unless you choose yourself with
`model`. When units were observed for different lengths of time or at different
sizes, give the exposure as `offset` and the results become rates. See the
[counts walkthrough](counts.html).

### `anova_glm()`: everything else

Any `stats::family`: Gamma or inverse Gaussian for positive, right-skewed
measurements; binomial with trial weights for proportions; quasi families when
the variance does not follow the family's rule. With the default Gaussian family
it is the classical ANOVA with Type II or Type III sums of squares. See the
[GLM walkthrough](glm.html).

## Common situations

| You have | Use | Not |
|---|---|---|
| Three or more groups, variances look different | `anova_welch()` | a classical ANOVA, which assumes equal variances |
| Two groups | `anova_welch()` (it is Welch's t test) | a pooled t test |
| Ratings on a 1–5 scale | `anova_kw()` | a comparison of means of ordinal codes |
| A baseline and a follow-up score, randomised groups | `anova_ancova()` with the baseline as covariate | change scores |
| The same people at three time points | `anova_rm()` | an ANOVA that treats each measurement as a new person |
| Height, weight and waist size on the same people | `anova_manova()` | three separate ANOVAs with no joint test |
| Passed / failed | `anova_bin()` | a comparison of 0/1 means |
| Number of visits over different lengths of follow-up | `anova_count()` with `offset` | counts per person ignoring exposure |
| Successes out of a known number of trials | `anova_glm()` with `family = "binomial"` and the trials as `weights` | percentages analysed as continuous |
| Costs or durations, positive and skewed | `anova_glm()` with `family = Gamma(link = "log")` | a log transformation, when you want means on the original scale |

## Arguments that work the same everywhere

Once you have chosen, the rest is shared. Every function takes the data first
and names columns as character strings. These arguments mean the same thing
wherever they appear:

| Argument | Accepted by | What it does |
|---|---|---|
| `conf_level` | all eight | The level of every interval the function returns. |
| `adjust` | all eight | The multiplicity adjustment for the pairwise comparisons. |
| `posthoc` | all eight | `FALSE` skips the pairwise comparisons. |
| `plots` | all eight | `FALSE` skips building the figures, which is faster. |
| `verbose` | all eight | Progress messages. |
| `interaction` | ancova, manova, bin, count, glm | Additive, full factorial, or up to a given order. |
| `type` | ancova, manova, bin, count, glm | Type II or Type III tests. |
| `vcov_type` | ancova, bin, count, glm | Heteroscedasticity-consistent (sandwich) standard errors. |
| `weights` | bin, count, glm | Prior weights, named as a column. |

The object every function returns is described in
[Working with results](results.html). Before you report anything from a fit,
read its `$notes`: it is where the function records what it decided on your
behalf and what it could not compute.
