---
title: "Counts and rates: analysis of deviance with anova_count()"
description: >
  A walkthrough of anova_count() on the warpbreaks and insurance-claims data:
  Poisson or negative binomial, rate ratios, marginal rates, exposure offsets,
  frequency weights and groups with no events.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

`anova_count()` fits a regression for a count response (the number of
breaks, claims, visits or incidents) across one or more grouping variables. It
starts from a Poisson model, checks it for overdispersion, moves to a negative
binomial model when the counts are more variable than a Poisson allows, and
reports an analysis of deviance table, incidence rate ratios, estimated
marginal rates and pairwise rate ratios. With an exposure column (time at
risk, number of policies) it compares rates rather than counts. For a yes/no
outcome use `anova_bin()` ([walkthrough](binary.html)); for other response
distributions, `anova_glm()` ([walkthrough](glm.html)). The
[decision guide](choosing.html) covers the choice in more detail.

## The data

`warpbreaks` records the number of warp breaks on each of 54 looms, where a
loom corresponds to a fixed length of yarn, for two types of wool (`A`, `B`)
at three levels of tension (`L`, `M`, `H`): a 2 by 3 factorial experiment with
9 looms in each cell.

```{r data}
head(warpbreaks)
table(warpbreaks$wool, warpbreaks$tension)
aggregate(breaks ~ wool + tension, data = warpbreaks,
          FUN = function(x) round(c(mean = mean(x), variance = var(x)), 1))
```

```{r data-inline, include = FALSE}
cell_stats <- aggregate(breaks ~ wool + tension, data = warpbreaks,
                        FUN = function(x) var(x) / mean(x))
```

The breaks are counts, so a Poisson model is the natural starting point. A
Poisson distribution has its variance equal to its mean, and here the variance
within each cell is between `r round(min(cell_stats$breaks), 1)` and
`r round(max(cell_stats$breaks), 1)` times the mean. The counts are
*overdispersed*, and dealing with that is the first thing the function does.

## Fitting the model

Columns are named as character strings: the response, then the grouping
variables. This is a designed factorial experiment, and whether the effect of
tension depends on the wool is part of the question, so the model includes the
interaction:

```{r fit}
fit <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                   interaction = TRUE)
fit
```

Reading the printout from the top:

- The method names the model actually fitted: a negative binomial regression,
  not the Poisson one you might have expected.
- `Observations used: 54`, all of them.
- The omnibus table, headed with what it holds: Type II tests,
  likelihood-ratio chi-square statistics.
- The notes: why the model is negative binomial, and a warning about what its
  tests are worth with 9 looms per cell. Both are discussed in the next
  section and [below](#what-the-notes-say).
- The plots that were built (they are not drawn until you ask).

`summary(fit)` prints this plus the assumption checks, the rate ratios, the
marginal means and the pairwise comparisons.

## Poisson, negative binomial or quasi-Poisson?

The `model` argument decides the model. Its default, `"auto"`, fits a Poisson
model first and computes its Pearson dispersion: the sum of squared Pearson
residuals divided by the residual degrees of freedom, which is about 1 when the
Poisson variance is right. Above `overdispersion_threshold` (1.5 by default)
the function fits a negative binomial model instead, which needs the **MASS**
package (without it, or if that fit fails, it uses quasi-Poisson).
`$model_type` records the outcome:

```{r model-type}
fit$model_type
fit$dispersion
fit$model_dispersion
```

`$dispersion` is the Pearson dispersion of the *Poisson* fit, the statistic
the decision was made on: `r round(fit$dispersion, 2)`, well above 1.5.
`$model_dispersion` is the same statistic for the model actually returned. At
`r round(fit$model_dispersion, 2)` it says the negative binomial's variance
function, `mu + mu^2 / theta`, has absorbed the extra variation. The second
note gives theta, `r round(fit$model$theta, 1)`, and goes on to say something
important: with 9 observations per cell the negative binomial's
likelihood-ratio tests and Wald intervals are markedly anti-conservative, and
quasi-Poisson F tests are better calibrated at this size.

Set `model` yourself to take the decision out of the function's hands. Here
is the interaction test under each of the three models:

```{r model-compare}
models <- c("poisson", "quasipoisson", "negbin")
fits <- lapply(models, function(m)
  anova_count(warpbreaks, "breaks", c("wool", "tension"), interaction = TRUE,
              model = m, plots = FALSE))
data.frame(model = vapply(fits, `[[`, "", "model_type"),
           test = vapply(fits, function(f) attr(f$anova, "statistic"), ""),
           statistic = vapply(fits, function(f) f$anova$statistic[3], 0),
           p_value = vapply(fits, function(f) f$anova$p_value[3], 0))
```

The Poisson model, which ignores the overdispersion, puts the p-value for the
interaction below one in a million. Both models that allow for it put it
between 0.01 and 0.05. An explicit `model = "poisson"` is kept even on data
like these, but not silently:

```{r model-poisson-notes}
fits[[1]]$notes[1]
```

The note puts a size on the problem: at a dispersion of
`r round(fits[[1]]$dispersion, 2)`, a nominal 5% test rejects roughly 31% of
true null hypotheses, six times as often as it should.

The quasi-Poisson model keeps the Poisson estimates and scales their variance
by the estimated dispersion. It has no likelihood, so its table uses F tests:

```{r model-quasi}
fits[[2]]$anova
fits[[2]]$notes
```

`overdispersion_threshold` moves the cut-off for `model = "auto"`. It is a
fixed cut-off, not a test, and a Poisson model kept below it still gets the
note on inflated statistics when the dispersion is beyond what chance would
produce:

```{r model-threshold}
lenient <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                       interaction = TRUE, overdispersion_threshold = 5,
                       plots = FALSE)
lenient$model_type
lenient$notes
```

Which to use? The negative binomial is a full likelihood model, and with more
data per group it is usually the better choice. With groups as small as these,
the note's advice to prefer quasi-Poisson is worth taking for the tests; the
rest of this article stays with the automatic choice so that you can see what
it reports.

## The omnibus test

```{r anova}
fit$anova
```

Each row is a likelihood-ratio test: `statistic` is the increase in deviance
when the term is dropped (with theta held at its estimate), referred to a
chi-square distribution on `df` degrees of freedom. Type II means each main
effect is tested after the other and without the interaction, and the
interaction after both. The `wool:tension` row says the effect of tension
depends on the wool (p = `r signif(fit$anova$p_value[3], 2)`), so the main
effects cannot be read on their own; the rate ratios and marginal means below
show what the interaction looks like.

## Effect sizes: rate ratios

`$effect_sizes` holds incidence rate ratios (IRRs), each a level of a factor
against its reference (first) level:

```{r effect-sizes}
print(fit$effect_sizes, digits = 3)
```

An IRR is the ratio of two expected counts (or two rates, with an offset): 1
means no difference, 0.5 half as many, 2 twice as many. `factor` names the
grouping variable and `comparison` says what is compared. In a model with an
interaction, a main-effect ratio holds at the reference level of the factor it
interacts with: `B vs A at tension = L` says wool B has
`r round(fit$effect_sizes$IRR[1], 2)` times the breaks of wool A *at low
tension*. The interaction rows are ratios of rate ratios:
`(B vs A) x (M vs L)` is `r round(fit$effect_sizes$IRR[4], 2)`, meaning the
B-to-A ratio at medium tension is `r round(fit$effect_sizes$IRR[4], 2)` times
the B-to-A ratio at low tension.

The intervals are Wald intervals on the log scale, exponentiated, and
`p_value` tests each ratio against 1. They are always computed under
treatment (reference) coding, whatever `type` or the global `contrasts`
option; `term` gives the name of the coefficient in that coding.

## Marginal means and comparisons

`$emmeans` holds the estimated marginal means on the count scale, here the
expected breaks per loom in each cell, with intervals built on the log scale
and back-transformed:

```{r emmeans}
print(fit$emmeans, digits = 3)
```

With the full interaction in the model these are the cell means of the data;
the model's contribution is the intervals. `$posthoc` compares every pair of
cells, as rate ratios:

```{r posthoc}
print(fit$posthoc[, c("contrast", "IRR", "conf_low", "conf_high", "p_value",
                      "p_adjusted", "adjustment")], digits = 3)
```

`IRR` is the rate of the first cell in `contrast` divided by the rate of the
second. `p_value` is unadjusted and `p_adjusted` is adjusted for all
`r nrow(fit$posthoc)` comparisons by the method in `adjustment` (Tukey by
default; `adjust` changes it). The intervals are simultaneous.

Most of those comparisons (`A L / B M`, say) answer no question anyone asked.
For the comparisons of tensions within each wool, pass `$emmeans_object` to
**emmeans**:

```{r posthoc-by}
emmeans::contrast(fit$emmeans_object, method = "pairwise", by = "wool")
```

With wool A, low tension gives clearly more breaks than medium or high, which
do not differ. With wool B, high tension gives fewer breaks than the other two,
less decisively.

## Rates: counts with an exposure

Counts are often collected over different amounts of exposure: more
policyholders, longer follow-up, larger areas. `MASS::Insurance` gives the
number of car insurance claims in 64 groups of policyholders, by district, car
group (engine size) and driver age, together with the number of policyholders
in each group. `District` is left out below to keep the example small.

```{r ins-data}
ins <- MASS::Insurance
head(ins)
```

A group with 1,000 policyholders will have more claims than one with 100, so
the counts must be compared per policyholder. Name the exposure column in
`offset`. Give it on its natural scale: the function takes the logarithm
itself and enters it as `offset(log(Holders))` in the model formula.

```{r ins-fit}
ins_fit <- anova_count(ins, "Claims", c("Group", "Age"), offset = "Holders")
ins_fit
print(formula(ins_fit$model), showEnv = FALSE)
```

```{r ins-inline, include = FALSE}
ins_dfr <- df.residual(ins_fit$model)
ins_p_over <- pchisq(ins_fit$dispersion * ins_dfr, ins_dfr, lower.tail = FALSE)
```

Here the Poisson model was kept: the dispersion,
`r round(ins_fit$dispersion, 2)`, is below the threshold and within what
chance produces (a Pearson test of overdispersion on its `r ins_dfr` residual
degrees of freedom gives p = `r round(ins_p_over, 2)`), so there is no note on
inflation. Both factors matter:

```{r ins-es}
print(ins_fit$effect_sizes, digits = 3)
```

The rate ratios are now ratios of claims *per policyholder*: drivers aged over
35 make `r round(ins_fit$effect_sizes$IRR[6], 2)` times as many claims per
policy as drivers under 25 with the same size of car, and cars over 2 litres
`r round(ins_fit$effect_sizes$IRR[3], 2)` times as many as cars under 1 litre.
(`Group` and `Age` are ordered factors. The rate ratios compare levels with
the first level all the same, as they would for unordered factors.)

The marginal means are rates at an exposure of 1, which a note records: claims
per policyholder. The unit of the exposure is the unit of the rate, so to
report claims per thousand policyholders, give the exposure in thousands:

```{r ins-rates}
ins$thousands <- ins$Holders / 1000
per_1000 <- anova_count(ins, "Claims", c("Group", "Age"),
                        offset = "thousands", plots = FALSE)
print(per_1000$emmeans, digits = 3)
grep("exposure", per_1000$notes, value = TRUE)
all.equal(per_1000$effect_sizes, ins_fit$effect_sizes)
```

The rate ratios, and everything else except the scale of the marginal means,
are unchanged. The model is additive, so the marginal rates are reported for
each factor separately, averaged over the other (the `term` column), and
`$posthoc` compares levels within each factor:

```{r ins-posthoc}
print(per_1000$posthoc[, c("term", "contrast", "IRR", "conf_low", "conf_high",
                           "p_adjusted")], digits = 3)
```

## Checking assumptions

`$assumptions` has three elements:

```{r assumptions}
fit$assumptions
```

- `poisson_dispersion` is the Pearson dispersion of the Poisson fit (the same
  number as `$dispersion`). Near 1 the Poisson variance is adequate; well
  above it, the counts are overdispersed. Below 1 they are underdispersed,
  which is rarer and makes Poisson tests conservative.
- `model_dispersion` is the same statistic for the model returned (the same
  number as `$model_dispersion`). For a negative binomial fit, a value near 1
  says the variance function fits; one well above 1 says it does not, and that
  a model for the source of the extra variation (a missing grouping variable,
  clustering) is needed.
- `cell_counts` is the number of rows in each cell of the design. Small cells
  are where the asymptotic tests are least trustworthy, as the note on theta
  said.

The model also assumes independent counts and that the log rate is additive
in the terms included, which the interaction test checks for the grouping
factors. More on the checks across the package is in the
[diagnostics article](diagnostics.html).

### A group with no events

When a group has no events at all, its fitted rate is zero, the log of it is
minus infinity, and every rate ratio involving it is undefined. `glm()` still
reports convergence, with an enormous negative coefficient. To show what the
function does, here are simulated weekly incident counts at three sites; the
true rate at `east` is 0.05 a week, small but not zero, and in ten weeks it had
no incidents:

```{r zero-data}
set.seed(1)
incidents <- data.frame(
  site = factor(rep(c("north", "south", "east"), each = 10),
                levels = c("north", "south", "east")),
  events = c(rpois(10, 2), rpois(10, 4), rpois(10, 0.05))
)
tapply(incidents$events, incidents$site, sum)
zero_fit <- anova_count(incidents, "events", "site")
zero_fit
```

Two notes flag the problem. The first counts the groups with fewer than 5
events in total, whose rates rest on very little information. The third names
the group with no events, and says what that does to the rate ratios and what
still works. The results bear it out:

```{r zero-es}
print(zero_fit$effect_sizes, digits = 3)
print(zero_fit$posthoc[, c("contrast", "IRR", "conf_low", "conf_high",
                           "p_adjusted")], digits = 3)
```

`east vs north` is estimated as practically zero, with an interval from 0 to
infinity and a p-value of `r round(zero_fit$effect_sizes$p_value[2], 2)`:
the Wald machinery has nothing to work with, however clear the difference.
The likelihood-ratio test in the omnibus table does work, and it is the number
to report for the overall comparison. For the rate ratios involving `east`,
follow the note's advice: collapse the level with another, or use an exact
method. A robust covariance would fail here too, and `vcov_type` falls back
to the model-based one with a note saying so:

```{r zero-robust}
grep("Robust", anova_count(incidents, "events", "site", vcov_type = "HC3",
                           plots = FALSE)$notes, value = TRUE)
```

## Plots

`fit$plots` holds two **ggplot2** objects, returned rather than drawn. Use
`plot(fit, "name")` or `fit$plots$name` to get one.

```{r plot-emmeans}
plot(fit, "emmeans")
```

The `emmeans` plot shows the estimated marginal means with their intervals.
The first grouping variable goes on the x axis and the second becomes the
colour, one line per level. What to look for: lines that are not parallel
show an interaction. For wool A, low tension stands well above medium and
high; for wool B, low and medium are together and high is lowest. The
low-tension line falls steeply from A to B while the medium one rises, so the
lines are far from parallel.

```{r plot-observed}
plot(fit, "observed")
```

The `observed` plot is a box plot of the raw counts in each cell, ordered by
median and labelled with the cell's number of rows. What to look for: spread
that grows with the median (expected for counts, and what the negative
binomial's variance function allows for), outlying looms, and whether the
order of the medians agrees with the marginal means.

For a model with an offset, keep in mind that this plot shows counts, not
rates. In the insurance data the groups with the most claims are those with the
most policyholders, which is not the same as those with the highest rate:

```{r plot-observed-ins, fig.height = 5}
plot(ins_fit, "observed")
```

The four cells for drivers over 35 have the highest median counts, although
that age band has the lowest claim rate in `$emmeans`: it simply holds most of
the policyholders. Plot `Claims / Holders` yourself, or rely on the `emmeans`
plot, when the rates are the point. Each plot is a ggplot and can be modified
with `+`; the [plots article](visuals.html) has more.

## Options worth knowing

### `interaction`

The model is additive by default. An additive model gives each factor's
effect averaged over the other, which is what you want when the interaction is
absent and misleading when it is not. For comparison, the additive fit of the
same data reports marginal means for each factor separately:

```{r additive}
additive <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                        plots = FALSE)
additive$anova
print(additive$emmeans, digits = 3)
```

The wool rows average over tensions, hiding the fact that the wool difference
runs in opposite directions at low and medium tension. Decide on the
interaction before looking at the answers, as the design here justified.

### `test_statistic`

`"LR"` (the default) gives likelihood-ratio tests; `"Wald"` gives Wald
chi-square tests; `"F"` gives F tests, which is what a quasi-Poisson model
uses anyway. On a negative binomial (or Poisson) model, `"F"` estimates a
dispersion from the Pearson residuals and divides the deviances by it, which
makes the table a quasi-likelihood test while the rate ratios still assume the
model's own dispersion. The note says so:

```{r test-statistic}
wald_fit <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                        interaction = TRUE, test_statistic = "Wald",
                        plots = FALSE)
wald_fit$anova
f_fit <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                     interaction = TRUE, test_statistic = "F", plots = FALSE)
f_fit$anova
grep("test_statistic", f_fit$notes, value = TRUE)
```

For quasi-likelihood inference throughout, use `model = "quasipoisson"`.

### `vcov_type`

`vcov_type` requests heteroskedasticity-consistent (sandwich) standard errors,
`"HC0"` to `"HC4"`, from the **sandwich** package. They reach the rate ratios,
the marginal means and the comparisons. A likelihood-ratio test compares
deviances and cannot use them, so the omnibus table uses them only with
`test_statistic = "Wald"`:

```{r vcov}
robust <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                      interaction = TRUE, vcov_type = "HC3",
                      test_statistic = "Wald", plots = FALSE)
robust$anova
data.frame(comparison = fit$effect_sizes$comparison,
           model_low = fit$effect_sizes$conf_low,
           model_high = fit$effect_sizes$conf_high,
           hc3_low = robust$effect_sizes$conf_low,
           hc3_high = robust$effect_sizes$conf_high)
grep("robust", robust$notes, value = TRUE)
```

The robust intervals are wider here: with 9 looms a cell, the sandwich
estimate is noisy, and it is better used as a check on the model-based
standard errors than as a replacement for them.

### Frequency weights

A table of counts with a column saying how often each count occurred is a
frequency table. `weights` names that column, and whole-number weights (some
above 1) are read as frequencies: each row stands for that many identical
observations. `InsectSprays` has 12 insect counts for each of six sprays;
here it is as a frequency table:

```{r weights-table}
spray_tab <- as.data.frame(table(spray = InsectSprays$spray,
                                 count = InsectSprays$count),
                           stringsAsFactors = FALSE)
spray_tab <- spray_tab[spray_tab$Freq > 0, ]
spray_tab$count <- as.numeric(spray_tab$count)
head(spray_tab)
c(rows = nrow(spray_tab), observations = sum(spray_tab$Freq))
```

The frequency table and the original 72 rows give the same analysis:

```{r weights-fit}
from_rows <- anova_count(InsectSprays, "count", "spray", plots = FALSE)
from_table <- anova_count(spray_tab, "count", "spray", weights = "Freq",
                          plots = FALSE)
c(rows = from_rows$dispersion, table = from_table$dispersion)
c(rows = from_rows$model_type, table = from_table$model_type)
all.equal(from_rows$anova, from_table$anova)
all.equal(from_rows$effect_sizes, from_table$effect_sizes)
all.equal(from_rows$emmeans, from_table$emmeans)
from_table$notes
```

The dispersion is the same in both, because with frequency weights it divides
by the number of observations the weights represent (minus the number of
parameters), not by the number of rows. Dividing by rows, as `glm()`'s own
residual degrees of freedom would, inflates it:

```{r weights-rows}
pois_table <- glm(count ~ spray, family = poisson, data = spray_tab,
                  weights = Freq)
sum(residuals(pois_table, type = "pearson")^2) / df.residual(pois_table)
```

That inflation could switch a Poisson model to a negative binomial one for no
reason. (Here both values are above the threshold, so the choice would be the
same either way; the correct one, `r round(from_rows$dispersion, 3)`, is only
just above it, and the theta note says the negative binomial has only weakly
overdispersed data to work with.) With `vcov_type`, the sandwich is also
computed as it would be on the expanded data:

```{r weights-robust}
hc_rows <- anova_count(InsectSprays, "count", "spray", vcov_type = "HC3",
                       plots = FALSE)
hc_table <- anova_count(spray_tab, "count", "spray", weights = "Freq",
                        vcov_type = "HC3", plots = FALSE)
all.equal(hc_rows$effect_sizes, hc_table$effect_sizes)
```

One part of the result still describes rows rather than observations:
`$assumptions$cell_counts` counts the rows of `$data_used`, so for the
frequency table it gives the number of distinct counts per spray rather than
the 12 observations behind each.

```{r weights-cells}
from_table$assumptions$cell_counts
```

### `type`, `adjust`, `conf_level` and `posthoc`

`type = "III"` fits the model under sum-to-zero contrasts, which changes the
main-effect rows of `$anova` when there is an interaction; the rate ratios are
unchanged.
`posthoc = FALSE` skips the pairwise comparisons, with a note:

```{r type3}
type3 <- anova_count(warpbreaks, "breaks", c("wool", "tension"),
                     interaction = TRUE, type = "III", posthoc = FALSE,
                     plots = FALSE)
type3$anova
all.equal(type3$effect_sizes, fit$effect_sizes)
grep("posthoc|still reports", type3$notes, value = TRUE)
```

`adjust` sets the multiplicity adjustment for `$posthoc` and `conf_level` the
level of every interval, as in the other model-based functions; the
[binary-outcomes walkthrough](binary.html) shows both.

## What the notes say

```{r notes}
fit$notes
```

- The first note records the model choice: the Poisson dispersion, the
  threshold, and the model fitted instead. It is also how to override it.
- The second gives the negative binomial's theta, the variance function it
  implies, and how far to trust the tests and intervals given the size of the
  smallest group. With 9 looms per cell it recommends quasi-Poisson F tests,
  and in a write-up of these data that is the version of the interaction test
  worth reporting alongside (or instead of) this one.

There is no note on the layout of the marginal means, because with the
interaction in the model they are cells; an additive fit (the insurance model,
for one) adds a note explaining the per-factor layout. Other fits in this
article showed the notes on an explicit Poisson model with overdispersion, on
quasi-Poisson F tests, on exposure, on sparse and empty groups, on robust
standard errors and on frequency weights. The [results article](results.html)
explains how `$notes` fits with the rest of the object.

## Reporting the result

```{r report-inline, include = FALSE}
a <- fit$anova
qp <- fits[[2]]$anova
es <- fit$effect_sizes
fmt_p <- function(p) {
  if (p < 0.001) "< 0.001" else paste("=", sprintf("%.3f", p))
}
```

A write-up might read:

> Warp breaks were analysed with a negative binomial regression on wool,
> tension and their interaction (Pearson dispersion of the Poisson model
> `r sprintf("%.2f", fit$dispersion)`; theta =
> `r sprintf("%.1f", fit$model$theta)`).
> The effect of tension depended on the wool (likelihood-ratio
> χ²(`r a$df[3]`) = `r sprintf("%.2f", a$statistic[3])`, p
> `r fmt_p(a$p_value[3])`; quasi-Poisson F(`r qp$df[3]`, `r qp$df[4]`) =
> `r sprintf("%.2f", qp$statistic[3])`, p `r fmt_p(qp$p_value[3])`). At low
> tension wool B produced fewer breaks than wool A (rate ratio
> `r sprintf("%.2f", es$IRR[1])`, 95% CI `r sprintf("%.2f", es$conf_low[1])`
> to `r sprintf("%.2f", es$conf_high[1])`), and with wool A both medium and
> high tension reduced breaks by about half relative to low tension (rate
> ratios `r sprintf("%.2f", es$IRR[2])` and `r sprintf("%.2f", es$IRR[3])`).

## See also

- `?anova_count` for every argument, and `?anovakit_fit` for the returned
  object.
- [Binary outcomes with `anova_bin()`](binary.html), the companion function
  for yes/no responses.
- [Generalised linear models with `anova_glm()`](glm.html) for other families.
- [Working with results](results.html) and [Diagnostics](diagnostics.html).
- [Plots and visual customisation](visuals.html).
