---
title: "Diagnostics: what anovakit checks and what to do about it"
description: >
  Every assumption check the eight functions run, organised by what is being
  checked: how to read each one, a worked example that makes it fire, and the
  argument or function that helps when it fails.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

```{r helpers, include = FALSE}
# Formatting helpers for numbers quoted in the prose.
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fp <- function(p) if (p < 0.001) "< 0.001" else paste("=", f3(p))
sg <- function(x, d = 2) format(signif(x, d))
note_with <- function(fit, pattern) grep(pattern, fit$notes, value = TRUE)
```

Every anovakit function checks the assumptions its method relies on, and only
those. The results go in `$assumptions`; anything that fails, could not be
computed, or changed what the function did is explained in `$notes`. This
article is organised by *what* is being checked rather than by function, so
that you can see, for example, every way the package looks at normality in one
place. The per-function walkthroughs ([Welch](welch.html),
[Kruskal-Wallis](kruskal-wallis.html), [ANCOVA](ancova.html),
[GLM](glm.html), [repeated measures](repeated-measures.html),
[MANOVA](manova.html), [binary](binary.html) and [counts](counts.html)) show
the same checks in the context of a single analysis.

## What each function stores

The components of `$assumptions` differ by method. Fitting each function once
and listing them is the quickest map:

```{r map}
map <- list(
  anova_welch  = anova_welch(chickwts, "weight", "feed", plots = FALSE),
  anova_kw     = anova_kw(chickwts, "weight", "feed", diagnostics = TRUE,
                          plots = FALSE),
  anova_ancova = anova_ancova(MASS::anorexia, "Postwt", "Treat", "Prewt",
                              plots = FALSE),
  anova_glm    = anova_glm(chickwts, "weight", "feed", plots = FALSE),
  anova_rm     = anova_rm(CO2, "uptake", subject = "Plant", within = "conc",
                          plots = FALSE),
  anova_manova = anova_manova(iris, names(iris)[1:4], "Species", plots = FALSE),
  anova_bin    = anova_bin(MASS::birthwt, "low", "race", plots = FALSE),
  anova_count  = anova_count(warpbreaks, "breaks", "tension", plots = FALSE)
)
data.frame(assumptions = vapply(map, function(f)
  paste(names(f$assumptions), collapse = ", "), character(1)))
```

`anova_kw()` stores nothing unless you ask for `diagnostics = TRUE`, because
the Kruskal-Wallis test assumes no distribution. `anova_glm()` keeps its
coefficient table there, since the intervals in it are where separation shows
up. `anova_rm()` also returns `$sphericity` at the top level, and
`anova_ancova()` and `anova_manova()` return `$slopes_test`. `summary(fit)`
prints every element of `$assumptions` together.

Two habits make all of this easier to use. First, read the tests and the
plots together: a p-value says whether a departure is detectable, a plot says
whether it is large, and neither says whether it matters for your inference.
Second, read `$notes`: the package already says there which check failed and
what it did about it.

## Normality

### Within each group: `anova_welch()` and `anova_kw()`

Welch's test allows each group its own variance, so pooling residuals across
groups would test the wrong thing: a mixture of normal distributions with
different spreads is not normal. `anova_welch()` therefore runs Shapiro-Wilk
separately in each group. `InsectSprays` counts insects on plots treated with
six sprays:

```{r ins-normality}
ins <- anova_welch(InsectSprays, "count", "spray")
ins$assumptions$normality
```

Groups C and D reject normality (p `r fp(ins$assumptions$normality$p_value[3])`
and p `r fp(ins$assumptions$normality$p_value[4])`). The Q-Q plot shows each
observation's deviation from its group mean in units of its group's standard
deviation, so every group contributes on its own scale:

```{r ins-qq}
plot(ins, "qq")
```

The upper tail rises above the line, the mark of right skew, and the short
horizontal runs are tied values: several plots in a group hold the same
number of insects. That is what small counts look like, and it points to the
remedy. For a count response the better model is `anova_count()` (see the
[counts article](counts.html)); for any skewed or ordinal response the
rank-based `anova_kw()` makes no normality assumption:

```{r ins-kw}
anova_kw(InsectSprays, "count", "spray", plots = FALSE)$anova
```

`anova_kw(diagnostics = TRUE)` computes the same per-group table and Q-Q plot,
but only as context, and says so:

```{r kw-note}
map$anova_kw$notes
```

### Why a normality test is a weak guide

A Shapiro-Wilk test answers "is the departure from normality detectable?",
and detectability depends on the sample size far more than on the size of the
departure. The simulation below, whose truth is known, draws two data sets. The
first has 1000 observations per group from a t distribution with 10 degrees of
freedom, whose tails are only slightly heavier than normal; with groups this
large the central limit theorem makes the group means very nearly normal, and
Welch's test keeps its error rate. The second has 8 observations per group from an
exponential distribution, which is strongly skewed, in samples too small for
the central limit theorem to help much.

```{r sim-normality}
set.seed(1)
big <- data.frame(g = rep(c("a", "b", "c"), each = 1000),
                  y = rt(3000, df = 10))
small <- data.frame(g = rep(c("a", "b", "c"), each = 8), y = rexp(24))
fit_big <- anova_welch(big, "y", "g")
fit_small <- anova_welch(small, "y", "g")
fit_big$assumptions$normality[, c("group", "n", "statistic", "p_value")]
fit_small$assumptions$normality[, c("group", "n", "statistic", "p_value")]
```

Every large group rejects, but only
`r sum(fit_small$assumptions$normality$p_value < 0.05)` of the three small,
strongly skewed groups does. The Q-Q plots tell the truer story. On the left
the points follow the line except in the far tails, beyond two standard
deviations; on the right they bow: the lowest values sit above the line, the
middle ones below it and the highest above it again, the shape of right skew.

```{r sim-qq, fig.show = "hold", out.width = "50%", fig.width = 5, fig.height = 4}
plot(fit_big, "qq") + ggplot2::labs(title = "t(10), 1000 per group",
                                    subtitle = NULL)
plot(fit_small, "qq") + ggplot2::labs(title = "Exponential, 8 per group",
                                      subtitle = NULL)
```

One draw could be luck, so here are the rejection rates over 500 repetitions
of each:

```{r sim-rates}
set.seed(2)
reject_rate <- function(draw, reps = 500) {
  mean(replicate(reps, shapiro.test(draw())$p.value < 0.05))
}
rates <- c(t10_n1000 = reject_rate(function() rt(1000, df = 10)),
           exponential_n8 = reject_rate(function() rexp(8)))
rates
```

The test flags the harmless departure `r round(100 * rates[1])`% of the time
and the serious one `r round(100 * rates[2])`% of the time. Use the p-value as
a prompt to look at the plot, not as a verdict, and decide by what the plot
shows and how large the groups are.

### On model residuals: `anova_ancova()` and `anova_glm()`

An ANCOVA assumes one residual variance for every group (Levene's test checks
that, see below), so its residuals can be pooled. `anova_ancova()` runs
Shapiro-Wilk on them and stores the `htest` object. On `MASS::anorexia`,
weight after treatment adjusted for weight before:

```{r anx}
anx <- anova_ancova(MASS::anorexia, "Postwt", "Treat", "Prewt")
anx$assumptions$normality
```

```{r anx-qq}
plot(anx, "qq")
```

Here the test does not reject (p `r fp(anx$assumptions$normality$p.value)`)
and the points stay close to the line. Shapiro-Wilk is skipped, with a note, when
there are fewer than 3 or more than 5000 residuals; above 5000, the Q-Q plot is
all there is, and with that many observations it is also all you need.

`anova_glm()` runs no normality test. For a Gaussian family the residuals are
approximately normal if the model is right, but for other families they are
not expected to be, so a test would reject for the wrong reason. It returns a
Q-Q plot of deviance residuals (`plot(fit, "qq")`), which for counts and
proportions is a rough guide to outlying observations rather than a check of
an assumption. For a Gaussian fit you can run
`shapiro.test(residuals(fit$model))` yourself.

When residual normality fails badly, the remedies are a transformation of the
response (create the column first, since responses are named as columns), a
family that matches the response in `anova_glm()`, or, for a one-way design,
`anova_kw()`.

### Within each within-subject cell: `anova_rm()`

The within-subject F tests of a repeated measures ANOVA depend on the
within-subject errors: what is left of each value after its subject's mean and
its cell's mean are removed. `anova_rm()` stores those in `$residuals` and runs
Shapiro-Wilk on them separately in each within-subject cell, for the same
reason `anova_welch()` tests each group. `CO2` measures carbon dioxide uptake
by 12 plants at seven ambient concentrations:

```{r co2}
co2 <- anova_rm(CO2, "uptake", subject = "Plant", within = "conc",
                between = c("Type", "Treatment"))
co2$assumptions$normality
```

```{r co2-qq}
plot(co2, "qq")
```

```{r co2-count, include = FALSE}
co2_norm <- co2$assumptions$normality
n_rej <- sum(co2_norm$p_value < 0.05)
```

`r n_rej` of the `r nrow(co2_norm)` cells rejects at the 5% level (the cell
at `r co2_norm$cell[co2_norm$p_value < 0.05]`, p
`r fp(co2_norm$p_value[co2_norm$p_value < 0.05])`). With seven tests, one
rejection is about what chance alone produces, so on its own it is weak
evidence; the Q-Q plot, standardised within each cell as the tests are, shows
no systematic curvature.

With a single two-level within-subject factor the two cells' residuals are
plus and minus half of each subject's centred difference, so both rows test
the normality of the differences, which is exactly the assumption of the
equivalent paired t-test. The `sleep` data make that visible:

```{r sleep}
sl <- anova_rm(sleep, "extra", subject = "ID", within = "group",
               plots = FALSE)
sl$assumptions$normality
with(sleep, extra[group == "2"] - extra[group == "1"])
```

The two rows are identical, and both reject
(p `r fp(sl$assumptions$normality$p_value[1])`). The differences themselves
show why: one subject's is
`r with(sleep, max(extra[group == "2"] - extra[group == "1"]))`, far above
the others. With ten subjects, one unusual one is enough to fail the test;
whether to keep it is a question about that subject, not about the test.
Normality of the subject means, which the
between-subject tests rely on, is not tested.

### Multivariate normality: Mardia's tests in `anova_manova()`

A MANOVA assumes the residual vectors are multivariate normal. `anova_manova()`
runs Mardia's tests of multivariate skewness and kurtosis on the residuals of
the multivariate model. For the four flower measurements of `iris`:

```{r iris}
irm <- anova_manova(iris, names(iris)[1:4], "Species")
irm$assumptions$mardia
note_with(irm, "^Mardia")
```

Both reject. The four multivariate statistics differ in how much they suffer
from non-normality, and Pillai's trace (the default, `test = "Pillai"`) is the
most robust of them; the note says so, and that it is the statistic in use.
With another `test`, the note suggests switching to it. Mardia's kurtosis test in particular
rejects too often when the sample is small relative to the square of the
number of responses, and both tests are skipped, with a note, unless the
residual degrees of freedom exceed the number of responses by at least 10
(see [Sparse cells](#sparse-cells-and-cell-counts)). The per-response Q-Q
plots in `$plots` (`qq_Sepal.Length` and so on) show which responses are
responsible; the [plots article](visuals.html) combines them into one figure.

## Equal variances

### The variance ratio: `anova_welch()`

Welch's test does not assume equal variances, so it does not test them either.
It reports the ratio of the largest group variance to the smallest in
`$assumptions$variance_ratio`, and adds a note when it exceeds 4:

```{r ins-ratio}
ins$assumptions$variance_ratio
note_with(ins, "variance")
```

A ratio of `r f2(ins$assumptions$variance_ratio)` is nothing to act on here:
it is the reason to use Welch's test. It matters when you fit the same
comparison with a method that pools the variance. `anova_glm()` with its
default Gaussian family does, and its model-based standard errors for the six
spray means are then identical. `vcov_type = "HC3"` gives each mean a
heteroskedasticity-consistent standard error, and
`test_statistic = "Wald"` makes the omnibus test use it too:

```{r ins-glm}
pooled <- anova_glm(InsectSprays, "count", "spray", plots = FALSE)
robust <- anova_glm(InsectSprays, "count", "spray", vcov_type = "HC3",
                    test_statistic = "Wald", plots = FALSE)
data.frame(spray = pooled$emmeans$spray,
           se_pooled = pooled$emmeans$se,
           se_HC3 = robust$emmeans$se)
note_with(robust, "anova_welch")
```

For a one-way design, `anova_welch()` remains the better calibrated choice,
as the note says.

### Levene's test: `anova_ancova()`

`anova_ancova()` pools the residual variance across groups, so it tests
whether it may: Levene's test on the model residuals across the cells, stored
in `$assumptions$levene`. `MASS::cats` records body and heart weight for 144
cats:

```{r cats}
cats <- anova_ancova(MASS::cats, "Hwt", "Sex", "Bwt", plots = FALSE)
cats$assumptions$levene
note_with(cats, "^Levene")
tapply(residuals(cats$model), cats$data_used$Sex, sd)
```

Levene's test rejects (p `r fp(cats$assumptions$levene$p_value)`): the
residual standard deviation of male heart weights is about
`r f2(tapply(residuals(cats$model), cats$data_used$Sex, sd)[["M"]] / tapply(residuals(cats$model), cats$data_used$Sex, sd)[["F"]])`
times that of females. The note
names the remedy: with `vcov_type = "HC3"` the F tests in `$anova` become Wald
F tests built from a heteroskedasticity-consistent covariance, and the
marginal means, comparisons and covariate slopes use it too.

```{r cats-hc3}
cats_hc3 <- anova_ancova(MASS::cats, "Hwt", "Sex", "Bwt", vcov_type = "HC3",
                         plots = FALSE)
attr(cats_hc3$anova, "statistic")
data.frame(sex = cats$simple_slopes$Sex,
           se_pooled = cats$simple_slopes$se,
           se_HC3 = cats_hc3$simple_slopes$se)
```

The slopes test chose the interaction model here (see
[Homogeneity of regression slopes](#homogeneity-of-regression-slopes)), so
each sex has its own slope. With the pooled variance the female slope's
standard error is overstated and the male one's understated; the robust
covariance corrects both. The homogeneity-of-slopes test itself stays the
model-based F test, as the notes say.

### Box's M: `anova_manova()`

A MANOVA assumes every group has the same covariance matrix. Box's M tests
that:

```{r boxm}
irm$assumptions$box_m
note_with(irm, "^Box")
```

It rejects decisively for `iris`. Box's M is notoriously sensitive to
non-normality, which Mardia's tests have already found, so treat the p-value
as a prompt to look at the group covariances. The standard deviations are a
good start:

```{r boxm-sd}
sapply(split(iris[1:4], iris$Species), function(x) round(apply(x, 2, sd), 2))
```

The petal measurements of setosa vary far less than those of the other two
species. As the note says, with groups of equal size, as here, the
multivariate tests are fairly robust to unequal covariance matrices, and
Pillai's trace most of all. With unequal groups as well that robustness is
weaker: use Pillai's trace and treat a borderline result with caution.

## Homogeneity of regression slopes

### `anova_ancova()`

An ANCOVA compares groups at a common value of the covariate. That comparison
is a single number only if the covariate's slope is the same in every group.
`anova_ancova()` fits the model with and without the covariate-by-group terms
and compares them; the result is in `$slopes_test` (and
`$assumptions$slopes`). In the anorexia data the slopes differ:

```{r anx-slopes}
anx$slopes_test
anx$simple_slopes[, c("Treat", "slope", "conf_low", "conf_high")]
```

In the control group, weight after treatment barely depends on weight before
(slope `r f2(anx$simple_slopes$slope[anx$simple_slopes$Treat == "Cont"])`);
in both treatment groups it does. The covariate plot shows the same thing:

```{r anx-covariate}
plot(anx, "covariate")
```

Because the test rejected at `homogeneity_alpha = 0.05`, the interaction model
was used and the notes say what that means:

```{r anx-notes}
anx$notes
```

The group row of `$anova` and the comparisons in `$posthoc` now compare the
treatments at the mean pre-treatment weight, not by a constant amount. Read
`$simple_slopes` alongside them. Choosing the model with a test on the same
data has a cost in both directions (the last note), so if you know the
design in advance, set `force_interaction = TRUE` (or `FALSE`) instead of
letting the test decide. When the groups were not randomised,
`force_interaction = TRUE` is the safer choice.

Two related checks look at where the covariate lies rather than at its slope.
When the additive model is used and the covariate means differ between groups
by more than half a pooled within-group standard deviation, a note warns that
a slope difference the test missed would bias the adjusted comparisons. And
when the covariate mean lies outside the range a group was observed over, that
group's adjusted mean is an extrapolation, and a note names it. Cars with more
cylinders are heavier, so both fire in `mtcars`:

```{r mt-balance}
mt <- transform(mtcars, cyl = factor(cyl))
mt_anc <- anova_ancova(mt, "mpg", "cyl", "wt", plots = FALSE)
note_with(mt_anc, "covariate means differ|outside the range")
```

No argument can repair covariates that barely overlap between groups; the
adjusted means there rest on the model's straight line rather than on data.
Report them as such, or restrict the comparison to the range the groups share.

### `anova_manova()` with covariates

With `covariates`, `anova_manova()` tests the same assumption with the chosen
multivariate statistic. In `mtcars`, fuel economy and quarter-mile time
depend on weight differently for automatic and manual cars:

```{r mancova}
mt$am <- factor(mt$am, labels = c("automatic", "manual"))
mcv <- anova_manova(mt, c("mpg", "qsec"), "am", covariates = "wt",
                    plots = FALSE)
mcv$slopes_test
note_with(mcv, "^Homogeneity")
```

The common-slope adjusted means then describe the groups at the covariate
mean only, and rejections by Mardia's tests or Box's M may be artefacts of the
misspecified slopes. `anova_manova()` does not fit the covariate-by-group
model itself; analyse each response with `anova_ancova()`, which does:

```{r mancova-ancova}
anova_ancova(mt, "mpg", "am", "wt", force_interaction = TRUE,
             plots = FALSE)$simple_slopes[, c("am", "slope", "conf_low",
                                              "conf_high")]
```

## Sphericity

The univariate F tests of a repeated measures ANOVA assume sphericity: that the
differences between every pair of within-subject levels have the same
variance. `anova_rm()` reads Mauchly's test and the Greenhouse-Geisser and
Huynh-Feldt epsilons from the fitted model and stores them in `$sphericity`
(and `$assumptions$sphericity`), one row per within-subject term with more
than one degree of freedom:

```{r co2-sph}
co2$sphericity
```

Mauchly's test rejects (p `r fp(co2$sphericity$p_value[1])`). Epsilon measures
how far the data are from sphericity: 1 means none, and its lower bound is
1/(k - 1), here `r f3(1 / 6)` for seven concentrations. The Greenhouse-Geisser
estimate of `r f2(co2$sphericity$gg_epsilon[1])` is well below 1; the
Huynh-Feldt estimate of `r f2(co2$sphericity$hf_epsilon[1])` is less
conservative. The correction multiplies both degrees of freedom of each
affected F test by epsilon. `correction` chooses which one `$anova` uses:

```{r co2-corrections}
by_corr <- lapply(c(GG = "GG", HF = "HF", none = "none"), function(cr) {
  anova_rm(CO2, "uptake", subject = "Plant", within = "conc",
           between = c("Type", "Treatment"), correction = cr, plots = FALSE)
})
do.call(rbind, lapply(names(by_corr), function(cr) {
  row <- by_corr[[cr]]$anova
  row <- row[row$term == "Treatment:conc", ]
  data.frame(correction = cr, num_df = row$num_df, den_df = row$den_df,
             p_value = row$p_value)
}))
note_with(by_corr$none, "^Mauchly")
```

The default, `"GG"`, is the safe choice when epsilon is well below 1; `"HF"`
is more powerful when it is close to 1; `"none"` is for when you have reason
to believe sphericity holds, and the note above warns when Mauchly's test
disagrees.
Mauchly's test has little power in small samples, so it is better to apply a
correction routinely than to decide by the test. The marginal means and
comparisons do not need a correction: they come from afex's multivariate
model, which gives each within-subject cell its own variance.

With fewer subjects than contrasts, the error matrix is singular and Mauchly's
test is undefined. The epsilons are still defined, and `anova_rm()` computes
them directly and says so. The six Quebec plants alone are too few for seven
concentrations:

```{r quebec}
quebec <- droplevels(subset(CO2, Type == "Quebec"))
qf <- anova_rm(quebec, "uptake", subject = "Plant", within = "conc",
               between = "Treatment", plots = FALSE)
qf$sphericity[, c("term", "mauchly_w", "gg_epsilon", "hf_epsilon")]
note_with(qf, "^car could not")
```

When every within-subject factor has two levels there is only one difference,
so sphericity holds automatically. The `sleep` fit above is an example:

```{r sleep-sph}
is.null(sl$sphericity)
attr(sl$anova, "correction")
sl$notes
```

The table records that no correction was applied, whatever `correction`
asks for, and `print()` labels it that way.

## Dispersion

### Counts: `anova_count()`

A Poisson model assumes the variance equals the mean. The Pearson dispersion
statistic, the sum of squared Pearson residuals divided by the residual
degrees of freedom, is about 1 when that holds. `anova_count()` computes it
from a Poisson fit and, by default (`model = "auto"`), moves to a negative
binomial model when it exceeds `overdispersion_threshold` (1.5). `warpbreaks`
counts the breaks in lengths of yarn woven under three tensions:

```{r warp}
wb <- anova_count(warpbreaks, "breaks", c("wool", "tension"), plots = FALSE)
c(poisson = wb$dispersion, fitted_model = wb$model_dispersion)
wb$model_type
note_with(wb, "dispersion")
```

The Poisson dispersion of `r f2(wb$dispersion)` means the variance is about
four times the mean. The negative binomial model has a dispersion of
`r f2(wb$model_dispersion)`, close to 1, as it should be. The note on theta
also says how well the negative binomial tests are calibrated at this sample
size. `$dispersion` (the statistic the choice was made on) and
`$model_dispersion` (that of the model returned) are also in `$assumptions`,
as `poisson_dispersion` and `model_dispersion`.

Forcing `model = "poisson"` keeps the Poisson model and says what that costs:

```{r warp-poisson}
note_with(anova_count(warpbreaks, "breaks", c("wool", "tension"),
                      model = "poisson", plots = FALSE), "^Pearson")
```

The threshold is a fixed cut-off, not a test, and a dispersion near it is
worth a second look whichever side it falls:

```{r ins-count}
ins_nb <- anova_count(InsectSprays, "count", "spray", plots = FALSE)
ins_nb$dispersion
note_with(ins_nb, "^Negative binomial")
ins_p <- anova_count(InsectSprays, "count", "spray",
                     overdispersion_threshold = 2, plots = FALSE)
note_with(ins_p, "^Pearson")
```

A dispersion of `r f2(ins_nb$dispersion)` is just over the default threshold,
so the default moved to a negative binomial, and the theta note reports that
theta is poorly determined but large: the data are only weakly overdispersed.
With the threshold raised to 2 the Poisson model is kept, and the note then
gives the Pearson test of overdispersion and what a dispersion of that size
does to the Poisson tests. Set `model` yourself (`"poisson"`, `"negbin"` or
`"quasipoisson"`) to take the decision out of the function's hands.

### Other families: `anova_glm()`

`anova_glm()` fits the family you give it and does not switch. It reports the
Pearson dispersion in `$assumptions$dispersion` and, for the binomial and
Poisson families, whose dispersion is fixed at 1, adds a note when the Pearson
test finds it above 1 beyond chance:

```{r glm-pois}
gp <- anova_glm(warpbreaks, "breaks", c("wool", "tension"),
                family = "poisson", plots = FALSE)
note_with(gp, "^Pearson")
```

The remedies are the ones the note names. `family = "quasipoisson"` estimates
the dispersion and scales every standard error and test by it (the omnibus
table becomes an F test); `anova_count(model = "negbin")` models the extra
variation explicitly.

The same check applies to binomial proportions with several trials per row.
`esoph` records oesophageal cancer cases and controls by age, alcohol and
tobacco group; modelled as proportions with the number of subjects as weights:

```{r esoph}
es <- transform(esoph, n = ncases + ncontrols,
                prop = ncases / (ncases + ncontrols))
e2 <- anova_glm(es, "prop", c("agegp", "alcgp"), family = "binomial",
                weights = "n", plots = FALSE)
e2$assumptions$dispersion
note_with(e2, "^Pearson")
```

Overdispersion is often a sign of a missing predictor rather than of a wrong
distribution. Adding tobacco use removes it:

```{r esoph-tob}
e3 <- anova_glm(es, "prop", c("agegp", "alcgp", "tobgp"), family = "binomial",
                weights = "n", plots = FALSE)
e3$assumptions$dispersion
note_with(e3, "^Pearson")
```

The dispersion falls to `r f2(e3$assumptions$dispersion)` and the note is
gone. If no missing term explains it, `family = "quasibinomial"` is the
remedy the note suggests.

### Why it is `NA` for 0/1 data

For a 0/1 response, `anova_bin()` (and `anova_glm()` with a binomial family)
reports the dispersion as `NA`. A single Bernoulli trial cannot be
overdispersed: its variance is fixed by its mean. And the Pearson statistic
carries no information: in a model with one parameter per cell it equals
N / (N - k) whatever the data. `MASS::birthwt` shows it, with a real outcome
and with pure noise:

```{r birthwt-disp}
bw <- transform(MASS::birthwt,
                race = factor(race, labels = c("white", "black", "other")),
                smoke = factor(smoke, labels = c("no", "yes")))
set.seed(3)
bw$noise <- rbinom(nrow(bw), 1, 0.3)
pearson <- function(f) {
  sum(residuals(f$model, type = "pearson")^2) / df.residual(f$model)
}
real <- anova_bin(bw, "low", c("race", "smoke"), interaction = TRUE,
                  plots = FALSE)
fake <- anova_bin(bw, "noise", c("race", "smoke"), interaction = TRUE,
                  plots = FALSE)
c(low_birth_weight = pearson(real), noise = pearson(fake),
  N_over_N_minus_k = nrow(bw) / (nrow(bw) - 6))
real$assumptions$dispersion
```

All three numbers are the same. If your binary data come in clusters
(patients within clinics, say), the concern is correlation within clusters,
which no dispersion statistic of 0/1 data can reveal; it needs a model for the
clustering.

## Separation and zero-count cells

### Binary responses

When every observation in a group has the same outcome, the maximum likelihood
estimate of that group's log-odds is infinite. `glm()` stops at a large
number and reports convergence anyway; the coefficient's standard error is
enormous, its Wald p-value is near 1 however strong the effect, and its Wald
interval covers everything. `anova_bin()`, and `anova_glm()` for binomial,
Poisson and negative binomial families, check the fitted linear predictor of
every cell and name the cells that have run away.

Every child in first and second class on the Titanic survived:

```{r titanic-table}
kids <- subset(as.data.frame(Titanic), Age == "Child")
xtabs(Freq ~ Class + Survived, kids)
```

With third class as the reference, the odds ratios for first and second class
are infinite:

```{r titanic}
tk <- anova_bin(kids, "Survived", "Class", weights = "Freq",
                reference = list(Class = "3rd"), plots = FALSE)
tk$notes
```

The eight rows with a frequency of zero (the crew's, and those for first- and
second-class children who died, of whom there were none) are dropped first. The separation note names the
cells and the coefficients. The next note describes what
`anova_bin()` does about the odds ratios: their profile-likelihood intervals
are open on the side the estimate ran off to, and the finite end is found by
profiling the likelihood directly:

```{r titanic-or}
tk$effect_sizes[, c("comparison", "odds_ratio", "conf_low", "conf_high",
                    "p_value")]
```

Read the estimates as "infinite" and the intervals as what the data support:
by the `r 100 * tk$conf_level`% profile intervals, the odds of survival for a
first-class child were at least `r f2(tk$effect_sizes$conf_low[1])` times those
of a third-class child, and for a second-class child at least
`r f2(tk$effect_sizes$conf_low[2])` times. The
Wald p-values beside them are useless, as the note warns. The separation note
also warns that the same goes for the separated cells' marginal
probabilities and the comparisons involving them: the intervals in `$emmeans`
and `$posthoc` are Wald intervals and get no profile treatment.

```{r titanic-emm}
tk$emmeans[, c("Class", "estimate", "conf_low", "conf_high")]
tk$posthoc[, c("contrast", "ratio", "conf_low", "conf_high", "p_adjusted")]
```

What to do:

- The likelihood-ratio omnibus test in `$anova` is still usable, although the
  sparse-data note warns that it is liberal with events this rare. For one
  grouping variable an exact test is better:
  `fisher.test(xtabs(Freq ~ Class + Survived, droplevels(subset(kids, Class != "Crew"))))`
  gives p = `r sg(fisher.test(xtabs(Freq ~ Class + Survived, droplevels(subset(kids, Class != "Crew"))))$p.value)`.
- Report the one-sided profile intervals from `$effect_sizes`, not the Wald
  quantities.
- Collapse the offending level into a neighbour if that makes substantive
  sense, or use a penalised fit such as `logistf::logistf()`.
- Do not ask for robust standard errors as a fix. At the boundary the sandwich
  covariance collapses towards zero, so `vcov_type` falls back to model-based
  standard errors there, with a note:

```{r titanic-hc3}
note_with(anova_bin(kids, "Survived", "Class", weights = "Freq",
                    vcov_type = "HC3", plots = FALSE), "^Robust")
```

### Counts

The count version of separation is a group with no events: its log rate is
minus infinity. The simulation below, whose truth is known, has a third site
whose true rate is small but not zero, and whose ten traps happened to catch
nothing:

```{r traps}
set.seed(10)
traps <- data.frame(site = factor(rep(c("A", "B", "C"), each = 10)))
traps$n <- rpois(30, c(A = 4, B = 2, C = 0.1)[as.character(traps$site)])
tapply(traps$n, traps$site, sum)
tr <- anova_count(traps, "n", "site", plots = FALSE)
note_with(tr, "^No events|fewer than 5 events|profiling")
tr$effect_sizes[, c("comparison", "IRR", "conf_low", "conf_high", "p_value")]
```

The rate ratio for site C against A is numerically zero and its Wald p-value
is near 1, as the note warns. Its interval, though, is one-sided: the lower
end is 0, and the upper end was found by profiling the likelihood directly,
where a Wald interval would run from 0 to infinity. So, by the
`r 100 * tr$conf_level`% profile interval, site C's rate is at most
`r f3(tr$effect_sizes$conf_high[tr$effect_sizes$comparison == "C vs A"])`
times site A's. `anova_glm()` with a Poisson family profiles the diverged
coefficient in its coefficient table the same way and finds the same end:

```{r traps-glm}
trg <- anova_glm(traps, "n", "site", family = "poisson", plots = FALSE)
co <- trg$assumptions$coefficients
exp(co[co$term == "siteC", c("conf_low", "conf_high")])
```

The likelihood-ratio omnibus test remains usable. If a zero group is a
structural zero (a species that cannot occur there), drop it; if not, report
the one-sided interval, or collapse the level.

## Sparse cells and cell counts

Many checks are about how much information each cell holds. The package
reports it in several ways.

**Expected events in a binary model.** When fewer than about 5 events or
non-events are expected in a cell under the null hypothesis, `anova_bin()`
warns that the likelihood-ratio test is liberal. Low birth weight by race and
smoking, with the interaction, has three such cells:

```{r bw-sparse}
note_with(real, "^Sparse")
```

The note gives the remedies: an exact test for one grouping variable, or the
score test, `anova(fit$model, test = "Rao")`. Fitting the grouping variables
additively (`interaction = FALSE`, the default) pools information across cells
and judges sparseness on each factor's levels instead.

**Events and empty combinations in a count model.** `anova_count()` notes
cells with fewer than 5 events in total (the trap example above has one), and
combinations of levels with no observations. Its
`$assumptions$cell_counts` gives the number of rows in each populated cell. In
`MASS::quine`, which records days absent from school, no slow learner is in
the oldest age group:

```{r quine}
qn <- anova_count(MASS::quine, "Days", c("Age", "Lrn"), interaction = TRUE,
                  plots = FALSE)
qn$assumptions$cell_counts
note_with(qn, "possible group|aliased|contain no data")
```

With the interaction, the empty cell's parameter cannot be estimated, it is
left out of `$emmeans` and `$posthoc`, and the interaction is tested on
`r qn$anova$df[qn$anova$term == "Age:Lrn"]` degrees of freedom instead of 3.
The additive model estimates every combination from the main effects, and the
note it gives instead says what that relies on:

```{r quine-add}
note_with(anova_count(MASS::quine, "Days", c("Age", "Lrn"), plots = FALSE),
          "possible group")
```

**Small groups for normality tests.** Shapiro-Wilk needs at least 3 values,
and at least 3 distinct ones to mean anything. `anova_welch()` names any group
it could not test, with the reason, both in the table's `note` column and in
`$notes`. In `mtcars`, by cylinders and transmission:

```{r mt-small}
mw <- anova_welch(mtcars, "mpg", c("cyl", "am"), plots = FALSE)
mw$assumptions$normality[, c("group", "n", "p_value", "note")]
note_with(mw, "^Shapiro")
```

The group with two cars is skipped, and so is a group of three cars, two of
which have the same fuel economy. Run on those three values anyway, the test
shows why:

```{r mt-small-sw}
six_manual <- mtcars$mpg[mtcars$cyl == 6 & mtcars$am == 1]
six_manual
shapiro.test(six_manual)[c("statistic", "p.value")]
```

W takes its smallest possible value for three observations and the p-value is
exactly `r shapiro.test(six_manual)$p.value`: a "rejection" produced by the tie,
not by the shape of any distribution. `anova_kw()` with
`diagnostics = TRUE` and `anova_rm()` name the groups and within-subject cells
they could not test in the same way.

**Small groups for the multivariate checks.** Box's M needs more observations
in every group than there are responses, and Mardia's tests need the residual
degrees of freedom to exceed the number of responses by at least 10.
`anova_manova()` stores `$assumptions$cell_counts` so you can see why. With
four flowers per species:

```{r iris-small}
tiny <- iris[c(1:4, 51:54, 101:104), ]
ms <- anova_manova(tiny, names(iris)[1:4], "Species", plots = FALSE)
ms$assumptions$cell_counts
note_with(ms, "^Mardia|^Box")
```

Neither check can be computed; more data, or fewer responses, are the only
remedies.

## Leverage and robust standard errors

A heteroskedasticity-consistent covariance estimates each observation's
contribution to the variance from its own residual. An observation that is
alone in its cell has leverage 1: the model fits it exactly, its residual is
zero by construction, and the sandwich would give that cell a standard error
of zero (HC0, HC1) or none at all (HC2 to HC4). `vcov_type` then falls back to
model-based standard errors, with a note. In `mtcars`, by cylinders and gears
with their interaction, two cells hold one car each:

```{r lev}
mt$gear <- factor(mt$gear)
lev <- anova_glm(mt, "mpg", c("cyl", "gear"), interaction = TRUE,
                 vcov_type = "HC3", plots = FALSE)
lev$data_used[hatvalues(lev$model) > 1 - 1e-8, c("cyl", "gear", "mpg")]
note_with(lev, "^Robust")
```

To get robust standard errors, give every cell more than one observation: fit
the grouping variables additively (the default for `anova_glm()`), collapse
sparse levels, or, for a one-way comparison, use `anova_welch()`. The additive
fit uses the robust covariance without complaint:

```{r lev-add}
lev_add <- anova_glm(mt, "mpg", c("cyl", "gear"), vcov_type = "HC3",
                     plots = FALSE)
any(grepl("were not used", lev_add$notes))
```

`vcov_type` also falls back at the boundary of the parameter space
(separation, or a group with no events, as above), when the sandwich package is
not installed, and when the robust covariance cannot be computed. In every case
`$notes` says which standard errors were used.

## Aliased coefficients and saturated models

### Aliased coefficients

An empty combination of levels in a model with interactions, or a covariate
that is a linear combination of others, leaves a coefficient that cannot be
estimated. The same `mtcars` fit has no eight-cylinder car with four gears:

```{r alias}
note_with(lev, "aliased|contain no data")
lev$anova
```

Under Type II (the default for `anova_glm()`), the table is computed and the
interaction is tested on
`r lev$anova$df[lev$anova$term == "cyl:gear"]` degrees of freedom instead of 4.
Type III tests are not defined when coefficients are aliased, and
`car::Anova()` refuses them. Ask for them anyway and the function computes
Type II tests instead, says so, and labels the result accordingly:

```{r alias-iii}
lev3 <- anova_glm(mt, "mpg", c("cyl", "gear"), interaction = TRUE,
                  type = "III", plots = FALSE)
lev3$method
attr(lev3$anova, "ss_type")
note_with(lev3, "Type III tests are not defined")
```

`anova_bin()`, `anova_count()` and `anova_ancova()` do the same, and so does
`anova_manova()`, for its multivariate table and its univariate follow-ups.
That matters most for `anova_ancova()`, which uses Type III by default: an
ANCOVA with an empty cell gets Type II tests without being asked.

```{r alias-ancova}
anova_ancova(mt, "mpg", c("cyl", "gear"), "wt", plots = FALSE)$method
# Days absent (no slow learner is in age group F3), and the Titanic (no
# child was crew): both ask for Type III and get Type II
c(anova_count = attr(anova_count(MASS::quine, "Days", c("Age", "Lrn"),
                                 interaction = TRUE, type = "III",
                                 plots = FALSE)$anova, "ss_type"),
  anova_bin = attr(anova_bin(as.data.frame(Titanic), "Survived",
                             c("Class", "Age"), interaction = TRUE,
                             type = "III", weights = "Freq",
                             plots = FALSE)$anova, "ss_type"))
```

Under Type II a main effect is tested ignoring the interaction terms that
contain it, which is a different hypothesis from the Type III one, so report
which type you have. The other remedies are to drop the interaction
(`interaction = FALSE`) or to analyse only the combinations that exist, for
example by combining the grouping variables into one factor of the populated
cells.

### Saturated models

A model with as many parameters as observations fits the data exactly. What
that costs depends on the family. For a Gaussian model there is no residual
variance left to test against, so nothing can be tested. The mean number of
warp breaks for each wool and tension is one number per cell:

```{r saturated}
cell_means <- aggregate(breaks ~ wool + tension, data = warpbreaks, FUN = mean)
sat <- anova_glm(cell_means, "breaks", c("wool", "tension"),
                 interaction = TRUE, plots = FALSE)
note_with(sat, "saturated")
sat$anova
```

Analyse the individual observations instead, or fit fewer terms.

For a family with a fixed dispersion (binomial, Poisson) the likelihood-ratio
tests remain valid: a saturated log-linear model is how a G-test of
independence is computed. Only goodness of fit and overdispersion become
uncheckable. Hair and eye colour of 592 students:

```{r haireye}
he <- as.data.frame(margin.table(HairEyeColor, c(1, 2)))
sat_pois <- anova_count(he, "Freq", c("Hair", "Eye"), interaction = TRUE,
                        plots = FALSE)
sat_pois$notes
sat_pois$anova
sat_pois$dispersion
```

The `Hair:Eye` row is the test of independence. `$dispersion` is `NA`: there
are no residual degrees of freedom to compute it from.

## Summary

| Check | Function(s) | Component | When it fails |
|:------|:------------|:----------|:--------------|
| Normality within groups | `anova_welch()`; `anova_kw()` with `diagnostics = TRUE` (context only) | `$assumptions$normality`, `$plots$qq` | Look at the Q-Q plot; `anova_kw()`; `anova_count()` for counts; transform the response |
| Normality of residuals | `anova_ancova()` (test); `anova_glm()` (Q-Q plot only) | `$assumptions$normality`, `$plots$qq` | Transform the response; a better-matched `family` in `anova_glm()` |
| Normality within each within-subject cell | `anova_rm()` | `$assumptions$normality`, `$residuals`, `$plots$qq` | Transform the response; judge by the plot when only one cell of many rejects |
| Multivariate normality (Mardia) | `anova_manova()` | `$assumptions$mardia` | `test = "Pillai"` (the default); inspect `$plots$qq_<response>` |
| Variance ratio | `anova_welch()` | `$assumptions$variance_ratio` | Nothing for Welch; in pooled-variance models use `vcov_type = "HC3"` |
| Levene's test | `anova_ancova()` | `$assumptions$levene` | `vcov_type = "HC3"` |
| Box's M | `anova_manova()` | `$assumptions$box_m` | Inspect group covariances; `test = "Pillai"`, especially with unequal groups |
| Homogeneity of regression slopes | `anova_ancova()`, `anova_manova()` with covariates | `$slopes_test` (`$assumptions$slopes` in ANCOVA), `$simple_slopes` | `force_interaction = TRUE`; read `$simple_slopes`; `anova_ancova()` per response after a MANCOVA |
| Covariate balance and extrapolation | `anova_ancova()` | `$notes`, `$covariate_means`, `$plots$covariate` | `force_interaction = TRUE`; restrict to the shared covariate range |
| Sphericity | `anova_rm()` | `$sphericity`, `$assumptions$sphericity` | `correction = "GG"` (default) or `"HF"` |
| Overdispersion (counts) | `anova_count()` | `$dispersion`, `$model_dispersion`, `$assumptions$poisson_dispersion` | `model = "negbin"` or `"quasipoisson"`; `overdispersion_threshold` |
| Overdispersion (other families) | `anova_glm()` | `$assumptions$dispersion` | `family = "quasipoisson"` or `"quasibinomial"`; `anova_count(model = "negbin")`; a missing predictor |
| Dispersion of 0/1 data | `anova_bin()`, `anova_glm()` | `$assumptions$dispersion` (`NA`) | Nothing to check; model clustering if there is any |
| Separation | `anova_bin()`, `anova_glm()` | `$notes`, `$effect_sizes` or `$assumptions$coefficients` (open intervals) | Exact test; one-sided profile intervals; collapse levels; `logistf::logistf()` |
| Zero-count cells | `anova_count()`, `anova_glm()` | `$notes`, `$effect_sizes`, `$assumptions$coefficients` (one-sided profile intervals) | Report the one-sided interval; collapse levels or drop structural zeros |
| Sparse cells | `anova_bin()`, `anova_count()`, `anova_manova()`, `anova_welch()`, `anova_kw()`, `anova_rm()` | `$notes`, `$assumptions$cell_counts`, `$assumptions$proportions` | `interaction = FALSE`; collapse levels; exact or score tests |
| Leverage 1 | `anova_ancova()`, `anova_glm()`, `anova_bin()`, `anova_count()` (with `vcov_type`) | `$notes` | Additive model; collapse levels; `anova_welch()` for one-way |
| Aliased coefficients | `anova_glm()`, `anova_ancova()`, `anova_bin()`, `anova_count()`, `anova_manova()` | `$notes`, `attr(fit$anova, "ss_type")` | Type III falls back to Type II by itself; `interaction = FALSE`; analyse populated cells only |
| Saturated model | `anova_glm()`, `anova_ancova()`, `anova_bin()`, `anova_count()`, `anova_manova()` | `$notes` | Fewer terms or more data (Gaussian); fine for tests with a fixed dispersion |

## See also

- [Which function?](choosing.html) for choosing a method in the first place.
- [Working with results](results.html) for `$notes` and the rest of the
  returned object.
- [Plots](visuals.html) for every diagnostic plot and how to customise it.
- The per-function articles: [Welch](welch.html),
  [Kruskal-Wallis](kruskal-wallis.html), [ANCOVA](ancova.html),
  [GLM](glm.html), [repeated measures](repeated-measures.html),
  [MANOVA](manova.html), [binary](binary.html), [counts](counts.html).
