---
title: "Analysis of deviance: factorial designs and non-normal responses"
description: >
  A walkthrough of anova_glm(): a two-way Gaussian ANOVA with Type II and
  Type III tests, then a Gamma model for a positive, right-skewed response,
  with its coefficient table, dispersion, robust standard errors and weights.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

```{r helpers, include = FALSE}
# Formatting helpers for numbers quoted in the prose.
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fp <- function(p) if (p < 0.001) "< 0.001" else paste("=", f3(p))
```

`anova_glm()` fits a generalised linear model of a response on one or more
grouping factors, in any `family`, and reports an analysis-of-deviance table,
effect sizes, estimated marginal means and pairwise comparisons. With the
default Gaussian family it is an ordinary factorial analysis of variance.
Other families model responses whose spread changes with their mean. It is the
general tool. For a one-way comparison of means with unequal variances,
`anova_welch()` is better calibrated ([walkthrough](welch.html)). For 0/1
outcomes use `anova_bin()` ([walkthrough](binary.html)), and for counts use
`anova_count()` ([walkthrough](counts.html)): both add diagnostics and effect
sizes that this function does not. To adjust for a numeric covariate use
`anova_ancova()` ([walkthrough](ancova.html)).

## The data

`ToothGrowth` records the length of odontoblasts (the cells responsible for
tooth growth) in 60 guinea pigs. Each animal received vitamin C at one of three
doses (0.5, 1 or 2 mg/day) by one of two delivery methods: orange juice (`OJ`)
or ascorbic acid (`VC`). `dose` is stored as a number, so make it a factor:
it is a grouping variable here, with three levels.

```{r data}
tg <- ToothGrowth
tg$dose <- factor(tg$dose)
head(tg)
table(tg$supp, tg$dose)
round(tapply(tg$len, list(tg$supp, tg$dose), mean), 2)
```

This is a balanced 2 x 3 factorial design with ten animals per cell. The
interesting question is not just whether dose and supplement matter, but
whether the effect of the supplement depends on the dose. The cell means
suggest it does: orange juice is ahead at the two lower doses, but not at
2 mg/day.

## Fitting the model

Columns are named as character strings. With `interaction = TRUE` the model
contains both factors and their interaction. Listing `dose` first puts it on
the x axis of the marginal-means plot.

```{r fit}
fit <- anova_glm(tg, "len", c("dose", "supp"), interaction = TRUE)
fit
```

Reading the printout from the top:

- The first line names the family, the link and the type of sums of
  squares. A Gaussian family with the identity link is an ordinary linear
  model.
- `Call:` records how the fit was made, and `Observations used: 60` says no
  rows were dropped.
- The heading `Omnibus test (Type II, F)` names the type of sums of squares
  and the test statistic. Both change what the numbers mean, and neither can
  be read off the table itself.
- There is no `Notes` block, because nothing needed saying. See
  [What the notes say](#what-the-notes-say).
- `Plots available` lists the four plots in `$plots`.

`summary(fit)` adds the assumption checks (the dispersion and the coefficient
table), the effect sizes, the marginal means and the pairwise comparisons.

## The omnibus test

```{r anova}
fit$anova
```

- `term` names the effect. `dose:supp` is the interaction.
- `sum_sq` and `df` are the sum of squares and degrees of freedom of each term.
  The `Residuals` row holds the error sum of squares and its degrees of
  freedom.
- `statistic` is the F statistic, the term's mean square divided by the
  residual mean square.
- `p_value` is its p-value.

The interaction is significant (F = `r f2(fit$anova$statistic[3])`,
p `r fp(fit$anova$p_value[3])`): the supplement effect is not the same at
every dose. When an interaction is present, a main-effect row averages over a
pattern that changes, so read the cell means below rather than the main
effects alone.

### Type II or Type III

With `type = "II"` (the default) each main effect is tested after the other
main effect, but not after the interaction. With `type = "III"` every term is
tested after every other, interaction included. In a balanced design like this
one the two coincide:

```{r type-balanced}
fit3 <- anova_glm(tg, "len", c("dose", "supp"), interaction = TRUE,
                  type = "III", plots = FALSE)
fit3$anova
```

They part company when the cells are unequal. Suppose four animals had been
lost from the `VC`, 2 mg cell and two from the `OJ`, 0.5 mg cell:

```{r type-unbalanced}
lost <- c(which(tg$supp == "VC" & tg$dose == "2")[1:4],
          which(tg$supp == "OJ" & tg$dose == "0.5")[1:2])
tg_u <- tg[-lost, ]
table(tg_u$supp, tg_u$dose)
anova_glm(tg_u, "len", c("dose", "supp"), interaction = TRUE,
          plots = FALSE)$anova
anova_glm(tg_u, "len", c("dose", "supp"), interaction = TRUE,
          type = "III", plots = FALSE)$anova
```

The interaction row is the same in both, because it is adjusted for
everything either way. The main-effect rows differ. Type II is the more
powerful test of a main effect when there is no interaction. Type III tests
each main effect averaged over the levels of the other factor, weighting each
cell equally. Some analysts prefer that when an interaction is present, since
it does not depend on the cell sizes.

**Why Type III is fitted under sum-to-zero contrasts.** A Type III test of
`supp` removes the columns that code `supp` while keeping the interaction. What
that tests depends on how the factors are coded. Under R's default treatment
contrasts, the `supp` columns describe the supplement effect at the reference
dose only, and the "Type III" row for `supp` becomes a test of the supplement
at 0.5 mg/day. Under sum-to-zero contrasts it is the supplement effect averaged
over the doses, which is what a main effect is supposed to be. So when you ask
for `type = "III"`, `anova_glm()` fits the model under `contr.sum`. Setting
`options(contrasts = )` afterwards would not help, because the contrasts are
stored on the fitted model. Here is what you would get by fitting with the
defaults and asking `car::Anova()` for Type III:

```{r naive-type3}
naive <- glm(len ~ dose * supp, data = tg)    # treatment contrasts
naive_tab <- car::Anova(naive, type = 3, test.statistic = "F")
naive_tab
unlist(fit3$model$contrasts)
```

The `supp` row has F = `r f2(naive_tab["supp", "F values"])`, the test of the
supplement at the reference dose, not the
F = `r f2(fit3$anova$statistic[2])` of the averaged main effect. The data are
the same, and the table carries the same label. The contrasts stored on
`fit3$model` show how `anova_glm()` avoids this.

## Effect sizes

For a Gaussian model `$effect_sizes` holds partial eta squared and partial
omega squared for each term, computed from the sums of squares of the table
above:

```{r effect-sizes}
fit$effect_sizes
```

- `partial_eta_sq` is the term's sum of squares divided by itself plus the
  residual sum of squares: the share of the variance not explained by the
  other terms that this term explains. Dose accounts for
  `r round(100 * fit$effect_sizes$partial_eta_sq[1])`% of it.
- `partial_omega_sq` is the same idea with a correction for the upward bias of
  eta squared in samples, so it is a little smaller. A negative value
  is floored at 0, with a note.

Both are *partial*: they do not add up to 1 across terms, and they should not
be read as a partition of the total variance.

## Marginal means and comparisons

Because the model contains the interaction, the estimated marginal means are
for the six cells. For a Gaussian model with the identity link they are the
cell means, with intervals that use the pooled residual variance:

```{r emmeans}
fit$emmeans
```

`$posthoc` compares every pair of cells, 15 in all, with a Tukey adjustment by
default. The three comparisons of the supplements at the same dose are:

```{r posthoc}
same_dose <- c("dose0.5 OJ - dose0.5 VC", "dose1 OJ - dose1 VC",
               "dose2 OJ - dose2 VC")
ph <- fit$posthoc
print(ph[ph$contrast %in% same_dose, ], digits = 3)
```

`estimate` is the first cell minus the second. `conf_low` and `conf_high` are
Tukey-adjusted (simultaneous) intervals. `p_value` is unadjusted, and
`p_adjusted` is adjusted across all 15 comparisons. Orange juice gives longer
odontoblasts than ascorbic acid at 0.5 and 1 mg/day, and the two are
indistinguishable at 2 mg/day.

If only those three comparisons interest you, adjusting for all 15 is more
conservative than it needs to be. `$emmeans_object` is the **emmeans** grid, so
you can ask for exactly the family you want. It carries **emmeans**' own default
level, so give `level =` explicitly:

```{r emm-object}
by_dose <- emmeans::contrast(fit$emmeans_object, method = "pairwise",
                             by = "dose")
summary(by_dose, infer = TRUE, level = 0.95)
```

With one comparison per dose there is nothing to adjust within each family.
The [results article](results.html) covers `$emmeans_object` further.

## Checking assumptions

A Gaussian model assumes independent observations, normally distributed
errors and the same error variance in every cell. `$assumptions` holds two
elements, and the residual plots in the next section do the rest.

```{r dispersion}
fit$assumptions$dispersion
sqrt(fit$assumptions$dispersion)
```

`dispersion` is the Pearson dispersion: the sum of squared Pearson residuals
divided by the residual degrees of freedom. For a Gaussian model that is the
residual mean square, so its square root,
`r f2(sqrt(fit$assumptions$dispersion))`, is the residual standard deviation
in the units of the response.

```{r coefficients}
fit$assumptions$coefficients
```

`coefficients` is the coefficient table, with t statistics (the dispersion is
estimated) and profile-likelihood intervals by default, which for a Gaussian
model equal `confint(lm())`. Under the default treatment contrasts the
intercept is the mean of the reference cell (0.5 mg/day, orange juice), and
the other coefficients are differences from it. `fit3`, fitted under
sum-to-zero contrasts, has different coefficients describing the same fitted
values. There the intercept is the average of the six cell means, `dose1` is
the average for 0.5 mg/day minus that grand mean, and `supp1` is the average
for orange juice minus it:

```{r coefficients-sum}
fit3$assumptions$coefficients[, c("term", "estimate")]
mean(fit$emmeans$estimate)
```

There is no formal normality test. Read the Q-Q plot below. For equal
variances, compare the cell standard deviations:

```{r cell-sds}
cell_sd <- tapply(tg$len, list(tg$supp, tg$dose), sd)
round(cell_sd, 2)
```

They range from `r f2(min(cell_sd))` to `r f2(max(cell_sd))`, so the largest
cell variance is `r f2((max(cell_sd) / min(cell_sd))^2)` times the smallest.
With equal cell sizes of ten, a balanced design tolerates that well. If the
variances differed badly, you could use heteroskedasticity-consistent
standard errors (`vcov_type`, below) or, for a one-way design,
`anova_welch()`. The [diagnostics article](diagnostics.html) covers the checks
across the package.

## Plots

```{r plot-box}
plot(fit, "box")
```

The `box` plot shows the observed response in each cell, ordered by median and
labelled with its size. Use it to see the raw data behind the means: spreads,
skewness and outliers.

```{r plot-emmeans}
plot(fit, "emmeans")
```

The `emmeans` plot shows the estimated marginal means with their 95% intervals.
The first grouping factor is on the x axis and the second sets the colour, one
line per supplement. Lines that are not parallel are the interaction: the gap
between the supplements closes at 2 mg/day.

```{r plot-residuals}
plot(fit, "residuals")
```

The `residuals` plot shows deviance residuals (for a Gaussian model, ordinary
residuals) against fitted values. The points fall in vertical stripes, one per
cell mean. The two 2 mg/day cells have almost the same mean, so their stripes
overlap at about 26. Look for stripes of clearly different heights (unequal
variances) or a funnel that widens with the fitted value (a variance that
grows with the mean, which suggests another family). Here the heights vary
somewhat, as the cell standard deviations did, but there is no funnel.

```{r plot-qq}
plot(fit, "qq")
```

The `qq` plot compares the deviance residuals with a normal distribution.
Points close to the dashed line support the normal-errors assumption, and
systematic curvature at the ends suggests skewness or heavy tails. Here the
points follow the line well. The lower tail is a little shorter than a normal
one, a mild departure for a balanced design with 60 observations.

All four are **ggplot2** objects that you can modify with `+`. See
[Plots and visual customisation](visuals.html).

## A non-Gaussian family: Gamma

### The data

`MASS::Cars93` lists 93 car models on sale in the US in 1993. Their `Price`
(in thousands of dollars) is positive and right-skewed, and within each `Type`
of car the spread grows with the typical price:

```{r cars}
cars <- MASS::Cars93[, c("Manufacturer", "Model", "Type", "Price")]
price <- split(cars$Price, cars$Type)
round(t(sapply(price, function(x) c(n = length(x), median = median(x),
                                     mean = mean(x), sd = sd(x),
                                     cv = sd(x) / mean(x)))), 2)
```

A Gaussian model assumes the same standard deviation in every group. It is
about 2 for small cars and about 12 for midsize ones. A Gamma family assumes
instead that the standard deviation is proportional to the mean: a constant
coefficient of variation (`cv`). The `cv` column shows that fits better,
though not perfectly (vans are unusually uniform). A log link makes the group
effects multiplicative, so comparisons come out as ratios of means.

### Fitting and the omnibus test

```{r gamma-fit}
gam <- anova_glm(cars, "Price", "Type", family = Gamma(link = "log"))
gam
```

`family` accepts a family object, as here, a family function, or a name such
as `"Gamma"`. The name alone gives the Gamma family's default inverse link,
which is rarely what you want for group comparisons (see
[below](#the-link-decides-what-a-comparison-is)).

**`test_statistic`.** By default the test is chosen after fitting. A family
that estimates its dispersion, like the Gaussian and the Gamma, gets an F test,
and a family whose dispersion is fixed at 1 (binomial, Poisson) gets a
likelihood-ratio test. In the F table of a non-Gaussian family the `sum_sq`
column holds deviances, as the `anova_glm()` help page explains. The `Type`
row is the drop in deviance when `Type` is added. The `Residuals` row is the
Pearson chi-squared statistic, and dividing it by its degrees of freedom gives
the dispersion. F is the ratio of the two per degree of freedom:

```{r f-by-hand}
a <- gam$anova
(a$sum_sq[1] / a$df[1]) / (a$sum_sq[2] / a$df[2])
```

You can ask for the likelihood-ratio or Wald statistic instead. The columns
come in the same order whatever the statistic, and only an F table has
`sum_sq` and a `Residuals` row:

```{r test-statistic}
for (ts in c("LR", "Wald")) {
  g <- anova_glm(cars, "Price", "Type", family = Gamma(link = "log"),
                 test_statistic = ts, plots = FALSE)
  cat("\n", attr(g$anova, "statistic"), "\n", sep = "")
  print(g$anova)
}
```

The likelihood-ratio statistic is the deviance drop divided by the estimated
dispersion, referred to a chi-squared distribution. That ignores the
uncertainty in the dispersion estimate, which the F test allows for, so for a
family that estimates its dispersion F is the better default. The Wald test is
built from the coefficients and their covariance matrix. It is the only one of
the three that can use a robust covariance (see
[`vcov_type`](#vcov_type-robust-standard-errors)). Here all three agree that
the car types differ.

### Effect sizes

```{r gamma-effects}
gam$effect_sizes
```

For a non-Gaussian family there are no sums of squares, so the effect size is
the `deviance_explained`: 1 minus the residual deviance over the null
deviance. Car type explains `r round(100 * gam$effect_sizes$estimate[1])`% of
the deviance. McFadden's pseudo R squared is reported only for the binomial,
Poisson and negative binomial families. For a Gamma model it is `NA`, and the
note in the printout explains why.

### The coefficient table and `ci_method`

```{r gamma-coefficients}
co <- gam$assumptions$coefficients
co
```

The coefficients are on the log scale. The intercept is the log of the mean
price of the reference type, `Compact`, and each other coefficient is the log
of a ratio to it: small cars cost exp(`r f3(co$estimate[4])`) =
`r f2(exp(co$estimate[4]))` times as much as compact ones on average. The
`distribution` column says the tests use t, because the dispersion is
estimated.

`ci_method = "profile"` (the default) gives profile-likelihood intervals, cut
at t quantiles so that they agree with the t tests beside them.
`ci_method = "wald"` gives estimate ± t × SE:

```{r ci-method}
wald <- anova_glm(cars, "Price", "Type", family = Gamma(link = "log"),
                  ci_method = "wald", plots = FALSE)$assumptions$coefficients
cbind(co[, c("term", "estimate")],
      profile = co[, c("conf_low", "conf_high")],
      wald = wald[, c("conf_low", "conf_high")])
```

The two differ in the third decimal place here. Profile intervals follow the
shape of the likelihood, which matters more in small samples or near a
boundary. Wald intervals are symmetric on the log scale. Exponentiate either to
get an interval for the ratio.

### Dispersion

```{r gamma-dispersion}
gam$assumptions$dispersion
sqrt(gam$assumptions$dispersion)
```

For a Gamma family the dispersion is the squared coefficient of variation. The
model therefore puts the standard deviation of prices at about
`r round(100 * sqrt(gam$assumptions$dispersion))`% of the mean in every car
type. The observed values above range from 10% (vans) to 45% (midsize cars), so
the constant-CV assumption is only roughly true. The
[`vcov_type`](#vcov_type-robust-standard-errors) section shows how to make the
comparisons robust to that.

### `$model_stats` and `$family`

```{r model-stats}
str(gam$model_stats)
gam$family
```

`$model_stats` holds the AIC, BIC, null and residual deviances and residual
degrees of freedom. `$family` is the family object actually used. Because
the Gaussian and Gamma fits are full likelihoods for the same response, their
AICs can be compared directly:

```{r aic}
gau <- anova_glm(cars, "Price", "Type", plots = FALSE)
c(gaussian = gau$model_stats$AIC, gamma = gam$model_stats$AIC)
```

The Gamma model is much better supported.

### Marginal means on the response scale

```{r gamma-emmeans}
gam$emmeans
```

The marginal means are on the response scale, in thousands of dollars. In a
one-factor model they are the group means. Their intervals are computed on the
log scale and back-transformed, so they are not symmetric about the estimate.

With a log link, a pairwise comparison is a **ratio** of means, in a column
called `ratio`, tested against a null value of 1:

```{r gamma-posthoc}
gph <- gam$posthoc
print(gph[gph$p_adjusted < 0.05,
          c("contrast", "ratio", "conf_low", "conf_high", "p_adjusted")],
      digits = 3)
```

Those are the comparisons that survive Tukey's adjustment. Midsize cars, for
example, average
`r f2(gph$ratio[gph$contrast == "Midsize / Small"])` times the price of small
ones.

### The link decides what a comparison is

With the log link comparisons are ratios, and with the logit link they are odds
ratios. With any other non-identity link they are differences between the
means on the response scale, and `$notes` says so. The Gamma family's default
link is the inverse:

```{r inverse-link}
inv <- anova_glm(cars, "Price", "Type", family = "Gamma", plots = FALSE)
inv$method
inv$notes[grepl("link", inv$notes)]
head(inv$posthoc[, c("contrast", "estimate", "conf_low", "conf_high")], 3)
```

### Plots for the Gamma fit

```{r gamma-residuals}
plot(gam, "residuals")
```

If the variance really is proportional to the mean squared, the deviance
residuals of a Gamma model have roughly the same spread at every fitted
value. Here the stripes differ in height. Two sit side by side at about 19:
vans (19.1) on the left and sporty cars (19.4) on the right. The vans' stripe
is by far the shortest, because their prices vary much less than the model
assumes. The midsize stripe, at about 27, is the tallest.

```{r gamma-qq}
plot(gam, "qq")
```

The deviance residuals of a Gamma model are approximately normal when the
model fits. Here the middle follows the line and the upper tail rises above
it: a few cars are dearer than a Gamma distribution with this dispersion
would suggest. `gam$model` is an ordinary `glm`, so you can find them
directly:

```{r gamma-top}
res <- residuals(gam$model, type = "deviance")
head(cars[order(res, decreasing = TRUE), ], 3)
```

```{r gamma-emm-plot}
plot(gam, "emmeans")
```

The `emmeans` plot shows the means on the response scale, with their
asymmetric intervals. `Type` has no natural order, so the points are not
joined by a line. Lines appear only across a second grouping factor, as in the
ToothGrowth plot, or along an ordered factor on the x axis.
`plot(gam, "box")` shows the raw prices by type, as for the Gaussian fit.

## Options worth knowing

### `vcov_type`: robust standard errors

`vcov_type` takes an HC type (`"HC0"` to `"HC4"`) from the **sandwich**
package. The robust covariance does not rely on the family's variance
function, so it protects the marginal means and comparisons when that function
is wrong, as the vans suggest it is here:

```{r vcov}
rob <- anova_glm(cars, "Price", "Type", family = Gamma(link = "log"),
                 vcov_type = "HC3", plots = FALSE)
cbind(rob$emmeans[, "Type", drop = FALSE],
      se_model = gam$emmeans$se, se_HC3 = rob$emmeans$se)
rob$notes
```

The standard error for vans shrinks and the one for midsize cars grows: each
now reflects that type's own spread. The notes spell out where the robust
covariance is used. It goes into the coefficient table (with Wald intervals,
since profile intervals cannot use it), the marginal means and the comparisons.
The omnibus F test compares deviances and cannot use it. To make the omnibus
test robust too, set `test_statistic = "Wald"`:

```{r vcov-wald}
anova_glm(cars, "Price", "Type", family = Gamma(link = "log"),
          vcov_type = "HC3", test_statistic = "Wald", plots = FALSE)$anova
```

### `weights`

`weights` names a column of prior weights. It is given as a column name, not a
vector, so the weights are subsetted with the data and cannot misalign when a
row is dropped. A natural use with the Gamma family is a response that is
itself an average. The mean of m prices has variance φμ²/m, so its prior
weight is m. Suppose you had only the average price per manufacturer and type,
with the number of models behind each average:

```{r weights-data}
agg <- aggregate(Price ~ Manufacturer + Type, data = cars, FUN = mean)
agg$n_models <- aggregate(Price ~ Manufacturer + Type, data = cars,
                          FUN = length)$Price
nrow(agg)
table(agg$n_models)
```

Weighting each average by its count recovers the means of the full data
exactly. Ignoring the counts does not:

```{r weights-fit}
w_fit <- anova_glm(agg, "Price", "Type", family = Gamma(link = "log"),
                   weights = "n_models", plots = FALSE)
u_fit <- anova_glm(agg, "Price", "Type", family = Gamma(link = "log"),
                   plots = FALSE)
data.frame(Type = gam$emmeans$Type, full_data = gam$emmeans$estimate,
           weighted = w_fit$emmeans$estimate,
           unweighted = u_fit$emmeans$estimate)
```

The standard errors differ slightly from the full-data fit, because the
dispersion is now estimated from `r nrow(agg)` averages rather than
`r nrow(cars)` cars.

A row with a weight of zero contributes nothing to the fit, so it is dropped
before fitting, counted in `$n_removed` and named in `$notes`:

```{r weights-zero}
agg0 <- agg
agg0$n_models[1] <- 0
z_fit <- anova_glm(agg0, "Price", "Type", family = Gamma(link = "log"),
                   weights = "n_models", plots = FALSE)
z_fit$n_removed
z_fit$notes[1]
```

### `interaction`

The default, `interaction = FALSE`, fits the grouping factors additively. The
marginal means and comparisons are then reported for each factor on its own,
averaged over the others, with a `term` column saying which, and each
factor's comparisons form their own multiplicity family:

```{r additive}
add <- anova_glm(tg, "len", c("dose", "supp"), plots = FALSE)
add$emmeans
add$notes
```

`interaction = TRUE` fits the full factorial. With three or more factors a
whole number sets the highest order of interaction: `interaction = 2` keeps
the two-way interactions and drops the rest.

### `adjust` and `conf_level`

`adjust` is applied by **emmeans** and accepts `"tukey"` (the default),
`"sidak"`, `"scheffe"`, `"dunnettx"`, `"bonferroni"`, `"holm"`, `"hochberg"`,
`"hommel"`, `"BH"`, `"BY"`, `"fdr"` or `"none"`. A step-down method such as
Holm cannot be turned into intervals, so **emmeans** uses Bonferroni intervals
beside Holm p-values, and `$notes` says so:

```{r adjust}
holm <- anova_glm(tg, "len", c("dose", "supp"), interaction = TRUE,
                  adjust = "holm", plots = FALSE)
holm$notes
```

`conf_level` sets the level of every interval: the coefficient table, the
marginal means and the comparisons.

## What the notes say

```{r notes}
fit$notes
gam$notes
```

The ToothGrowth fit has no notes. No rows were dropped, the model is far from
saturated, nothing failed to compute, and the Gaussian family has partial eta
squared to report. The Gamma fit has one, explaining why McFadden's pseudo R
squared is `NA`. Other notes you have seen above: the robust covariance and
where it is used, the per-factor reporting of an additive model, Bonferroni
intervals beside Holm p-values, and differences rather than ratios under the
inverse link. Read `$notes` on every fit.

## Reporting the result

```{r report-values, include = FALSE}
tga <- fit$anova
ga <- gam$anova
tph <- fit$posthoc
row_of <- function(x) tph[tph$contrast == x, ]
r05 <- row_of("dose0.5 OJ - dose0.5 VC")
r1 <- row_of("dose1 OJ - dose1 VC")
r2 <- row_of("dose2 OJ - dose2 VC")
ms <- gph[gph$contrast == "Midsize / Small", ]
```

Write-ups built from the fits with inline R, so the numbers cannot drift from
the analysis:

> A two-way analysis of variance (Type II) found that odontoblast length
> depended on the dose of vitamin C,
> F(`r tga$df[1]`, `r tga$df[4]`) = `r f2(tga$statistic[1])`,
> p `r fp(tga$p_value[1])`,
> partial η² = `r f2(fit$effect_sizes$partial_eta_sq[1])`, and on the
> supplement, F(`r tga$df[2]`, `r tga$df[4]`) =
> `r f2(tga$statistic[2])`, p `r fp(tga$p_value[2])`, with an interaction
> between them, F(`r tga$df[3]`, `r tga$df[4]`) = `r f2(tga$statistic[3])`,
> p `r fp(tga$p_value[3])`. Orange juice produced longer odontoblasts than
> ascorbic acid at 0.5 mg/day (difference `r f2(r05$estimate)`,
> Tukey-adjusted p `r fp(r05$p_adjusted)`) and 1 mg/day
> (`r f2(r1$estimate)`, adjusted p `r fp(r1$p_adjusted)`), but not at
> 2 mg/day (`r f2(r2$estimate)`, adjusted p `r fp(r2$p_adjusted)`).

> In a Gamma model with a log link, car prices differed by type,
> F(`r ga$df[1]`, `r ga$df[2]`) = `r f2(ga$statistic[1])`,
> p `r fp(ga$p_value[1])`, with type explaining
> `r round(100 * gam$effect_sizes$estimate[1])`% of the deviance. Midsize cars
> cost `r f2(ms$ratio)` times as much as small cars on average (95% CI
> `r f2(ms$conf_low)` to `r f2(ms$conf_high)`, Tukey-adjusted).

For a model on a non-identity link, say which family and link you used, and
report comparisons in the form they take on it: ratios for a log link.

## See also

- `anova_bin()` ([walkthrough](binary.html)) and `anova_count()`
  ([walkthrough](counts.html)) for binary and count responses.
- `anova_ancova()` ([walkthrough](ancova.html)) to add a numeric covariate.
- `anova_welch()` ([walkthrough](welch.html)) for a one-way comparison of
  means with unequal variances.
- [Working with results](results.html) for the fit object, `$emmeans_object`
  and reporting.
- [Diagnostics across the package](diagnostics.html) and
  [Plots and visual customisation](visuals.html).
- The [Get Started vignette](anovakit.html) for a tour of every function.
