---
title: "Kruskal-Wallis: comparing groups by ranks"
description: >
  A walkthrough of anova_kw(): the Kruskal-Wallis test, epsilon squared,
  Dunn's comparisons with the tie correction, missing and infinite values,
  and the optional diagnostics.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

```{r helpers, include = FALSE}
# Formatting helpers for numbers quoted in the prose.
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
f4 <- function(x) formatC(x, format = "f", digits = 4)
fp <- function(p) if (p < 0.001) "< 0.001" else paste("=", f3(p))
```

`anova_kw()` compares a numeric response across groups using ranks rather
than the values themselves: the Kruskal-Wallis test across all groups, then
Dunn's pairwise comparisons. Because only the order of the values matters, it
suits responses that are skewed, have outliers, are measured on a coarse or
ordinal scale, or include values that are known only to be "more extreme than
everything else". It asks whether values in some groups tend to rank higher than
in others. That is not a test of means, and it is a test of medians only when
the groups' distributions have the same shape. When the means are the question,
use `anova_welch()` ([walkthrough](welch.html)). The
[decision guide](choosing.html) covers the choice in more detail.

## The data

`airquality` holds daily air-quality readings from New York between May and
September 1973. The response here is `Ozone`, the mean ozone concentration in
parts per billion. The month is stored as a number, so give it readable labels
first. A factor keeps its level order, so the months stay in calendar order.

```{r data}
aq <- airquality
aq$Month <- factor(month.abb[aq$Month], levels = month.abb[5:9])
head(aq[, c("Ozone", "Month", "Day")])

oz <- split(aq$Ozone, aq$Month)
round(t(sapply(oz, function(x) c(days = length(x), missing = sum(is.na(x)),
                                  median = median(x, na.rm = TRUE),
                                  mean = mean(x, na.rm = TRUE),
                                  max = max(x, na.rm = TRUE)))), 1)
```

Two features make a rank analysis attractive. The readings are right-skewed:
in four of the five months the mean is above the median, and in every month
the largest reading is more than twice the median. And
`r sum(is.na(aq$Ozone))` of the `r nrow(aq)` days have no reading,
`r sum(is.na(aq$Ozone[aq$Month == "Jun"]))` of them in June, so the groups are
unequal in size.

## Fitting the model

Columns are named as character strings: the response, then one or more grouping
columns.

```{r fit}
kw <- anova_kw(aq, "Ozone", "Month")
kw
```

Reading the printout from the top:

- The first line names the method, and `Call:` records how the fit was made.
- `Observations used: 116` counts the rows analysed. The
  `r kw$n_removed` rows with no ozone reading were dropped, and the printout
  points to `$notes` for the details.
- `Omnibus test` is the Kruskal-Wallis test, described in the next section.
- `Notes` says what was dropped and why.
- `Plots available` lists what is in `$plots`: only the box plot, unless you
  ask for the optional diagnostics.

Only the analysed columns count towards missingness. `Solar.R` also has
missing values, but it was not part of the analysis, so it removed nothing:

```{r removed}
kw$n_removed
nrow(kw$data_used) + kw$n_removed == nrow(aq)
```

`summary(kw)` prints the same header followed by the effect sizes, the group
summaries and Dunn's comparisons. The sections below take them one at a time.

## The omnibus test

```{r anova}
kw$anova
```

- `term` is the grouping column (or a label for the combined cells when there
  are several; see [below](#several-grouping-columns)).
- `statistic` is the Kruskal-Wallis H. All `r nrow(kw$data_used)` values are
  ranked together, and H measures how far each month's mean rank lies from the
  overall mean rank, weighted by the month's size. It includes the correction
  for tied values.
- `df` is the number of groups minus one. H is referred to a chi-squared
  distribution on these degrees of freedom.
- `p_value` is the probability of an H at least this large if all five months
  shared one distribution of ozone.

The test is `stats::kruskal.test()`, and its `htest` object is kept in
`$model`. Here it agrees with calling `kruskal.test()` directly:

```{r kruskal-check}
kruskal.test(Ozone ~ Month, data = aq)$statistic
```

(`anova_kw()` passes `kruskal.test()` the ranks rather than the raw values, so
that ties are counted exactly as they were ranked. The two can differ slightly
only when the data contain values that differ by floating-point error alone;
see [below](#values-that-differ-only-by-rounding-error).)

The mean ranks differ between months far more than chance would produce
(p `r fp(kw$anova$p_value)`). The test does not say which months differ. That
is what Dunn's comparisons are for.

## Effect sizes

```{r effect-sizes}
kw$effect_sizes
```

With H the tie-corrected statistic, n observations and k groups:

- `epsilon_squared` is H / (n - 1). It is exactly the proportion of the
  variance of the ranks that lies between groups: the R² of a one-way analysis
  of variance on the ranks. You can check that:

  ```{r r-squared}
  summary(lm(rank(Ozone) ~ Month, data = kw$data_used))$r.squared
  ```

- `eta_squared_H` is (H - k + 1) / (n - k), a bias-adjusted version that
  subtracts what k groups would produce by chance. A negative value is reported
  as 0, with a note.

Take care with the names, which follow Tomczak and Tomczak (2014) and the
**effectsize** package. Here epsilon squared is the unadjusted measure and
therefore never smaller than eta squared, which is the reverse of what the same
two names mean in a parametric analysis. Whichever you report, name it.

About `r round(100 * kw$effect_sizes$estimate[1])`% of the variation in the
ranks of the ozone readings lies between months. No interval is reported for
either measure.

## Group summaries and Dunn's comparisons

No model is fitted, so `$emmeans` holds plain group summaries:

```{r emmeans}
print(kw$emmeans, digits = 3)
```

The `mean_rank` column is the quantity Dunn's comparisons are built from.
The `median` is usually the more useful summary to report alongside them. The
`mean`, `sd` and the t interval for the mean (`conf_low`, `conf_high`) are
there for reference. They describe the raw values, and play no part in the
test.

`$posthoc` holds Dunn's comparisons, one row per pair of months:

```{r posthoc}
print(kw$posthoc, digits = 3)
```

- `mean_rank_diff` is the mean rank of `group1` minus that of `group2`.
- `z` is that difference divided by its standard error, so it has the same
  sign.
- `p_value` is the two-sided p-value from the normal distribution, unadjusted.
- `p_adjusted` is adjusted by the method in `adjustment`: Benjamini-Hochberg
  (`"BH"`) by default.

Dunn's test uses the ranks from the single joint ranking of all
`r nrow(kw$data_used)` observations. It does not re-rank each pair, so it is not
the same as running a Wilcoxon rank-sum test on each pair: the comparisons
use the same ranking as the omnibus test. `rstatix::dunn_test()` reports the
same comparisons as `group2` minus `group1`, so its statistics have the
opposite sign.

### The tie correction

Ozone was recorded in whole parts per billion, so many days share a value, and
tied values share their average rank. Ties reduce the variance of the ranks, and
Dunn's standard error allows for it. For groups i and j,

$$
\mathrm{SE}_{ij} = \sqrt{\left(\frac{N(N+1)}{12} -
  \frac{\sum_t (t^3 - t)}{12(N-1)}\right)\left(\frac{1}{n_i} +
  \frac{1}{n_j}\right)},
$$

where the sum runs over the sets of tied values and t is the size of each set.
Reproducing the May against July comparison by hand:

```{r dunn-by-hand}
d <- kw$data_used
N <- nrow(d)
r <- rank(d$Ozone)
t_sizes <- table(r)                     # one entry per distinct value
tie_term <- sum(t_sizes^3 - t_sizes) / (12 * (N - 1))
c(tied_values = sum(t_sizes > 1), untied = N * (N + 1) / 12,
  tie_term = tie_term)

rbar <- tapply(r, d$Month, mean)
n <- table(d$Month)
se <- sqrt((N * (N + 1) / 12 - tie_term) * (1 / n[["May"]] + 1 / n[["Jul"]]))
(rbar[["May"]] - rbar[["Jul"]]) / se
```

That matches the `z` for May against July in `$posthoc`. Here
`r sum(t_sizes > 1)` distinct values are shared by two or more days, but the
correction is small against the untied variance, so it barely changes any
comparison. It matters more when a few values account for much of the data,
as with ratings on a five-point scale.

### Reading the comparisons

```{r posthoc-values, include = FALSE}
ph <- kw$posthoc
row_of <- function(a, b) ph[ph$group1 == a & ph$group2 == b, ]
p_may <- max(row_of("May", "Jul")$p_adjusted, row_of("May", "Aug")$p_adjusted)
```

July and August both rank well above May (both adjusted p `r fp(p_may)`)
and above September (adjusted p `r fp(row_of("Jul", "Sep")$p_adjusted)` and
`r f3(row_of("Aug", "Sep")$p_adjusted)`). July and August do not differ from
each other (adjusted p `r fp(row_of("Jul", "Aug")$p_adjusted)`), and neither
do May, June and September. June sits between: its difference from July is
just inside 5% after adjustment
(`r f4(row_of("Jun", "Jul")$p_adjusted)`), and from August just outside
(`r f4(row_of("Jun", "Aug")$p_adjusted)`). With only
`r kw$emmeans$n[kw$emmeans$group == "Jun"]` June readings, the June
comparisons are the least precise in the table.

### Choosing `adjust`

The default, Benjamini-Hochberg, controls the false discovery rate: among the
comparisons you call significant, the expected proportion of false ones. That
suits the usual follow-up question "which of these pairs differ?" When you
need control of the chance of even one false positive (the family-wise error
rate), use `"holm"`, which is uniformly more powerful than `"bonferroni"`. Any
`stats::p.adjust()` method is accepted, spelled out in full. `"tukey"` is not:
Dunn's test is a comparison of ranks, not a linear contrast.

```{r adjust}
adj <- sapply(c("BH", "holm", "bonferroni", "none"), function(a)
  anova_kw(aq, "Ozone", "Month", adjust = a, plots = FALSE)$posthoc$p_adjusted)
cbind(ph[, c("group1", "group2")], signif(adj, 3))
```

Only `p_adjusted` changes. Under Holm, the June against July comparison no
longer passes at 5%
(`r f4(adj[ph$group1 == "Jun" & ph$group2 == "Jul", "holm"])`), while the
comparisons of July and August with May and September all still do.

## Checking assumptions

Kruskal-Wallis assumes no particular distribution, so `$assumptions` is empty
by default:

```{r assumptions}
kw$assumptions
```

The test does rest on two things no diagnostic in the fit checks. The first
is that the observations are independent. Daily readings are not: a
high-ozone day tends to be followed by another.

```{r autocorrelation}
# Each day's reading against the next day's, within the same month
o <- aq$Ozone
same_month <- aq$Month[-1] == aq$Month[-nrow(aq)]
lag1 <- cor(o[-1][same_month], o[-nrow(aq)][same_month],
            use = "complete.obs")
lag1
```

With a day-to-day correlation of `r f2(lag1)`, the `r nrow(kw$data_used)`
readings carry less independent information than their number suggests, so
treat the p-values here as somewhat optimistic. The second is that if you want
to read a difference as a shift in the median, the groups' distributions need
similar shapes. The box plot below is the place to judge that.

With `diagnostics = TRUE` the function adds a per-group Shapiro-Wilk table in
`$assumptions$normality`, a Q-Q plot in `$plots$qq`, and a note:

```{r diagnostics}
kwd <- anova_kw(aq, "Ozone", "Month", diagnostics = TRUE)
kwd$assumptions$normality
kwd$notes
names(kwd$plots)
```

As the note says, these are context only: nothing in the test depends on them.
They can help you explain why you chose a rank test. Here May and September are
clearly non-normal, which is the kind of evidence that argues against comparing
means with `anova_welch()`. The [diagnostics article](diagnostics.html) covers
the checks across the whole package.

## Plots

```{r plot-box}
plot(kw, "box")
```

The `box` plot is always built. It shows the distribution of each group,
ordered by median and labelled with its size. What to look for:

- **Where the groups sit.** July and August sit clearly above the other three
  months, the pattern Dunn's comparisons found (with the June comparisons the
  least certain).
- **Shape.** In most months the upper whisker is longer than the lower one,
  and May, June, September and August each have at least one high outlier.
  That long upper tail is what pulls the means above the medians.
- **Spread.** The spreads differ: the boxes for July and August are much
  taller than those for the other three months. So the groups differ in more
  than location, and the result is best read as "ozone tends to be higher in
  July and August" rather than as a statement about medians alone.

```{r plot-qq}
plot(kwd, "qq")
```

The `qq` plot appears only with `diagnostics = TRUE`. Each value is expressed
as its deviation from its own group's mean, divided by its own group's
standard deviation, and the points are compared with a normal distribution.
Here the upper tail curves well above the dashed line and the lowest points
also sit above it, the pattern of right skew: a long upper tail and a short
lower one. Like the Shapiro-Wilk table, it is context only.

Both are ordinary **ggplot2** objects that you can modify with `+`:

```{r plot-modify}
plot(kw, "box") +
  ggplot2::labs(title = "Daily ozone by month, New York 1973",
                x = NULL, y = "Ozone (ppb)")
```

See [Plots and visual customisation](visuals.html) for more.

## Infinite values

Missing values are dropped and counted in `$n_removed`, as above. Infinite
values are not. In a rank analysis `Inf` is simply larger than everything else,
which is how you might record, say, a participant who never completed a task.
The small constructed example below records the minutes taken to solve a
puzzle, with three kinds of hint. Anyone who gave up is recorded as `Inf`:

```{r inf-data}
puzzle <- data.frame(
  hint = factor(rep(c("none", "text", "video"), each = 8),
                levels = c("none", "text", "video")),
  minutes = c(20.8, Inf, 13.9, 15.5, Inf, 11.5, Inf, 11.6,
              20.2, 8.8, 15.2, Inf, 5.2, 8.1, 8.5, 11.6,
              5.4, 2.1, 2.3, 10.2, 5.3, 2.9, 5.6, 9.8)
)
pz <- anova_kw(puzzle, "minutes", "hint")
pz
pz$emmeans
```

All 24 rows are used. The four people who gave up rank above everyone who
finished, and take a tied rank among themselves. The mean ranks and medians are
exact. A group containing `Inf` has an infinite mean and no standard deviation,
so those columns are reported as `Inf` and `NA`. The note records that the
infinite values were kept and ranked, and warns that the box plot cannot draw
them, so for data like these read the medians and mean ranks rather than the
plot.

A comparison of means has no use for an infinite value. `anova_welch()` drops
those rows, as it does missing ones:

```{r inf-welch}
anova_welch(puzzle, "minutes", "hint", plots = FALSE)$n_removed
```

## Options worth knowing

### `conf_level`

`conf_level` affects only the t intervals for the group means in `$emmeans`.
Nothing in the test or in Dunn's comparisons depends on it:

```{r conf-level}
k90 <- anova_kw(aq, "Ozone", "Month", conf_level = 0.90, plots = FALSE)
k90$emmeans[, c("group", "mean", "conf_low", "conf_high")]
```

### `posthoc`

There are `choose(k, 2)` comparisons, so with many groups you may want only the
omnibus test. The effect sizes come from H, so they are still reported:

```{r posthoc-false}
anova_kw(aq, "Ozone", "Month", posthoc = FALSE, plots = FALSE)$notes
```

### Several grouping columns

Given several grouping columns, `anova_kw()` combines them into one factor of
the level combinations that contain data, and runs one test across those
cells. `warpbreaks` has two wools at three tensions:

```{r cells}
wk <- anova_kw(warpbreaks, "breaks", c("wool", "tension"), plots = FALSE)
wk$anova
wk$notes
```

The test compares the six cells. As the note says, it cannot separate a wool
effect from a tension effect or test their interaction.

### Values that differ only by rounding error

Values are ranked exactly as stored. Two numbers that should be equal but
were computed differently, such as `0.1 + 0.2` and `0.3`, are therefore ranked
as distinct values, not as a tie. The function looks for such near-ties and
says when it finds them:

```{r near-ties}
nt <- data.frame(g = rep(c("a", "b"), each = 3),
                 y = c(0.1 + 0.2, 0.2, 0.25, 0.3, 0.9, 1.1))
nt_fit <- anova_kw(nt, "y", "g", plots = FALSE)
nt_fit$notes
c(anova_kw = nt_fit$anova$statistic,
  kruskal.test = unname(kruskal.test(y ~ g, data = nt)$statistic))
```

`kruskal.test()` on the raw values ranks the two numbers as distinct but counts
them as a tie when it corrects for ties, because it counts ties on the printed
values. That is why its statistic differs slightly from the one `anova_kw()`
reports, which counts ties on the ranks it actually used. If such values are
meant to be equal, round the response before the analysis. They then become a
genuine tie, the note goes away, and the two statistics agree:

```{r near-ties-rounded}
nt$y <- round(nt$y, 10)
nt_round <- anova_kw(nt, "y", "g", plots = FALSE)
nt_round$notes
c(anova_kw = nt_round$anova$statistic,
  kruskal.test = unname(kruskal.test(y ~ g, data = nt)$statistic))
```

## What the notes say

```{r notes}
kw$notes
```

For the ozone fit the only note records the `r kw$n_removed` rows dropped for
missing values, and names the columns that were checked. Other notes you have
seen above: infinite values kept, the contextual status of the diagnostics,
`posthoc = FALSE`, combined cells, and near-ties. The function also writes a
note when `eta_squared_H` is negative (and floored at 0) or undefined. Read
`$notes` on every fit.

## Reporting the result

```{r report-values, include = FALSE}
a <- kw$anova
es <- kw$effect_sizes
med <- setNames(kw$emmeans$median, kw$emmeans$group)
```

A write-up built from the fit with inline R, so the numbers cannot drift from
the analysis:

> Daily ozone concentrations differed between months (Kruskal-Wallis
> H(`r a$df`) = `r f2(a$statistic)`, p `r fp(a$p_value)`,
> ε² = `r f2(es$estimate[1])`; n = `r nrow(kw$data_used)` days with a
> reading). Median ozone was `r med[["Jul"]]` ppb in July and `r med[["Aug"]]`
> ppb in August, against `r med[["May"]]` ppb in May and `r med[["Sep"]]` ppb
> in September. In Dunn's comparisons with Benjamini-Hochberg adjustment, July
> and August each ranked higher than May (both adjusted p `r fp(p_may)`)
> and than September (adjusted p `r fp(row_of("Jul", "Sep")$p_adjusted)` and
> `r f3(row_of("Aug", "Sep")$p_adjusted)` respectively).

Say which effect size you report (epsilon squared as defined here) and which
adjustment you used. Also say that readings were missing on
`r kw$n_removed` days, mostly in June.

## See also

- `anova_welch()` and the [Welch walkthrough](welch.html) for comparing means.
- `anova_glm()` and the [analysis-of-deviance walkthrough](glm.html) for a
  model of a skewed response, with a Gamma family for example.
- [Working with results](results.html) for the structure of the returned
  object and more on reporting.
- [Diagnostics across the package](diagnostics.html) and
  [Plots and visual customisation](visuals.html).
- The [Get Started vignette](anovakit.html) for a tour of every function.
