---
title: "MANOVA: comparing groups on several responses at once"
description: >
  A walkthrough of anova_manova() on the iris measurements: the multivariate
  test, univariate follow-ups, canonical discriminant analysis, Box's M and
  Mardia's tests, and MANCOVA.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

`anova_manova()` compares two or more numeric responses *jointly* across
groups. It reports a multivariate test of each term, follows it up with one
ANOVA per response, and describes how the groups separate with a canonical
discriminant analysis; with covariates it becomes a MANCOVA. Use it when you
have several related measurements on each unit and want one test of "do the
groups differ on these measurements taken together", or want to know which
combination of them separates the groups. For a single response use
`anova_glm()`, `anova_welch()` or, with a covariate, `anova_ancova()`; for the
same response measured repeatedly on each subject, `anova_rm()`. The
[decision guide](choosing.html) has more.

## The data

`datasets::iris` gives four measurements, in centimetres, of 50 flowers from
each of three species of iris.

```{r data}
resp <- c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")
head(iris)
table(iris$Species)
aggregate(. ~ Species, data = iris, FUN = mean)
```

The four measurements are strongly correlated with each other (bigger flowers
are bigger in every direction), so four separate ANOVAs would test largely the
same thing four times. A MANOVA tests them together, taking the correlations
into account.

## Fitting the model

Responses are named as a character vector, then the grouping variable(s).

```{r fit}
fit <- anova_manova(iris, resp, "Species")
fit
```

Reading the printout:

- The method line names the multivariate statistic (Pillai's trace, the default
  `test`) and the type of test (Type II, the default `type`).
- `Observations used: 150`: no rows were dropped.
- The table is the multivariate test, one row per term.
- The notes say that Mardia's tests reject multivariate normality and that Box's
  M rejects equal covariance matrices. Both are discussed
  [below](#checking-assumptions).
- There are two plots per response (residuals and Q-Q), a marginal-means plot
  and a canonical plot.

`summary(fit)` adds the assumption checks, the effect sizes, marginal means and
comparisons for every response, the canonical axes and each response's
univariate ANOVA table.

## The multivariate test

```{r multivariate}
fit$multivariate
attr(fit$multivariate, "statistic")
```

`$multivariate` (the same table as `$anova`) has one row per term:

- `df` is the term's hypothesis degrees of freedom (a three-level factor has 2).
- `statistic` is the multivariate statistic, here Pillai's trace, which lies
  between 0 and the smaller of the number of responses and `df`.
- `approx_f`, `num_df` and `den_df` are its F approximation, and `p_value` the
  p-value from it.

The tests are computed by `car::Manova()` on the multivariate linear model
`lm(cbind(<responses>) ~ <groups>)`, kept in `$model`. They are Type II by
default: each term is adjusted for every other term that does not contain it,
so the result does not depend on the order of `groups`. `type = "III"` fits
under sum-to-zero contrasts and adjusts each term for all others. With one
grouping variable the two agree, and both agree with
`summary(stats::manova())`, which is sequential and so order-dependent once
there are several terms.

Here the species differ overwhelmingly on the four measurements taken
together (Pillai's trace = `r round(fit$multivariate$statistic, 3)`, p far
below any threshold).

### Which statistic: `test`

There are four standard multivariate statistics, and `test` chooses one:
`"Pillai"` (the default), `"Wilks"`, `"Hotelling-Lawley"` or `"Roy"`. They are
all functions of the same eigenvalues and usually agree, as here:

```{r tests}
tests <- c("Pillai", "Wilks", "Hotelling-Lawley", "Roy")
do.call(rbind, lapply(tests, function(t) {
  f <- anova_manova(iris, resp, "Species", test = t, plots = FALSE)
  data.frame(test = t, f$multivariate[, c("statistic", "approx_f", "num_df",
                                          "den_df", "p_value")])
}))
```

They differ in how they behave when the assumptions fail and in which
alternatives they detect best. Pillai's trace is the most robust to unequal
covariance matrices and non-normality, which is why it is the default. Roy's
largest root uses only the first eigenvalue: it is the most powerful when the
groups differ along a single direction, but with more than one hypothesis
degree of freedom its F statistic is only an upper bound, so its p-value is a
lower bound, and a note says so:

```{r roy}
roy <- anova_manova(iris, resp, "Species", test = "Roy", plots = FALSE)
roy$notes
roy$test
```

## Univariate follow-ups

A significant multivariate test says the groups differ *somewhere* in the
four measurements. The follow-ups say where. One ANOVA (of the same type as the
multivariate test) is fitted per response and kept under `$univariate`, each
with its own model, table, effect sizes, marginal means, emmeans grid and
comparisons:

```{r univariate}
names(fit$univariate)
names(fit$univariate$Sepal.Width)
fit$univariate$Sepal.Width$anova
```

The univariate p-values are *not* adjusted across responses, and they are
reported whether or not the multivariate test is significant. When a term's
multivariate test is not significant at the 0.05 level, a note says that its
follow-ups are not protected by it. Treat a small univariate p-value there as a
lead, not a finding.

## Effect sizes

`$effect_sizes` stacks the partial eta squared and partial omega squared of
every term in every response's ANOVA, with a `response` column:

```{r effect-sizes}
fit$effect_sizes
```

```{r es-read, include = FALSE}
es <- setNames(fit$effect_sizes$partial_eta_sq, fit$effect_sizes$response)
s_pillai <- min(length(resp), fit$multivariate$df)
```

Species accounts for `r round(100 * es[["Petal.Length"]])`% of the variation in
petal length but only `r round(100 * es[["Sepal.Width"]])`% of the variation in
sepal width. No multivariate effect size is reported; a common one is Pillai's
trace divided by the smaller of the number of responses and `df`, here
`r round(fit$multivariate$statistic, 3)` / `r s_pillai` =
`r sprintf("%.2f", fit$multivariate$statistic / s_pillai)`.

## Marginal means and comparisons

The marginal means and comparisons of every response are stacked in one table
each, with a `response` column:

```{r emmeans}
print(fit$emmeans, digits = 3)
```

```{r posthoc}
print(fit$posthoc[, c("response", "contrast", "estimate", "conf_low",
                      "conf_high", "p_value", "p_adjusted")], digits = 3)
```

`p_value` is unadjusted and `p_adjusted` is adjusted by the method in the
`adjustment` column (Tukey by default). The `adjust` argument controls the
adjustment *within* each response only: here each response's three comparisons
are one family, and nothing is adjusted across the four responses. If you
report comparisons for all responses, say so, or adjust the `p_value` column
across the whole table yourself with `p.adjust()`. The many identical
`p_adjusted` values of about 3e-15 are not a coincidence: Tukey p-values come
from `ptukey()`, whose upper tail has an absolute error of about 1e-14, so
digits below about 1e-10 are noise, and `summary(fit)` prints such values as
`< 1e-10`.

Without covariates the marginal means are the raw group means, since there is
nothing to adjust for. `$emmeans_object` is `NULL` because there is one emmeans
grid per response; take the one you need from `$univariate`:

```{r emmeans-object}
is.null(fit$emmeans_object)
class(fit$univariate$Petal.Width$emmeans_object)
```

## Canonical discriminant analysis

The multivariate test says the species differ; the canonical discriminant
analysis says *how*. It finds the weighted combinations of the responses (the
canonical axes) that separate the groups best relative to the variation within
groups:

```{r canonical}
fit$canonical
fit$canonical_term
```

- `eigenvalue` is the ratio of between-group to within-group variation along
  the axis.
- `canonical_r` is the correlation between the axis and the group membership.
- `prop_variance` is the axis's share of the term's between-group variation.

There are at most min(number of responses, `df`) axes, here two. The first
carries `r round(100 * fit$canonical$prop_variance[1], 1)`% of the separation:
the three species lie almost along a single line. To see what the axes mean,
read the structure coefficients, the pooled within-group correlations between
each response and each axis:

```{r structure}
fit$assumptions$structure_coefficients
```

`Can1` is mainly petal length and petal width; `Can2` is mainly sepal width,
with petal width. The sign of an axis is arbitrary; the function makes each
axis's largest structure coefficient positive so the orientation is
reproducible.

`$canonical_term` names the term the analysis describes, spelled as in
`$multivariate$term`. With one grouping variable that is the grouping term.
With several, it is their full interaction when the model contains it, and
otherwise the first grouping variable; see [below](#several-grouping-variables).

## Checking assumptions

MANOVA assumes that the residuals are multivariate normal and that every group
has the same covariance matrix. Iris is the textbook example where the second
assumption fails.

```{r assumptions}
names(fit$assumptions)
fit$assumptions$mardia
fit$assumptions$box_m
fit$assumptions$cell_counts
```

- `mardia` holds Mardia's tests of multivariate skewness and kurtosis, on the
  residuals of the multivariate model. Both reject here (skewness p =
  `r signif(fit$assumptions$mardia$p_value[1], 2)`, kurtosis p =
  `r signif(fit$assumptions$mardia$p_value[2], 2)`). They are asymptotic, and
  the kurtosis test rejects too often when the sample is small relative to the
  square of the number of responses. The tests are skipped, with a note,
  unless the residual degrees of freedom exceed the number of responses by at
  least 10.
- `box_m` is Box's M test of equal covariance matrices across the cells of the
  grouping variables. It rejects decisively (statistic
  `r round(fit$assumptions$box_m$statistic, 1)` on
  `r fit$assumptions$box_m$df` df).
- `cell_counts` gives the number of observations in each cell.
- `structure_coefficients` belongs to the canonical analysis above.

The variances of each measurement within each species show what Box's M is
reacting to:

```{r group-variances}
vars <- sapply(split(iris[resp], iris$Species), function(d) diag(var(d)))
round(vars, 3)
```

Setosa's petals vary far less than the other two species': the variance of its
petal length is `r sprintf("%.3f", vars["Petal.Length", "setosa"])` cm^2^,
against `r sprintf("%.3f", vars["Petal.Length", "versicolor"])` and
`r sprintf("%.3f", vars["Petal.Length", "virginica"])`. So what should you do? Box's M is notoriously sensitive to
non-normality, and it will reject for differences that do not matter in large
samples, so treat it as a prompt to look at the group covariances, as above,
rather than as a verdict. What matters for the multivariate test is how the
covariance difference combines with the group sizes. With equal group sizes, as
here, Pillai's trace is robust to unequal covariance matrices; with unequal
sizes and unequal covariances, the test is liberal when the smaller groups have
the larger covariances, and Pillai's trace is still the safest choice. The
Box's M note makes the same distinction: it checks whether the group sizes are
equal and says whether Pillai's trace is the statistic in use (see
[What the notes say](#what-the-notes-say)). With a separation this strong, no
reasonable violation would change the conclusion.

The [diagnostics article](diagnostics.html) covers the checks across the
package. The checks cost time on a large data set; `assumptions = FALSE` skips
Mardia's tests and Box's M, leaving only the cell counts and the structure
coefficients:

```{r no-assumptions}
quick <- anova_manova(iris, resp, "Species", assumptions = FALSE,
                      plots = FALSE)
names(quick$assumptions)
quick$notes
```

## Plots

Each plot is a ggplot2 object in `$plots`, returned by `plot(fit, "name")`
and modifiable with `+` (see [Plots and visual customisation](visuals.html)).

```{r plot-names}
names(fit$plots)
```

### Residuals and Q-Q plots, per response

Each response's univariate model has its own residuals-against-fitted plot,
`residuals_<response>`, and normal Q-Q plot, `qq_<response>`. The two for petal
length:

```{r plot-residuals}
plot(fit, "residuals_Petal.Length")
```

The fitted values of a one-way model are the group means, so the points fall
in three vertical strips, one per species. What to look for: similar spread in
each strip. Here the setosa strip (fitted value near 1.5 cm) is far narrower
than the other two, the unequal variances seen above.

```{r plot-qq}
plot(fit, "qq_Petal.Length")
```

What to look for: points along the dashed line. These residuals follow it
closely in the middle, and both tails are a little longer than a normal
distribution's (points above the line at the top, below it at the bottom). The
`residuals_Sepal.Length`, `qq_Sepal.Length` and so on are read the same way.

### Marginal means

```{r plot-emmeans, fig.height = 5}
plot(fit, "emmeans")
```

One panel per response, each on its own scale, with the marginal means and
their confidence intervals. The panels show at a glance that setosa is the odd
one out, and that sepal width is the only measurement on which it is the
largest.

### Canonical plot

```{r plot-canonical, fig.height = 5}
plot(fit, "canonical")
```

Every flower's scores on the first two canonical axes, with crosses at the
species centroids. The axes are scaled to unit pooled within-group variance,
so the distance between centroids is in within-group standard deviations. What
to look for: how far apart the groups are along each axis. Here the species are
spread along `Can1` with setosa well clear of the other two, while `Can2`
separates them far less, as its eigenvalue says. With only one estimable axis
the plot becomes a box plot of `Can1` by group, subtitled "Only one
discriminant axis is estimable", as in the examples below.

## MANCOVA: adding covariates

`covariates` turns the analysis into a MANCOVA. `MASS::survey` records the
span of the writing hand (`Wr.Hnd`) and the other hand (`NW.Hnd`), in
centimetres, and the height of 237 students. Men have larger hands, but they
are also taller; adjusting for height asks whether the hands differ between
men and women *of the same height*.

```{r mancova}
sv <- MASS::survey
mc <- anova_manova(sv, c("Wr.Hnd", "NW.Hnd"), "Sex", covariates = "Height")
mc$multivariate
mc$n_removed
```

`r mc$n_removed` rows with a missing value in one of the four columns were
dropped (the notes say which columns). The multivariate table now has a row for
the covariate as well as for `Sex`.

### Centring and adjusted means

Covariates are mean-centred before fitting. Centring changes no test here; it
keeps a covariate with a large offset and a small spread from being mistaken
for a constant, and it means the adjusted means are read at the covariate
means, which are kept in `$covariate_means`:

```{r mancova-means}
mc$covariate_means
print(mc$emmeans, digits = 4)
aggregate(cbind(Wr.Hnd, NW.Hnd) ~ Sex, data = mc$data_used, FUN = mean)
```

```{r mancova-read, include = FALSE}
raw <- aggregate(cbind(Wr.Hnd, NW.Hnd) ~ Sex, data = mc$data_used, FUN = mean)
raw_gap <- diff(raw$Wr.Hnd)
adj <- mc$emmeans[mc$emmeans$response == "Wr.Hnd", ]
adj_gap <- diff(adj$estimate)
```

The raw difference in writing-hand span between the sexes, on these same rows,
is `r round(raw_gap, 2)` cm; at the mean height of
`r round(mc$covariate_means[["Height"]], 1)` cm it is `r round(adj_gap, 2)` cm.
About `r round(100 * (1 - adj_gap / raw_gap))`% of the raw difference goes with
height, and a clear difference remains.

### Homogeneity of regression slopes

A MANCOVA assumes each covariate has the same slope in every group. The
function tests this by comparing the model with one in which the covariate
interacts with every grouping term, using the chosen multivariate statistic:

```{r slopes}
mc$slopes_test
```

The slopes are consistent with being equal here (p =
`r round(mc$slopes_test$p_value, 2)`). When the test rejects at the 0.05 level
a note says so: the common-slope adjusted means then describe the groups at
the covariate mean only, and the advice is to model the interaction yourself or
analyse each response with `anova_ancova()`, which chooses between the two
models and reports per-group slopes.

With covariates, the assumption checks and the canonical analysis work in the
space the multivariate test operates in: Mardia's tests use the residuals of
the multivariate model, Box's M compares the covariance matrices of the
responses with the covariate effects removed, and the canonical scores have
the covariate effects removed. A note records this:

```{r mancova-notes}
mc$notes
```

`Sex` has one degree of freedom, so there is one canonical axis, and the
canonical plot shows it as a box plot by group, subtitled "Only one
discriminant axis is estimable". Because the model has two terms, the subtitle
also names the term the axis describes and says that the other terms' effects
(here, height's) were removed:

```{r mancova-canonical}
plot(mc, "canonical")
```

Men score higher on the axis, but the boxes overlap: at the same height, hand
span separates the sexes only partly (canonical r =
`r round(mc$canonical$canonical_r, 2)`).

## Several grouping variables

`MASS::crabs` has five body measurements (in millimetres) of 200 crabs, 50 of
each sex (`sex`) in each of two colour forms (`sp`, blue or orange).

```{r crabs}
crabs <- MASS::crabs
crab_resp <- c("FL", "RW", "CL", "CW", "BD")
cr <- anova_manova(crabs, crab_resp, c("sp", "sex"), interaction = TRUE)
cr$multivariate
cr$assumptions$cell_counts
```

`interaction` is `FALSE` by default in `anova_manova()`, so the grouping
variables enter additively unless you ask; `TRUE` fits the full factorial and a
whole number keeps interactions up to that order. Here the colour-by-sex
interaction is clear: the difference between the sexes in these measurements
is not the same in the two colour forms.

With the full factorial, `$emmeans` and `$posthoc` compare the cells, and the
marginal-means plot puts the first grouping variable on the x axis with one
line per level of the others:

```{r crabs-emmeans, fig.height = 5}
plot(cr, "emmeans")
```

The canonical analysis describes a single term, which `$canonical_term`
names. With the full interaction in the model it is that interaction, and the
scores have the fitted effects of the main effects removed, so the plot shows
the interaction alone:

```{r crabs-canonical}
cr$canonical_term
cr$canonical
plot(cr, "canonical")
```

`sp:sex` has one degree of freedom, so there is one axis, and the subtitle says
so. With the main effects removed, the scores show the interaction contrast itself: blue females and
orange males on one side, orange females and blue males on the other. Without the
interaction, the analysis describes the main effect of the first grouping
variable; list another variable first in `groups` to describe it instead. In
an additive model, `$emmeans` and `$posthoc` are reported for each factor
separately, averaged over the others, with a `term` column:

```{r crabs-additive}
cr_add <- anova_manova(crabs, crab_resp, c("sex", "sp"), plots = FALSE)
cr_add$canonical_term
head(cr_add$emmeans[, c("response", "term", "sex", "sp", "estimate")], 4)
```

## Options worth knowing

- `test` chooses the multivariate statistic ([above](#which-statistic-test)).
- `type = "III"` gives Type III tests, for the multivariate table and the
  univariate follow-ups alike. If the model has aliased coefficients (an empty
  cell, collinear covariates), Type III is not defined, and Type II is used
  with a note.
- `interaction` chooses how several grouping variables combine
  ([above](#several-grouping-variables)).
- `adjust` sets the adjustment of the comparisons within each response
  (default `"tukey"`); `posthoc = FALSE` skips them.
- `conf_level` sets the level of every interval.
- `assumptions = FALSE` skips Mardia's tests and Box's M
  ([above](#checking-assumptions)); the slopes test is always computed when
  there are covariates.
- `plots = FALSE` skips building the plots, which saves time: there are two
  per response.
- A single response is accepted, and falls back to a univariate analysis with a
  note; there is then no multivariate test and no canonical analysis.

## What the notes say

```{r notes}
fit$notes
```

1. Mardia's tests reject multivariate normality, and the note says that the
   table already uses Pillai's trace, the most robust of the four statistics
   to this.
2. Box's M rejects equal covariance matrices. The note reminds you that the
   test is very sensitive to non-normality and asks you to inspect the group
   covariances. Because the group sizes are equal, it adds that the
   multivariate tests are then fairly robust to unequal covariances, Pillai's
   trace most of all, and that Pillai's trace is the statistic used here.

With another `test`, both notes recommend `test = "Pillai"` instead, as the
`roy` fit [above](#which-statistic-test) shows.

Other notes you may see: dropped rows, centred covariates, a rejected slopes
test, the Roy bound, which term the canonical analysis describes, a term whose
multivariate test is not significant (so its follow-ups are unprotected),
aliased coefficients, skipped checks, and anything car or emmeans said while
the models were being fitted.

## Reporting the result

```{r report-values, include = FALSE}
mv <- fit$multivariate
fmt_p <- function(p) {
  if (p < 0.001) "p < 0.001" else paste("p =", format(signif(p, 2)))
}
cn <- fit$canonical
bm <- fit$assumptions$box_m
u <- function(r) fit$univariate[[r]]$anova[1, ]
```

> A one-way MANOVA compared the three iris species on sepal length, sepal
> width, petal length and petal width. Box's M test rejected equality of the
> covariance matrices (&chi;^2^(`r bm$df`) = `r round(bm$statistic, 1)`,
> `r fmt_p(bm$p_value)`), so Pillai's trace, which is robust to this with equal
> group sizes, was used. The species differed on the measurements jointly,
> Pillai's trace = `r round(mv$statistic, 2)`, F(`r mv$num_df`, `r mv$den_df`)
> = `r round(mv$approx_f, 1)`, `r fmt_p(mv$p_value)`. The first canonical axis
> accounted for `r round(100 * cn$prop_variance[1], 1)`% of the separation
> (canonical r = `r round(cn$canonical_r[1], 2)`) and was defined mainly by the
> petal measurements. Univariate follow-up ANOVAs showed species differences on
> every measurement, from F(`r u("Sepal.Width")$df`,
> `r fit$univariate$Sepal.Width$anova$df[2]`) =
> `r round(u("Sepal.Width")$statistic, 1)` for sepal width (partial
> &eta;^2^ = `r sprintf("%.2f", es[["Sepal.Width"]])`) to
> F(`r u("Petal.Length")$df`, `r fit$univariate$Petal.Length$anova$df[2]`) =
> `r round(u("Petal.Length")$statistic, 0)` for petal length (partial
> &eta;^2^ = `r sprintf("%.2f", es[["Petal.Length"]])`), all `r fmt_p(max(sapply(resp, function(r) u(r)$p_value)))`.

## See also

- `?anova_manova` for every argument and the details of each computation.
- [Working with results](results.html) for the `anovakit_fit` object and the
  emmeans grids.
- `anova_ancova()` ([article](ancova.html)) for one response with covariates,
  including the choice between common and separate slopes.
- `anova_glm()` ([article](glm.html)) for one response in any family.
- `anova_rm()` ([article](repeated-measures.html)) when the "responses" are the
  same measurement taken repeatedly.
- [Diagnostics](diagnostics.html) and [Plots and visual
  customisation](visuals.html).

Box, G. E. P. (1949). A general distribution theory for a class of likelihood
criteria. *Biometrika*, 36(3/4), 317-346.

Mardia, K. V. (1970). Measures of multivariate skewness and kurtosis with
applications. *Biometrika*, 57(3), 519-530.
