---
title: "Repeated measures ANOVA: the same subjects under several conditions"
description: >
  A walkthrough of anova_rm(): mixed designs, sphericity corrections,
  generalised eta squared, incomplete subjects and within-subject residual
  checks.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

`anova_rm()` fits a repeated measures ANOVA, in which every subject is measured
in every cell of one or more *within-subject* factors, optionally with
*between-subject* factors that split the subjects into groups. It takes data in
long format (one row per subject per cell), fits the model with
`afex::aov_ez()` and so needs the afex package. If each subject contributes a
single observation, the groups are independent and `anova_welch()` or
`anova_glm()` is the right tool; for a before-and-after design where the
question is about the "after", `anova_ancova()` with the baseline as a
covariate is often better. The [decision guide](choosing.html) has more.

## The data

`afex::obk.long` is the long form of a contrived data set from O'Brien and
Kaiser (1985): 16 subjects in three treatments (`control`, `A`, `B`), of both
genders, each measured at a pre-test, a post-test and a follow-up (`phase`),
five times an hour apart within each session (`hour`). So there are two
between-subject factors and two within-subject factors, 3 x 5 = 15 cells per
subject and 16 x 15 = 240 rows. The phases are stored in alphabetical order;
putting them in time order makes every table easier to read.

```{r data}
obk <- afex::obk.long
obk$phase <- factor(obk$phase, levels = c("pre", "post", "fup"))
head(obk)
nrow(obk)
# subjects in each between-subject cell
with(obk[!duplicated(obk$id), ], table(treatment, gender))
```

```{r data-means}
round(with(obk, tapply(value, list(treatment, phase), mean)), 2)
```

The control group barely changes across the phases, while both treatments rise
after the pre-test. A repeated measures ANOVA tests effects like that against
the variation *within* subjects, which is usually much smaller than the
variation between them.

## Fitting the model

Every column is named as a string. `subject` identifies the subject, `within`
lists the within-subject factors, and `between` the between-subject factors,
each of which must take one value per subject (the function checks).

```{r fit}
fit <- anova_rm(obk, "value", subject = "id", within = c("phase", "hour"),
                between = c("treatment", "gender"))
fit
```

Reading the printout:

- The heading of the table says the sums of squares are Type III, the
  statistic is F, and the degrees of freedom are Greenhouse-Geisser corrected
  (the default `correction = "GG"`).
- `Observations used: 240`: every subject has every cell, so nothing was
  removed. An incomplete subject would be dropped and named in the notes
  ([below](#incomplete-designs)).
- The table has a row for every term of the full factorial model, 15 in all.
- Two notes: which terms failed Mauchly's test, and a warning that the default
  marginal means average over a factor that interacts with a factor they show.
- Four plots are available.

`summary(fit)` adds the assumption checks, effect sizes, marginal means,
comparisons and the sphericity table.

## The omnibus test

```{r anova}
print(fit$anova, digits = 3)
```

- `num_df` and `den_df` are the numerator and denominator degrees of freedom.
  For a term that involves a within-subject factor they have been multiplied by
  the Greenhouse-Geisser epsilon, which is why they are not whole numbers.
  The between-subject terms (`treatment`, `gender` and their interaction) need
  no correction.
- `mse` is the mean square of the error term the effect is tested against:
  between-subject terms against the variation between subjects, within-subject
  terms against their own subject-by-term interaction.
- `statistic` is F, `partial_eta_sq` is the partial eta squared, and `p_value`
  is computed on the corrected degrees of freedom.

```{r anova-read, include = FALSE}
a <- fit$anova
p_of <- function(t) a$p_value[a$term == t]
fmt_p <- function(p) {
  if (p < 0.001) "p < 0.001" else paste("p =", format(signif(p, 2)))
}
```

`phase` (`r fmt_p(p_of("phase"))`) and `hour`
(`r fmt_p(p_of("hour"))`) are clear, and so is the `treatment:phase`
interaction (`r fmt_p(p_of("treatment:phase"))`): the treatments differ
in how the scores change across the phases, as the table of means suggested.
The type and correction are stored as attributes of the table:

```{r anova-attrs}
attr(fit$anova, "ss_type")
attr(fit$anova, "correction")
```

## Sphericity

A within-subject F test assumes *sphericity*: that the differences between
every pair of levels of a within-subject factor have the same variance. When
the assumption fails the F test rejects too often. The usual remedy is to shrink
both degrees of freedom by an estimate of epsilon, which is 1 under sphericity
and as low as 1/(k - 1) for a factor with k levels.

```{r sphericity}
print(fit$sphericity, digits = 3)
```

```{r sph-read, include = FALSE}
sp <- fit$sphericity
sp_h <- sp[sp$term == "hour", ]
sp_p <- sp[sp$term == "phase", ]
```

There is one row per within-subject term with more than one degree of freedom
(a two-level factor has nothing to test; see [below](#a-within-only-design)).
Terms that share the same within-subject factors share the same error matrix,
so `treatment:hour`, `gender:hour` and `hour` have the same Mauchly's test
and the same epsilons.

- `mauchly_w` and `p_value` are Mauchly's test. It rejects for `hour`
  (p = `r signif(sp_h$p_value, 2)`) and its interactions, and not for `phase`
  (p = `r signif(sp_p$p_value, 2)`).
- `gg_epsilon` is the Greenhouse-Geisser estimate
  (`r round(sp_h$gg_epsilon, 3)` for `hour`), and `hf_epsilon` the
  Huynh-Feldt estimate with Lecoutre's correction
  (`r round(sp_h$hf_epsilon, 3)`), capped at 1. `hf_epsilon_raw` is the
  uncapped value; the two differ only when the estimate exceeds 1, and a note
  then says so.
- `p_gg` and `p_hf` are each term's p-value under each correction, so you can
  see both whichever one `$anova` uses.

The `correction` argument chooses what `$anova` reports: `"GG"` (the default),
`"HF"` or `"none"`.

```{r corrections}
hf <- anova_rm(obk, "value", subject = "id", within = c("phase", "hour"),
               between = c("treatment", "gender"), correction = "HF",
               plots = FALSE)
none <- anova_rm(obk, "value", subject = "id", within = c("phase", "hour"),
                 between = c("treatment", "gender"), correction = "none",
                 plots = FALSE)
rows <- fit$anova$term %in% c("phase", "hour")
data.frame(term = fit$anova$term[rows],
           df_GG = fit$anova$num_df[rows], p_GG = fit$anova$p_value[rows],
           df_HF = hf$anova$num_df[rows], p_HF = hf$anova$p_value[rows],
           df_none = none$anova$num_df[rows], p_none = none$anova$p_value[rows])
none$notes[1]
```

Greenhouse-Geisser is the conservative choice and Huynh-Feldt the less
conservative one; with no correction the p-values for `hour` are the smallest
of the three, and the note warns that they may be too small. A common rule is
to use Huynh-Feldt when the Greenhouse-Geisser epsilon is above about 0.75 and
Greenhouse-Geisser otherwise, but deciding in advance, rather than after seeing
the p-values, matters more than which you pick. Mauchly's test has little power
in small samples, so a non-significant result is weak reassurance; correcting
by default costs little.

The same table is in `fit$assumptions$sphericity`. When there are fewer
subjects than a term has contrasts, Mauchly's test is undefined; the epsilons
are then computed directly and a note says so.

## Effect sizes

```{r effect-sizes}
print(fit$effect_sizes, digits = 3)
```

`partial_eta_sq` is the effect's sum of squares over itself plus its own error
sum of squares. In a repeated measures design each within-subject term has its
own, relatively small, error term, so partial eta squared for those terms is
large and not comparable with the same effect measured in a between-subjects
design. `generalised_eta_sq` (Olejnik and Algina, 2003) puts the subject
variation back into the denominator, so it is comparable across designs, and it
is the measure to report when you want to compare with other studies. Neither
has an interval.

Generalised eta squared depends on which factors were *manipulated* and which
were merely *observed*. By default every factor is treated as manipulated.
Gender was not assigned by the experimenter, so it belongs in `observed`
(passed on to afex):

```{r observed}
obs <- anova_rm(obk, "value", subject = "id", within = c("phase", "hour"),
                between = c("treatment", "gender"), observed = "gender",
                plots = FALSE)
data.frame(term = fit$effect_sizes$term,
           ges_default = fit$effect_sizes$generalised_eta_sq,
           ges_observed = obs$effect_sizes$generalised_eta_sq)[1:8, ]
attr(obs$effect_sizes, "observed")
```

Every generalised eta squared shrinks, because the variation gender accounts
for now counts as part of the individual differences in the denominator.
Partial eta squared, the F tests and the p-values do not change.

## Marginal means and comparisons

By default `$emmeans` holds the marginal means of the within-subject cells,
here the 15 combinations of phase and hour, averaged over the between-subject
factors:

```{r emmeans}
head(fit$emmeans)
nrow(fit$posthoc)
attr(fit$emmeans, "emmeans_model")
```

They come from afex's *multivariate* model, so each cell's standard error and
degrees of freedom come from that cell's own variance, not from a pooled error
term (`df` is the number of subjects minus the number of between-subject
cells, `r fit$emmeans$df[1]`). The `emmeans_model` attribute records this.

The second note of the fit warns that averaging over `treatment` is risky here,
because `treatment:phase` is significant: the average change across phases
hides the fact that the control group does not change. `emm_specs` chooses the
factors the means are computed over. Asking for phase by treatment, averaged
over hour and gender:

```{r emm-specs}
tp <- anova_rm(obk, "value", subject = "id", within = c("phase", "hour"),
               between = c("treatment", "gender"),
               emm_specs = c("phase", "treatment"))
print(tp$emmeans, digits = 3)
```

The comparisons now cover all 36 pairs of the nine cells. Those within a
treatment across phases are usually the ones of interest:

```{r emm-specs-posthoc}
ph <- tp$posthoc
within_trt <- c("pre control - fup control", "pre A - fup A", "pre B - fup B")
print(ph[ph$contrast %in% within_trt,
         c("contrast", "estimate", "conf_low", "conf_high", "p_value",
           "p_adjusted")], digits = 3)
tp$notes[2]
```

`p_value` is unadjusted and `p_adjusted` is adjusted by the method in the
`adjustment` column (Tukey by default, across all 36 comparisons, which is
conservative when you only care about three of them). `adjust` chooses another
method for the whole family. If you planned only a few comparisons, take their
unadjusted `p_value`s and adjust them among themselves with `p.adjust()`. The
note confirms which factors were averaged over.

When you need a contrast the wrapper does not compute -- say, the change from
pre to follow-up compared between treatments -- `$emmeans_object` is the
emmeans grid with your factor names and labels, ready for `emmeans::contrast()`.

## Checking assumptions

Two assumptions matter: sphericity, covered above, and normality of the
within-subject errors.

```{r normality}
print(fit$assumptions$normality, digits = 3)
```

The within-subject F tests depend on what is left of each value once its
subject's mean and its cell's mean (within its between-subject group) are
removed. Those residuals are `fit$residuals`, one per row of `$data_used` and
in the same order, and Shapiro-Wilk is run on them separately for each
within-subject cell. Pooling them would mix cells of different spread, which
can fail a normality test even when every cell is normal. Here no cell rejects
at the 5% level; the smallest p-value is
`r signif(min(fit$assumptions$normality$p_value), 2)`, in the
`r fit$assumptions$normality$cell[which.min(fit$assumptions$normality$p_value)]`
cell. With 15 tests, one or two small p-values are expected by chance, so look
for a pattern rather than a single cell. A cell that cannot be tested (fewer
than three subjects, or no variation) gets an `NA` p-value and the reason in
the `note` column.

```{r residuals}
head(cbind(fit$data_used, residual = round(fit$residuals, 3)))
```

The normality of the subject means, which the between-subject tests rely on, is
not tested. The [diagnostics article](diagnostics.html) covers the checks
across the package.

## Plots

Each plot is a ggplot2 object; `plot(fit, "name")` returns it for you to print
or modify (see [Plots and visual customisation](visuals.html)).

### Residuals against fitted values

```{r plot-residuals}
plot(fit, "residuals")
```

The residuals and fitted values are the within-subject ones described above
(a fitted value is the observed value minus its residual). What to look for:
the same even band as in any residual plot, with no funnel. The values in
this data set are whole numbers on a short scale, which is why the points line
up along diagonal stripes.

### Normal Q-Q plot

```{r plot-qq}
plot(fit, "qq")
```

The residuals here are standardised within each within-subject cell, as they
are tested in `$assumptions$normality`, so cells with different spreads are
put on the same footing. What to look for: points along the dashed line, with
no systematic curve at the ends. Here both ends bend in towards the middle: the
residuals have lighter tails than a normal distribution, which is common for
whole-number scores on a short scale and is generally harmless for the F
tests. Heavy tails, bending away from the line, would be the worrying
pattern.

### Residuals in row order

```{r plot-index}
plot(fit, "index")
```

The residuals in the row order of `$data_used`, which is the order the rows
came in (here, 15 consecutive rows per subject). What to look for: no trends,
and no runs of residuals on the same side of zero. A drift within each block
of rows would suggest practice or fatigue effects that the model does not
capture; if your data are sorted by time of testing, this is where a drift
over the course of the study would show. Nothing of the kind is visible here.

### Marginal means

```{r plot-emmeans}
plot(fit, "emmeans")
plot(tp, "emmeans")
```

The first factor in `emm_specs` goes on the x axis and any others are mapped
to colour, with one line per level. In the first plot (the default, phase by
hour) the pre-test is the lowest phase at every hour and the third hour is the
highest in every phase. The second, phase by treatment, shows the control group
staying flat while `A` and `B` rise, which is the `treatment:phase`
interaction.

## Incomplete designs

A repeated measures ANOVA needs every subject in every within-subject cell.
`datasets::ChickWeight` records the weights of 50 chicks on four diets, every
two days from birth to day 20 and on day 21. Some chicks died or were not
weighed on every day:

```{r chicks}
cw <- as.data.frame(datasets::ChickWeight)
table(table(cw$Chick))   # rows per chick: most have all 12
```

`Time` is numeric; like every within-subject column it is turned into a
factor, so each day is a level.

```{r chicks-fit}
chick <- anova_rm(cw, "weight", subject = "Chick", within = "Time",
                  between = "Diet", plots = FALSE)
chick$subjects_dropped
c(n_removed = chick$n_removed,
  missing = chick$n_removed_missing,
  unbalanced = chick$n_removed_unbalanced,
  aggregated = chick$n_removed_aggregated)
chick$notes[1]
```

Five chicks without a full set of weighings were removed before fitting, and
they are listed in `$subjects_dropped` in the order of the subject column's
levels (`Chick` is an ordered factor, which is why they are not in numeric
order). `$n_removed` counts every input row that is not in `$data_used`, and
three components break it down by cause: rows with a missing value, rows
belonging to incomplete subjects, and rows combined by aggregation (next
section). Here all `r chick$n_removed` removed rows belong to incomplete
subjects.

A missing value costs more than its own row. Setting one weight to `NA`:

```{r chicks-na}
cw_na <- cw
cw_na$weight[cw_na$Chick == "1" & cw_na$Time == 10] <- NA
chick_na <- anova_rm(cw_na, "weight", subject = "Chick", within = "Time",
                     between = "Diet", plots = FALSE)
c(n_removed = chick_na$n_removed,
  missing = chick_na$n_removed_missing,
  unbalanced = chick_na$n_removed_unbalanced)
chick_na$subjects_dropped
```

The row with the missing weight is dropped, which leaves chick 1 without a
day-10 weighing, so its other 11 rows go too.
`nrow(data_used) + n_removed` is always the number of rows supplied. If losing
whole subjects is too costly, a mixed model (for example with lme4 or afex's
`mixed()`) can use incomplete subjects; `anova_rm()` cannot.

These growth data also show how badly sphericity can fail: weights spread out
as chicks grow, so the variance of the differences grows with the gap between
days.

```{r chicks-sphericity}
print(chick$sphericity[, c("term", "mauchly_w", "p_value", "gg_epsilon",
                           "hf_epsilon")], digits = 3)
```

A Greenhouse-Geisser epsilon of `r round(chick$sphericity$gg_epsilon[1], 2)`
is close to its minimum of 1/11 for a 12-level factor: the corrected test has
barely more degrees of freedom than a single comparison.

### Internal names

afex builds a model formula from column names and names its wide-format
columns after the within-subject levels, so a column called `subject id` or
levels such as `0, 2, 4` would be mangled. `anova_rm()` therefore gives afex
internal names and labels, and maps every table and plot back to yours.
`$model` is the afex fit itself and keeps the internal names;
`$internal_names` gives the correspondence:

```{r internal-names}
chick$internal_names
head(chick$model$data$long, 3)
```

The day levels 0, 2, 4, ... are `L01`, `L02`, `L03`, ... in `$model`, in
level order. You only need this table if you work with `$model` directly.

## Repeated rows: `fun_aggregate`

Sometimes a subject has several rows in the same cell: several trials per
condition, or, if you leave a factor out of the analysis, the rows it used to
distinguish. Analysing `obk` by phase alone leaves five rows (the hours) per
subject and phase. They are combined into one value before fitting, with the
mean by default:

```{r aggregate}
by_phase <- anova_rm(obk, "value", subject = "id", within = "phase",
                     between = "treatment", plots = FALSE)
by_phase$notes[1]
c(used = nrow(by_phase$data_used), removed = by_phase$n_removed,
  aggregated = by_phase$n_removed_aggregated)
head(by_phase$data_used, 3)
```

`$data_used` holds the aggregated rows (one per subject and phase), and the
192 rows combined away are counted in `$n_removed` and
`$n_removed_aggregated`. Pass `fun_aggregate` to use another function, for
example the median for skewed trial-level data:

```{r aggregate-median}
by_phase_md <- anova_rm(obk, "value", subject = "id", within = "phase",
                        between = "treatment", fun_aggregate = median,
                        plots = FALSE)
by_phase_md$notes[1]
all.equal(by_phase_md$data_used$value, by_phase$data_used$value)
```

In these contrived data the median and the mean of each subject's five hourly
values happen to coincide, so the result is the same; the note records the
function either way. Aggregation is a convenience: if the repeated rows are a
design factor, like `hour`, it is usually better to keep it in `within`.

## A within-only design

`between` is optional. `datasets::sleep` gives the extra hours of sleep of ten
patients under two drugs (`group`), each patient taking both, so `group` is a
within-subject factor and `ID` the subject:

```{r sleep}
sl <- anova_rm(datasets::sleep, "extra", subject = "ID", within = "group")
sl
```

With a two-level factor there is only one difference per subject, so
sphericity holds automatically: `$sphericity` is `NULL` and the note says why.
(The heading of the table still names the default Greenhouse-Geisser
correction, but there is nothing to correct: the degrees of freedom are the
uncorrected `r sl$anova$num_df` and `r sl$anova$den_df`.) The ANOVA is then the paired t-test in another form, with F = t^2^:

```{r sleep-t}
d <- with(datasets::sleep, extra[group == "2"] - extra[group == "1"])
tt <- t.test(d)
c(t_squared = unname(tt$statistic)^2, F = sl$anova$statistic,
  p_t = tt$p.value, p_F = sl$anova$p_value)
```

The same equivalence holds for the normality check. With one two-level
within-subject factor, the residuals are plus and minus half of each subject's
centred difference, so both rows of `$assumptions$normality` test the
normality of the differences, which is exactly the assumption of the paired
t-test:

```{r sleep-normality}
sl$assumptions$normality
shapiro.test(d)$p.value
```

Both rows give p = `r signif(sl$assumptions$normality$p_value[1], 2)`: the
differences are not convincingly normal, because one patient's difference
(`r max(d)` hours) is far larger than the others (without that patient,
Shapiro-Wilk gives p = `r signif(shapiro.test(d[-which.max(d)])$p.value, 2)`).
Ten subjects is too few to lean on the test, but it is a reason to check the
result with a nonparametric paired test.

## Options worth knowing

- `correction` (`"GG"`, `"HF"`, `"none"`), shown [above](#sphericity).
- `emm_specs`, shown [above](#marginal-means-and-comparisons). When it leaves
  out a factor of the model, the means are averaged over it and a note says so.
- `adjust` sets the multiplicity adjustment for `$posthoc`: `"tukey"` (the
  default), `"holm"`, `"bonferroni"`, `"BH"` and others; `posthoc = FALSE`
  skips the comparisons.
- `conf_level` sets the level of every interval.
- A few arguments of `afex::aov_ez()` pass through `...`: `observed` (above),
  `fun_aggregate` (above), `type = 2` for Type II sums of squares, and
  `anova_table = list(p_adjust_method = "holm")` to adjust the p-values of
  `$anova` across its terms. Anything else is an error rather than silently
  ignored. `covariate` is refused: `anova_rm()` does not fit covariates.

```{r type-2}
t2 <- anova_rm(obk, "value", subject = "id", within = c("phase", "hour"),
               between = c("treatment", "gender"), type = 2, plots = FALSE)
attr(t2$anova, "ss_type")
print(t2$anova[1:3, c("term", "statistic", "p_value")], digits = 3)
```

The design here is unbalanced between subjects (two to four subjects per
treatment-by-gender cell), so Type II and Type III tests differ for most terms.
Only the terms that contain both between-subject factors, such as
`treatment:gender`, are the same under both.

- `factorize = TRUE` (the default) converts a response stored as a factor or
  as text to numbers *through its labels*, so a factor with levels 11, 17, 21
  becomes 11, 17, 21 rather than 1, 2, 3; a note records the conversion.

## What the notes say

```{r notes}
fit$notes
```

1. Mauchly's test rejects sphericity for `hour` and the terms involving it, and
   the table uses the Greenhouse-Geisser correction. The corrected p-values are
   the ones to report.
2. The default marginal means average over `treatment` and `gender`, but
   `treatment:phase` is significant, so those averages can hide the
   difference between treatments. The note suggests the full cell grid; a
   smaller `emm_specs` such as `c("phase", "treatment")`, used above, is often
   enough.

Other notes you may see: subjects removed for incomplete cells, rows
aggregated, a response converted from a factor, epsilons computed directly
because the error matrix is singular, a Huynh-Feldt estimate above 1, a design
with no error degrees of freedom, and anything afex or emmeans said while the
model was being fitted.

## Reporting the result

```{r report-values, include = FALSE}
row <- function(t) a[a$term == t, ]
f2 <- function(x) sprintf("%.2f", x)
ge <- setNames(fit$effect_sizes$generalised_eta_sq, fit$effect_sizes$term)
r_ph <- row("phase"); r_tp <- row("treatment:phase"); r_hr <- row("hour")
em <- tp$emmeans
m <- function(t, p) em$estimate[em$treatment == t & em$phase == p]
```

> A 3 (treatment) x 2 (gender) x 3 (phase) x 5 (hour) mixed ANOVA was run on
> the scores, with Type III sums of squares and Greenhouse-Geisser corrected
> degrees of freedom (Mauchly's test rejected sphericity for hour, W =
> `r f2(sp_h$mauchly_w)`, `r fmt_p(sp_h$p_value)`; epsilon =
> `r f2(sp_h$gg_epsilon)`). Scores changed across phases,
> F(`r f2(r_ph$num_df)`, `r f2(r_ph$den_df)`) = `r f2(r_ph$statistic)`,
> `r fmt_p(r_ph$p_value)`, generalised &eta;^2^ = `r f2(ge[["phase"]])`, and
> across hours, F(`r f2(r_hr$num_df)`, `r f2(r_hr$den_df)`) =
> `r f2(r_hr$statistic)`, `r fmt_p(r_hr$p_value)`, generalised &eta;^2^ =
> `r f2(ge[["hour"]])`. The change across phases depended on treatment,
> F(`r f2(r_tp$num_df)`, `r f2(r_tp$den_df)`) = `r f2(r_tp$statistic)`,
> `r fmt_p(r_tp$p_value)`, generalised &eta;^2^ =
> `r f2(ge[["treatment:phase"]])`: averaged over hour and gender, the control
> group scored `r f2(m("control", "pre"))` at pre-test and
> `r f2(m("control", "fup"))` at follow-up, against `r f2(m("A", "pre"))` to
> `r f2(m("A", "fup"))` for treatment A and `r f2(m("B", "pre"))` to
> `r f2(m("B", "fup"))` for treatment B.

## See also

- `?anova_rm` for every argument, including the details of the sphericity
  computation and the residuals.
- [Working with results](results.html) for the `anovakit_fit` object and
  `$emmeans_object`.
- `anova_ancova()` ([article](ancova.html)) for a pre-post design analysed with
  the baseline as a covariate.
- `anova_welch()` ([article](welch.html)) and `anova_kw()`
  ([article](kruskal-wallis.html)) for independent groups.
- [Diagnostics](diagnostics.html) and [Plots and visual
  customisation](visuals.html).

O'Brien, R. G., & Kaiser, M. K. (1985). MANOVA method for analyzing repeated
measures designs: An extensive primer. *Psychological Bulletin*, 97, 316-333.

Olejnik, S., & Algina, J. (2003). Generalized eta and omega squared statistics:
measures of effect size for some common research designs. *Psychological
Methods*, 8(4), 434-447.
