---
title: "Working with results: the anovakit_fit object"
description: >
  What every function returns, how to print, extract and reuse it, how to go
  further with emmeans and the fitted model, and how to report a result.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

Every anovakit function returns an object of class `anovakit_fit`. It is a
plain named list: every table is a data frame, every figure a **ggplot2**
object, and nothing is hidden behind accessor functions. This article shows
what is in it and how to use it. The examples use `chickwts`, the weights of
71 chicks raised on six feed supplements.

```{r data}
table(chickwts$feed)
```

```{r fit}
fit <- anova_glm(chickwts, "weight", "feed")
```

`anova_glm()` with its default Gaussian family is a classical one-way ANOVA,
which is a convenient model-based example. Everything below applies to all
eight functions; where one differs, it says so.

## What is in a fit

```{r names}
class(fit)
names(fit)
```

The first fourteen components are the same for every function:

| Component | What it holds |
|---|---|
| `method` | The analysis that was run, as printed. |
| `call` | The call that made the fit. |
| `model` | The fitted model (`lm`, `glm`, `mlm`, `afex_aov`), or the `htest` for `anova_welch()` and `anova_kw()`. |
| `anova` | The omnibus test table. |
| `effect_sizes` | The method's effect sizes, with intervals where they exist. |
| `emmeans` | Estimated marginal means (group summaries for Welch and Kruskal-Wallis). |
| `emmeans_object` | The **emmeans** grid behind `$emmeans`, for further contrasts. |
| `posthoc` | Pairwise comparisons. |
| `assumptions` | The assumption checks this method needs. |
| `plots` | Named list of ggplot2 figures. |
| `data_used` | The rows and columns the analysis actually used. |
| `n_removed` | How many input rows are not in `data_used`. |
| `conf_level` | The level of every interval in the fit. |
| `notes` | Everything the function decided or could not compute. |

After those come the components a function adds for itself: here `family` and
`model_stats`. `anova_ancova()` adds the slopes test and the simple slopes,
`anova_rm()` the sphericity table and the subjects it dropped, and so on. Each
walkthrough describes its own, and `?anovakit_fit` lists them all.

## Printing

`print()` shows the omnibus table, any notes, and the names of the plots:

```{r print}
fit
```

`summary()` shows the rest: effect sizes, marginal means, comparisons and the
assumption checks. It returns an object, so you can keep it and print it later
with different `digits`.

```{r summary}
s <- summary(fit, digits = 3)
class(s)
s
```

Both use `digits` for display only; the numbers stored in the fit are never
rounded.

## The tables are ordinary data frames

Every table is a data frame with a documented column layout, so base R, dplyr
or data.table all work on it without conversion.

```{r anova}
fit$anova
```

The model-based functions attach the test details to the omnibus table as
attributes, which is how `print()` knows what heading to write:

```{r anova-attr}
attributes(fit$anova)[c("ss_type", "statistic")]
```

The pairwise comparisons have one row per pair. `p_value` is the *unadjusted*
p-value, `p_adjusted` the adjusted one, and `adjustment` names the method, so
nothing about the adjustment has to be remembered from the call:

```{r posthoc}
names(fit$posthoc)
unique(fit$posthoc$adjustment)
```

Selecting and sorting is plain subsetting. Here are the comparisons that
survive the Tukey adjustment, largest difference first:

```{r posthoc-filter}
ph <- fit$posthoc
sig <- ph[ph$p_adjusted < 0.05, c("contrast", "estimate", "conf_low",
                                  "conf_high", "p_adjusted")]
sig[order(-abs(sig$estimate)), ]
```

The intervals in `$posthoc` from the model-based functions are simultaneous.
With a single-step method such as Tukey's they use the same adjustment as
`p_adjusted`, so an interval that excludes zero agrees with an adjusted p-value
below `1 - conf_level`. Step-down methods such as Holm's cannot be turned into
intervals; **emmeans** then uses Bonferroni intervals, and `$notes` says so.

Writing a table out is one line:

```{r write, eval = FALSE}
write.csv(fit$posthoc, "comparisons.csv", row.names = FALSE)
```

## Read the notes

`$notes` is a character vector. It records every decision the function made on
your behalf (which model it chose, which rows it dropped, which correction it
applied), everything it could not compute and why, and anything **car**,
**emmeans**, **stats** or **afex** said while fitting. None of it is printed to
the console as a warning, so this is the one place to look.

This simple fit needed no decisions, so its notes are empty:

```{r notes-empty}
fit$notes
```

Ask for heteroscedasticity-consistent standard errors and there is more to
say: which parts of the output use them, and what they cannot do.

```{r notes}
robust <- anova_glm(chickwts, "weight", "feed", vcov_type = "HC3",
                    plots = FALSE)
writeLines(robust$notes)
```

Treat the notes as a checklist to read before you quote a number. They are
plain text, so you can also search them, for example to flag which fits in a
batch used robust standard errors:

```{r notes-search}
fits <- list(
  welch  = anova_welch(chickwts, "weight", "feed", plots = FALSE),
  glm    = fit,
  robust = robust
)
vapply(fits, function(f) length(f$notes), integer(1))
vapply(fits, function(f) any(grepl("robust", f$notes, ignore.case = TRUE)),
       logical(1))
```

## Which rows were used

Every function drops rows it cannot analyse (missing or infinite values in the
analysed columns, zero weights, and in `anova_rm()` subjects with an incomplete
design) and counts them. `nrow(fit$data_used) + fit$n_removed` is always the
number of rows you supplied:

```{r rows}
gappy <- chickwts
gappy$weight[c(3, 17, 40)] <- NA
g <- anova_glm(gappy, "weight", "feed", plots = FALSE)
c(supplied = nrow(gappy), used = nrow(g$data_used), removed = g$n_removed)
```

`$data_used` holds only the analysed columns, in the form the model saw them:
grouping columns as factors, and covariates mean-centred when
`anova_ancova()` or `anova_manova()` centred them.

```{r data-used}
str(g$data_used)
```

## Going further with emmeans

`$emmeans` and `$posthoc` cover the usual questions: the mean of each group,
and every pairwise difference. For anything else, `$emmeans_object` is the
**emmeans** reference grid the tables were built from.

```{r emm-object}
grid <- fit$emmeans_object
class(grid)
```

For example, comparing every feed with `horsebean` (Dunnett-style comparisons
with a control) instead of all 15 pairs:

```{r emm-dunnett}
ct <- emmeans::contrast(grid, "trt.vs.ctrl", ref = "horsebean")
confint(ct, level = fit$conf_level)
```

Or a custom contrast: the four plant-based feeds against the two animal-based
ones. The coefficients follow the order of the factor levels.

```{r emm-custom}
levels(chickwts$feed)
plant_vs_animal <- list(
  "plant - animal" = c(casein = -1/2, horsebean = 1/4, linseed = 1/4,
                       meatmeal = -1/2, soybean = 1/4, sunflower = 1/4)
)
confint(emmeans::contrast(grid, plant_vs_animal), level = fit$conf_level)
```

Two things to keep in mind:

- The grid carries **emmeans**' default confidence level, not the fit's
  `conf_level`. Pass `level = fit$conf_level` (as above) when you want the two
  to agree.
- When grouping factors enter a model additively, the means are reported per
  factor and `$emmeans_object` is a named list with one grid per factor. In
  `anova_manova()` there is one grid per response, under
  `fit$univariate[[response]]$emmeans_object`. `anova_welch()` and `anova_kw()`
  fit no model **emmeans** can use, so theirs is `NULL`.

## Going further with the model

`$model` is the fitted model itself: an `lm`, `glm`, `negbin`, multivariate
`mlm` or `afex_aov`. Its call refers to the analysed rows through an
environment the fit carries, so the usual model functions work on it directly,
from anywhere:

```{r model}
class(fit$model)
null <- update(fit$model, . ~ 1)
anova(null, fit$model, test = "F")
AIC(null, fit$model)
```

`anova_welch()` and `anova_kw()` fit no model; for them `$model` is the
`htest` returned by `oneway.test()` or `kruskal.test()`.

```{r htest}
fits$welch$model
```

## Plots

`$plots` is a named list of ggplot2 objects, built but never drawn.
`plot(fit, "name")` returns one, and `plot(fit)` returns the first:

```{r plots}
names(fit$plots)
plot(fit, "emmeans")
```

They are ordinary ggplot2 objects, so `+` adds to them. The
[plots guide](visuals.html) shows every figure the package makes and how to
customise it. When you are fitting many models and do not need the figures,
`plots = FALSE` skips building them and keeps the objects small.

## Many analyses at once

Because every function has the same signature and returns the same structure,
running one analysis over several outcomes or subsets is a loop. Here the
built-in `ToothGrowth` data compare the two vitamin C supplements separately at
each dose:

```{r loop}
tg <- ToothGrowth
by_dose <- lapply(split(tg, tg$dose), function(d) {
  anova_welch(d, "len", "supp", plots = FALSE)
})
data.frame(
  dose      = names(by_dose),
  statistic = vapply(by_dose, function(f) f$anova$statistic, numeric(1)),
  df        = vapply(by_dose, function(f) f$anova$den_df, numeric(1)),
  p_value   = vapply(by_dose, function(f) f$anova$p_value, numeric(1)),
  row.names = NULL
)
```

If you then need to adjust across those tests, `p.adjust()` on the
`p_value` column does it.

## Reporting a result

Everything a write-up needs is in the tables. A small helper keeps the
formatting in one place:

```{r report-helper}
fmt_p <- function(p) {
  if (p < 0.001) "p < .001" else sprintf("p = %s", sub("^0", "", sprintf("%.3f", p)))
}
a <- fit$anova[fit$anova$term == "feed", ]
res_df <- fit$anova$df[fit$anova$term == "Residuals"]
es <- fit$effect_sizes[fit$effect_sizes$term == "feed", ]
```

```{r report-es}
es
```

> Chick weight differed between feed supplements,
> F(`r a$df`, `r res_df`) = `r sprintf("%.2f", a$statistic)`,
> `r fmt_p(a$p_value)`, partial η² = `r sprintf("%.2f", es$partial_eta_sq)`,
> partial ω² = `r sprintf("%.2f", es$partial_omega_sq)`. Of the
> `r nrow(fit$posthoc)` pairwise comparisons,
> `r sum(fit$posthoc$p_adjusted < 0.05)` were significant after Tukey's
> adjustment.

Because the numbers come from the fit rather than being typed, they cannot drift
from the analysis when the data change.

## Saving and reproducing

A fit is an ordinary R object, so `saveRDS()` stores it whole, figures
included. `$call` records how it was made:

```{r call}
fit$call
```

Fits made with `plots = FALSE` are much smaller, which matters when you store
many of them.
