---
title: "Plots: every figure anovakit returns, and how to customise them"
description: >
  A tour of every ggplot2 figure the eight functions build, grouped by kind,
  with what to look for in each; then how to change them, build your own from
  the result tables, combine and save them.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

```{r helpers, include = FALSE}
f2 <- function(x) formatC(x, format = "f", digits = 2)
```

Every anovakit function builds its figures with **ggplot2** and returns them
in `$plots`. Nothing is drawn as a side effect: a plot appears only when you
print it, and until then it is an ordinary ggplot object that you can change
with `+`, save, or throw away. This article shows every plot each function
returns, grouped by what it shows, and then how to customise them. For the
statistics behind the diagnostic plots, see the
[diagnostics article](diagnostics.html).

## The plots each function returns

One fit per function, on data used throughout this article:

```{r fits}
bw <- transform(MASS::birthwt,
                race = factor(race, labels = c("white", "black", "other")),
                smoke = factor(smoke, labels = c("no", "yes")))
tg <- transform(ToothGrowth, dose = factor(dose))

fits <- list(
  welch  = anova_welch(chickwts, "weight", "feed"),
  kw     = anova_kw(InsectSprays, "count", "spray", diagnostics = TRUE),
  ancova = anova_ancova(MASS::anorexia, "Postwt", "Treat", "Prewt"),
  rm     = anova_rm(CO2, "uptake", subject = "Plant", within = "conc",
                    between = c("Type", "Treatment"),
                    emm_specs = c("conc", "Type")),
  manova = anova_manova(iris, names(iris)[1:4], "Species"),
  bin    = anova_bin(bw, "low", c("race", "smoke")),
  count  = anova_count(warpbreaks, "breaks", c("tension", "wool"),
                       interaction = TRUE),
  glm    = anova_glm(tg, "len", c("dose", "supp"), interaction = TRUE)
)
lapply(fits, function(f) names(f$plots))
```

`anova_kw()` returns its Q-Q plot only with `diagnostics = TRUE`.
`anova_manova()` returns a residual plot and a Q-Q plot for each response,
named after it. `anova_ancova()` names its covariate plot `covariate` when
there is one covariate and `covariate_<name>` for each of several.
`print(fit)` lists the plots a fit holds, and `plot(fit, "name")` returns one.

## Group summaries

These plots describe the data as observed, before any model.

### Group means: `anova_welch()`

```{r welch-means}
plot(fits$welch, "means")
```

Each bar is a group mean and each error bar its t interval at `conf_level`,
computed from that group alone, so a small or variable group gets a wide
interval. The axis labels carry each group's size. These intervals describe
the means one at a time; whether two groups differ is a question for
`$posthoc`, whose intervals are for the differences.

### Box plots: `anova_welch()`, `anova_kw()`, `anova_glm()` and `anova_count()`

Four functions draw the response by group as box plots, with the groups
ordered by their medians and each label showing the number of observations in
that cell. `anova_welch()` and `anova_kw()` call it `box`:

```{r boxes-1, fig.show = "hold", out.width = "50%", fig.width = 5, fig.height = 4}
plot(fits$welch, "box")
plot(fits$kw, "box")
```

`anova_glm()` calls it `box` too (for a numeric response), and
`anova_count()` calls it `observed`. With several grouping variables the
boxes are the cells of their combinations:

```{r boxes-2, fig.show = "hold", out.width = "50%", fig.width = 5, fig.height = 4}
plot(fits$glm, "box")
plot(fits$count, "observed")
```

What to look for: boxes of very different heights (unequal spread, which
`anova_welch()` and robust standard errors allow for and a pooled-variance
model does not), a median off-centre in its box or a long whisker on one side
(skew), and points beyond the whiskers. In the insect counts on the right of
the first pair, the sprays with low counts also have the smallest spread, as
counts usually do: a hint that a count model suits them better than a
comparison of means.

### Observed proportions: `anova_bin()`

```{r bin-proportions}
plot(fits$bin, "proportions")
```

Each bar is one cell of the grouping variables, split into the proportions of
the two outcomes and labelled with the percentages; the subtitle names the
level being modelled. The cells are the combinations of every grouping
variable, even when the model is additive, so this is the place to see an
interaction the model does not include.

## Marginal means with intervals

Every model-based function plots its estimated marginal means (`$emmeans`)
with error bars at `conf_level`; the subtitle says which level. The first
grouping variable goes on the x axis and any others are mapped to colour, with
one line per level. As with the group means, the error bars are intervals for
each mean, not for the differences between them: two bars can overlap while
`$posthoc` finds a clear difference.

`anova_ancova()` plots means adjusted to the covariate mean, which is why the
title says "covariate-adjusted":

```{r ancova-emmeans}
plot(fits$ancova, "emmeans")
```

`anova_glm()` with an interaction plots one line per level of the second
factor, so lines that are not parallel are the interaction. In the tooth
growth data the advantage of orange juice over vitamin C at the two lower
doses has gone at the highest:

```{r glm-emmeans}
plot(fits$glm, "emmeans")
```

`anova_count()` plots the estimated mean count in each cell, on the count
scale (a rate per unit of exposure when there is an `offset`). Here wool A
breaks far more often than wool B at low tension and no more often at the
other two:

```{r count-emmeans}
plot(fits$count, "emmeans")
```

`anova_rm()` plots the factors named in `emm_specs` (by default the
within-subject factors). Here the plant's origin was added, so the lines
compare the two types across concentrations, and the gap between them widens
as the concentration rises:

```{r rm-emmeans}
plot(fits$rm, "emmeans")
```

`anova_manova()` plots one panel per response, each with its own y axis:

```{r manova-emmeans, fig.height = 5}
plot(fits$manova, "emmeans")
```

When the grouping variables enter a model additively (the default for
`anova_bin()`, `anova_count()`, `anova_glm()` and `anova_manova()` with more
than one), the marginal means are reported for each factor separately,
averaged over the others, and the plot has one panel per factor:

```{r bin-emmeans}
plot(fits$bin, "emmeans")
```

## Residual diagnostics

### Q-Q plots

A Q-Q plot puts the sorted residuals against the quantiles a normal
distribution would give them. Points on the dashed line are consistent with
normality; an S shape (both tails beyond the line) means heavy tails, a bow
means skew, and horizontal steps mean tied or discrete values. What goes into
the plot differs by function. `anova_welch()` standardises each value within
its own group, as Welch's test allows each group its own variance;
`anova_kw()` does the same, as context only. The chick weights follow the
line closely:

```{r welch-qq, fig.height = 3.5}
plot(fits$welch, "qq")
```

The insect counts rise above the line in the upper tail, the mark of right
skew, and show short steps where counts are tied:

```{r kw-qq, fig.height = 3.5}
plot(fits$kw, "qq")
```

### Residuals against fitted values

The model-based functions pair a Q-Q plot with a plot of residuals against
fitted values. In a model with only grouping variables the fitted values take
one value per cell, so the points line up in vertical strips; compare the
strips' spread. `anova_glm()` plots deviance residuals, which for a Gaussian
model are the ordinary residuals:

```{r glm-resid, fig.show = "hold", out.width = "50%", fig.width = 5, fig.height = 4}
plot(fits$glm, "residuals")
plot(fits$glm, "qq")
```

What to look for: a funnel, with the spread growing with the fitted value,
means the variance depends on the mean (use `vcov_type = "HC3"`, or a family
or transformation that expects it); a curve means the model misses something
systematic. The tooth-growth strips have similar spreads. With a covariate in
the model the fitted values spread out continuously, as in `anova_ancova()`:

```{r ancova-resid, fig.show = "hold", out.width = "50%", fig.width = 5, fig.height = 4}
plot(fits$ancova, "residuals")
plot(fits$ancova, "qq")
```

`anova_manova()` gives both plots for each response, from that response's
univariate model:

```{r manova-resid, fig.show = "hold", out.width = "50%", fig.width = 5, fig.height = 4}
plot(fits$manova, "residuals_Petal.Width")
plot(fits$manova, "qq_Petal.Width")
```

Petal width shows both patterns worth knowing. The strips widen from left to
right, a funnel: the species with wider petals vary more (residual standard
deviations of
`r paste(f2(tapply(residuals(fits$manova$univariate$Petal.Width$model), iris$Species, sd)), collapse = ", ")`
for setosa, versicolor and virginica),
which is part of what Box's M detects. And the Q-Q plot moves in steps,
because petal width is recorded to the nearest 0.1 cm, so many residuals are
tied.

### Repeated measures: residuals, Q-Q and index

`anova_rm()` plots the within-subject residuals, what is left of each value
after its subject's mean and its cell's mean are removed, against the fitted
values:

```{r rm-resid, fig.height = 3.5}
plot(fits$rm, "residuals")
```

Its Q-Q plot standardises them within each within-subject cell, as the
normality tests in `$assumptions$normality` do:

```{r rm-qq, fig.height = 3.5}
plot(fits$rm, "qq")
```

The third plot shows the same residuals in the row order of `$data_used`:

```{r rm-index, fig.height = 3.5}
plot(fits$rm, "index")
```

It is only as informative as that order. `CO2` is sorted by plant and, within
plant, by concentration, so a run of residuals of one sign would point to a
plant whose response curve has a different shape from the others'. In data
stored in the order they were collected, a trend here would point to drift
over time.

## Special plots

### The covariate plot: `anova_ancova()`

```{r ancova-covariate}
plot(fits$ancova, "covariate")
```

The response against the covariate, one colour and one fitted line per group,
with a confidence band at `conf_level` for each line. It shows the two things
an ANCOVA rests on. The slopes: an ANCOVA with common slopes assumes parallel
lines, and here the control group's line is flat while both treatment lines
rise, which is why the homogeneity-of-slopes test rejected and the fit kept
the interaction (see `$slopes_test` and `$simple_slopes`). And the overlap:
the adjusted means are read at the covariate mean, and a group whose points
do not reach that value is being extrapolated along its line. The covariate
is shown on its original scale even though the model is fitted on the centred
covariate.

### The canonical plot: `anova_manova()`

```{r manova-canonical, fig.height = 5}
plot(fits$manova, "canonical")
```

The canonical discriminant analysis finds the combinations of the responses
that separate the groups best. Each point is one observation's score on the
first two axes, and the crosses are the group centroids. The axes are scaled
to unit pooled within-group variance, so a distance of 1 is one within-group
standard deviation: centroids several units apart are well separated, as the
three species are along `Can1`. `$canonical` gives each axis's share of the
between-group variation, and `$assumptions$structure_coefficients` says which
responses each axis is made of:

```{r canonical-tables}
fits$manova$canonical
fits$manova$assumptions$structure_coefficients
```

`Can1` carries `r round(100 * fits$manova$canonical$prop_variance[1], 1)`%
of the separation and is dominated by the petal measurements. With only one
estimable axis (two groups, or a single-degree-of-freedom term) the plot is a
box plot of `Can1` by group instead.

## Customising

### Getting a plot

`plot(fit, "name")` returns the stored object, the same as
`fit$plots$name`; `plot(fit, 2)` takes the second by position, and
`plot(fit)` the first. Asking for a name that does not exist lists the ones
that do:

```{r get-plot}
p <- plot(fits$welch, "means")
identical(p, fits$welch$plots$means)
try(plot(fits$welch, "residuals"))
```

### Skipping them: `plots = FALSE`

Building the plots takes time, and each plot carries its own copy of the data
it draws. When you only want the numbers, in a simulation or a loop over many
responses, turn them off:

```{r no-plots}
quick <- anova_manova(iris, names(iris)[1:4], "Species", plots = FALSE)
length(quick$plots)
identical(quick$multivariate, fits$manova$multivariate)
c(with_plots = length(serialize(fits$manova, NULL)),
  without = length(serialize(quick, NULL)))
plot(quick)
```

The tables are unchanged, the time spent building the plots is saved, and
without its ten plots the MANOVA fit is
`r round(length(serialize(fits$manova, NULL)) / length(serialize(quick, NULL)))`
times smaller when saved. `plot()` on such a fit says why it has nothing to
return.

### Changing a plot with `+`

The rest of this article attaches ggplot2:

```{r ggplot2}
library(ggplot2)
```

Anything you can add to a ggplot you can add to these: labels, scales,
themes, coordinates, facets, extra layers. The package's plots use
`theme_minimal()` with a few adjustments: a bold title, the legend at the
bottom, and x labels rotated by 45 degrees when there are more than six groups
or any label is longer than eight characters. Adding a *complete* theme such
as `theme_bw()` replaces all of that; adding `theme()` changes only the
elements you name.

A horizontal layout often suits long group names. `coord_flip()` moves the
group labels to the vertical axis but leaves the rotation on the horizontal
one, which is now the numeric axis, so undo it there. Note too that `labs()`
names aesthetics, not positions: after the flip, `x` is still the groups.

```{r modify, fig.height = 4.5}
plot(fits$welch, "box") +
  coord_flip() +
  scale_y_continuous(breaks = seq(100, 450, by = 50)) +
  labs(title = "Chick weight by feed supplement", subtitle = NULL,
       x = NULL, y = "Weight at six weeks (g)") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        plot.title = element_text(size = 15))
```

### Colour-blind-safe palettes

The default ggplot2 palette separates groups by hue at equal lightness, which
readers with red-green colour blindness cannot always tell apart. Two
alternatives need nothing beyond R and ggplot2: the Okabe-Ito palette, which
ships with R as `palette.colors()`, and the viridis scales built into
ggplot2. Mapping a second channel, such as point shape, helps too, and
survives greyscale printing:

```{r palette}
okabe_ito <- unname(palette.colors(palette = "Okabe-Ito"))[-1]  # drop black
plot(fits$glm, "emmeans") +
  aes(shape = .series) +
  scale_colour_manual(values = okabe_ito) +
  labs(colour = "Delivery", shape = "Delivery", x = "Dose (mg/day)",
       y = "Tooth length")
```

Giving the colour and the shape the same legend title merges their legends.
For the filled plots (bars, boxes and the proportions plot) the equivalents
are `scale_fill_manual(values = okabe_ito)` and `scale_fill_viridis_d()`.

To map something, you need the plot's own column names. They are in
`names(p$data)`; the line grouping of an emmeans plot with several factors is
`.series`, which is what the shape was mapped to above.

### Facets, and the per-factor plot of an additive model

A plot can be split by any column of its data. The covariate plot's group
column is called `.cell`:

```{r facet-covariate, fig.height = 3.5}
p <- plot(fits$ancova, "covariate")
names(p$data)
p + facet_wrap(~ .cell) + theme(legend.position = "none")
```

With one panel per group the three slopes, and the range of the covariate
each group was observed over, are easier to compare.

The marginal means of an additive model are already faceted, one panel per
factor, by a column called `.term`. Adding a new `facet_wrap()` replaces that
faceting, which is how to change its layout or its strip labels:

```{r facet-additive}
plot(fits$bin, "emmeans") +
  facet_wrap(~ .term, scales = "free_x",
             labeller = as_labeller(c(race = "Mother's race",
                                      smoke = "Smoked during pregnancy"))) +
  scale_y_continuous(labels = function(x) paste0(100 * x, "%")) +
  labs(y = "Probability of low birth weight")
```

### Building your own figures from `$emmeans` and `$posthoc`

The tables behind the plots are ordinary data frames, so a figure the package
does not draw is a few lines away. A forest plot of every pairwise difference
makes the comparisons, rather than the means, the thing you see. For the
chick weights, from a Gaussian model, whose `$posthoc` intervals are
Tukey-adjusted:

```{r forest, fig.height = 5}
cw <- anova_glm(chickwts, "weight", "feed", plots = FALSE)
ph <- cw$posthoc
ph$contrast <- factor(ph$contrast, levels = rev(ph$contrast))
ggplot(ph, aes(x = estimate, y = contrast)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(aes(xmin = conf_low, xmax = conf_high)) +
  labs(title = "Pairwise differences in chick weight",
       subtitle = sprintf("%s%% simultaneous intervals (adjust = \"%s\")",
                          100 * cw$conf_level, ph$adjustment[1]),
       x = "Difference in mean weight (g)", y = NULL) +
  theme_minimal()
```

An interval that crosses the dashed line is a difference the data cannot
distinguish from zero at the adjusted level. The intervals in `$posthoc` from
the model-based functions are adjusted by `adjust`; those from
`anova_welch()` are per-comparison intervals, so label them accordingly.
`$emmeans` works the same way: its `estimate`, `conf_low` and `conf_high`
columns are what the marginal-means plots draw.

Ratios belong on a log scale, where an odds ratio of 2 and one of 0.5 are the
same distance from 1. `$effect_sizes` of `anova_bin()` holds each level's odds
ratio against the reference level, with profile-likelihood intervals:

```{r odds-ratios, fig.height = 3}
or <- fits$bin$effect_sizes
ggplot(or, aes(x = odds_ratio, y = comparison)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_pointrange(aes(xmin = conf_low, xmax = conf_high)) +
  scale_x_log10(breaks = c(1, 2, 4, 8)) +
  facet_grid(factor ~ ., scales = "free_y", space = "free_y") +
  labs(title = "Odds of low birth weight",
       x = "Odds ratio (log scale)", y = NULL) +
  theme_minimal()
```

The same pattern works for incidence rate ratios (`IRR` in the
`$effect_sizes` of `anova_count()`) and for the `ratio` column of a
`$posthoc` on the log or logit scale. If you summarise `$emmeans_object`
yourself with **emmeans**, pass `level = fit$conf_level`: the grid carries
emmeans' own default level, not the fit's.

### Combining plots

Because each plot keeps its data, several plots of one kind can be combined
into one faceted figure without any extra package. The four Q-Q plots of the
MANOVA:

```{r combine, fig.height = 5}
responses <- names(iris)[1:4]
qq_data <- do.call(rbind, lapply(responses, function(r) {
  data.frame(response = r,
             value = fits$manova$plots[[paste0("qq_", r)]]$data$value)
}))
ggplot(qq_data, aes(sample = value)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line(linetype = "dashed") +
  facet_wrap(~ response, scales = "free_y") +
  labs(title = "Normal Q-Q plots of residuals, by response",
       x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal()
```

Placing *different* plots side by side is a job for a layout package.
**patchwork** is the usual choice; it is not an anovakit dependency, so the
chunk below is not run here:

```{r patchwork, eval = FALSE}
library(patchwork)
plot(fits$ancova, "residuals") + plot(fits$ancova, "qq")
```

In a report built with knitr, a chunk with `fig.show = "hold"` and
`out.width = "50%"` puts consecutive plots side by side without any package,
which is how the pairs in this article were drawn.

### Saving

`ggsave()` saves a ggplot to a file, with the format taken from the
extension. Give the size in inches and, for raster formats, the resolution:

```{r save}
out <- file.path(tempdir(), "anorexia-covariate.png")
ggsave(out, plot(fits$ancova, "covariate"), width = 7, height = 4.5,
       dpi = 300)
file.exists(out)
```

A loop saves every plot of a fit, named after it:

```{r save-all}
dir <- file.path(tempdir(), "manova-plots")
dir.create(dir, showWarnings = FALSE)
for (nm in names(fits$manova$plots)) {
  ggsave(file.path(dir, paste0(nm, ".pdf")), fits$manova$plots[[nm]],
         width = 6, height = 4)
}
list.files(dir)
```

PDF and SVG keep lines and text sharp at any size, which suits print; PNG at
300 dpi suits slides and documents.

### `conf_level` sets every error bar

Every interval a function returns, and so every error bar and band in its
plots, uses `conf_level`. The titles and subtitles follow it:

```{r conf-level}
w99 <- anova_welch(chickwts, "weight", "feed", conf_level = 0.99)
plot(w99, "means")$labels$title
a99 <- anova_ancova(MASS::anorexia, "Postwt", "Treat", "Prewt",
                    conf_level = 0.99)
plot(a99, "emmeans")$labels$subtitle
plot(a99, "covariate")$labels$subtitle
```

The bars and bands are then wider by exactly the change in the critical value:
for the casein group, the 99% interval for the mean is
`r f2(diff(unlist(w99$emmeans[1, c("conf_low", "conf_high")])) / diff(unlist(fits$welch$emmeans[1, c("conf_low", "conf_high")])))`
times as wide as the 95% one. To show a different level, refit with
`conf_level` rather than editing the plot: the intervals in the tables and the
figures then agree.

## See also

- [Diagnostics](diagnostics.html) for what the residual plots and the tests
  beside them check, and what to do when they fail.
- [Working with results](results.html) for `$emmeans`, `$posthoc` and the
  rest of the returned object.
- The per-function articles, each with its own plots in context:
  [Welch](welch.html), [Kruskal-Wallis](kruskal-wallis.html),
  [ANCOVA](ancova.html), [GLM](glm.html),
  [repeated measures](repeated-measures.html), [MANOVA](manova.html),
  [binary](binary.html), [counts](counts.html).
