---
title: "Welch's ANOVA: comparing means when variances differ"
description: >
  A walkthrough of anova_welch(): the omnibus Welch test, pairwise Welch t
  tests, standardised mean differences that do not assume equal variances,
  and the three diagnostic plots.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 96, out.width = "100%")
```

```{r libs}
library(anovakit)
```

```{r helpers, include = FALSE}
# Formatting helpers for numbers quoted in the prose.
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fp <- function(p) if (p < 0.001) "< 0.001" else paste("=", f3(p))
pct <- function(x) paste0(formatC(100 * x, format = "f", digits = 1), "%")
```

`anova_welch()` compares the mean of a numeric response across two or more
groups without assuming that the groups share a variance. It is the package's
default choice for a one-way comparison of means: when the variances are equal
it gives up almost nothing to the classical F test, and when they are not it
keeps its stated error rate where the classical test does not. If the response
is heavily skewed, ordinal, or has values that are only "more extreme" rather
than measured, compare ranks with `anova_kw()`
([walkthrough](kruskal-wallis.html)). To adjust for a covariate use
`anova_ancova()`, and to separate the main effects and interaction of two or
more factors use `anova_glm()` ([walkthrough](glm.html)). The
[decision guide](choosing.html) covers the choice in more detail.

## The data

`PlantGrowth` records the dried weight of 30 plants, ten grown under control
conditions and ten under each of two treatments.

```{r data}
head(PlantGrowth)
table(PlantGrowth$group)
sds <- tapply(PlantGrowth$weight, PlantGrowth$group, sd)
round(cbind(mean = tapply(PlantGrowth$weight, PlantGrowth$group, mean),
            sd = sds), 3)
```

The question is whether the treatments change the mean weight. The standard
deviations are not the same: `trt1` varies about
`r f2(sds[["trt1"]] / sds[["trt2"]])` times as much as `trt2`, so its variance
is about `r f2((sds[["trt1"]] / sds[["trt2"]])^2)` times as large. With
groups this small you cannot tell whether that is a real difference in spread
or sampling noise, and that uncertainty is exactly the case for a test that
does not need to know.

## Fitting the model

Columns are named as character strings: the response, then one or more grouping
columns.

```{r fit}
fit <- anova_welch(PlantGrowth, "weight", "group")
fit
```

Reading the printout from the top:

- The first line names the method, and `Call:` records how the fit was made.
- `Observations used: 30` is the number of rows analysed. Had any rows been
  dropped for a missing or infinite value, the count would appear here with a
  pointer to `$notes`.
- `Omnibus test` is the Welch test itself, described in the next section.
- There is no `Notes` block, because the function had nothing to report for
  these data. [What the notes say](#what-the-notes-say) below explains when it
  does.
- `Plots available` lists the three plots stored in `$plots`.

`summary(fit)` prints the same header followed by every table the fit holds:
the assumption checks, effect sizes, group summaries and pairwise comparisons.
The sections below take those one at a time.

## The omnibus test

```{r anova}
fit$anova
```

```{r df-equal, include = FALSE}
# Welch's denominator df for these group sizes when the sample variances are
# identical: rescale each group to unit SD about its own mean.
pg <- PlantGrowth
g_m <- ave(pg$weight, pg$group)
g_s <- ave(pg$weight, pg$group, FUN = sd)
pg$equal_sd <- g_m + (pg$weight - g_m) / g_s
df_equal <- unname(oneway.test(equal_sd ~ group, data = pg)$parameter[2])
```

- `term` is the grouping column (or a label for the combined cells when there
  are several; see [below](#several-grouping-columns)).
- `statistic` is Welch's F. It is built from each group's own variance, with
  each group's mean weighted by its precision (its size over its variance), so
  a noisy group counts for less.
- `num_df` is the number of groups minus one.
- `den_df` is Welch's approximate denominator degrees of freedom, estimated
  from the group sizes and variances. It is not on the same footing as the
  classical test's 30 - 3 = 27. With three groups of ten it would be
  `r round(df_equal, 2)` even if the three sample variances were identical,
  and it falls as they diverge: here it is `r f2(fit$anova$den_df)`.
- `p_value` is the probability of an F at least this large if all the group
  means were equal.

The test is `stats::oneway.test()` with `var.equal = FALSE`, and the `htest`
object it returns is kept in `$model`. There is no fitted model, so
`$emmeans_object` is `NULL`.

```{r model}
fit$model
oneway.test(weight ~ group, data = PlantGrowth, var.equal = TRUE)
```

The classical F test, shown second, happens to reach a similar conclusion here
(p = `r f3(oneway.test(weight ~ group, PlantGrowth, var.equal = TRUE)$p.value)`
against `r f3(fit$anova$p_value)`). That will not always be so.

## Why not the classical F test?

`InsectSprays` counts the insects left on plots treated with six sprays, twelve
plots per spray. The spreads differ a great deal:

```{r insects-data}
round(tapply(InsectSprays$count, InsectSprays$spray, sd), 2)
ins <- anova_welch(InsectSprays, "count", "spray")
ins$anova
ins$notes
```

The note fires whenever the largest group variance is more than four times the
smallest. Here it is `r f2(ins$assumptions$variance_ratio)` times. The
difference between the sprays is so large that both tests reject
overwhelmingly, so these data cannot show what goes wrong. For that you need a
case where the truth is known.

The simulation below draws data in which every group has the **same** mean, so
any rejection is a false positive, and in which the smallest group is the
noisiest. It also includes two equal-variance settings, one with no difference
and one with a real one, to show what Welch's test costs when its protection is
not needed. `anova_welch()`'s omnibus test is `oneway.test()`, so the
simulation calls that directly.

```{r simulation}
set.seed(2024)
reject_rate <- function(n, sd, mu = rep(0, length(n)), reps = 2000) {
  g <- factor(rep(seq_along(n), n))
  p <- replicate(reps, {
    y <- rnorm(sum(n), mean = rep(mu, n), sd = rep(sd, n))
    c(classical = oneway.test(y ~ g, var.equal = TRUE)$p.value,
      welch     = oneway.test(y ~ g)$p.value)
  })
  rowMeans(p < 0.05)
}
sim <- rbind(
  "equal SDs, no difference"   = reject_rate(c(10, 10, 10), c(1, 1, 1)),
  "equal SDs, means 0, 0.5, 1" = reject_rate(c(10, 10, 10), c(1, 1, 1),
                                             mu = c(0, 0.5, 1)),
  "SDs 4/2/1, n 10/20/30, no difference" =
    reject_rate(c(10, 20, 30), c(4, 2, 1))
)
sim
```

Each rate is estimated from 2,000 data sets, so a rate near 5% is accurate to
about half a percentage point, and one near 45% to about one point. At a
nominal 5% level:

- With equal variances and no difference, both tests reject about 5% of the
  time (`r pct(sim[1, "classical"])` classical, `r pct(sim[1, "welch"])`
  Welch).
- With equal variances and a real difference, Welch's test is only a little
  less powerful (`r pct(sim[2, "welch"])` against
  `r pct(sim[2, "classical"])`).
- When the smallest group is the noisiest, the classical test rejects a true
  null in `r pct(sim[3, "classical"])` of data sets, about
  `r round(sim[3, "classical"] / 0.05)` times its nominal rate, while Welch's
  stays at `r pct(sim[3, "welch"])`.

The classical test pools the variances, so a small, noisy group is treated as
if it were as precise as the others, and its chance deviations look like real
differences. Unequal variances paired with unequal group sizes are common in
observational data, and you rarely know in advance that you are not in that
situation. That is why Welch's test is the default.

## Effect sizes

```{r effect-sizes}
print(fit$effect_sizes, digits = 3)
```

Each row is a standardised mean difference for one pair of groups, in the
same direction as the pairwise difference (`group1` minus `group2`).

**The standardiser.** Each difference in means is divided by
$s^* = \sqrt{(s_1^2 + s_2^2)/2}$, the square root of the average of the two
groups' variances, as the `standardiser` column records. The usual pooled
standard deviation assumes the two variances are equal: its meaning and its
sampling variance both depend on that. The average-variance standardiser does
not, so the effect size makes the same assumption as the test. You can check
the arithmetic for `ctrl` against `trt1`:

```{r d-by-hand}
m <- tapply(PlantGrowth$weight, PlantGrowth$group, mean)
s_star <- sqrt((sds[["ctrl"]]^2 + sds[["trt1"]]^2) / 2)
d_ct <- (m[["ctrl"]] - m[["trt1"]]) / s_star
d_ct
```

**The bias correction.** That value is Cohen's d on the average-variance
standardiser. By default (`hedges_correction = TRUE`) the estimate is multiplied
by Hedges' small-sample correction, evaluated at the Satterthwaite degrees of
freedom of the standardiser, and the column is called `hedges_g`. The
correction shrinks the estimate slightly towards zero, here from
`r f3(d_ct)` to `r f3(fit$effect_sizes$hedges_g[1])`.

**The interval.** `conf_low` and `conf_high` are an interval for the population
value $\delta^* = (\mu_1 - \mu_2) / \sqrt{(\sigma_1^2 + \sigma_2^2)/2}$. It is
found by inverting the noncentral t distribution of Welch's statistic on
Welch's degrees of freedom, then rescaling to the average-variance
standardiser. Because the interval is for $\delta^*$ itself, it is the same
whether or not the point estimate is bias-corrected. It is the interval that
`effectsize::cohens_d(pooled_sd = FALSE)` reports:

```{r effectsize-check}
two <- droplevels(subset(PlantGrowth, group %in% c("ctrl", "trt1")))
effectsize::cohens_d(weight ~ group, data = two, pooled_sd = FALSE)
```

**The labels.** `magnitude` applies Cohen's conventional thresholds of 0.2,
0.5 and 0.8 to the absolute value. The labels are a rough guide. Look at the
interval too: the `ctrl` versus `trt1` estimate is labelled "medium", but its
interval runs from `r f2(fit$effect_sizes$conf_low[1])` to
`r f2(fit$effect_sizes$conf_high[1])` and so is consistent with anything from a
small effect in the other direction to a large one.

## Group means and comparisons

No model is fitted, so `$emmeans` holds plain group summaries: each group's
size, mean, standard deviation and standard error, with a t interval for the
mean built from that group's own standard deviation.

```{r emmeans}
fit$emmeans
```

`$posthoc` holds one row per pair of groups:

```{r posthoc}
print(fit$posthoc, digits = 3)
```

This is **not** the Games-Howell procedure, although it is related. Both use,
for each pair, a standard error and degrees of freedom built from those two
groups alone. Games-Howell then refers the statistic to the studentised range
distribution to get simultaneous intervals. `anova_welch()` does something
simpler:

1. For each of the `choose(k, 2)` pairs it runs `stats::t.test()` with
   `var.equal = FALSE` on the two groups' data. That gives `difference`
   (`group1` minus `group2`), the interval `conf_low` to `conf_high`, the
   Welch t `statistic`, its own `df`, and the unadjusted two-sided `p_value`.
2. It then passes all the p-values to `stats::p.adjust()` with the method in
   `adjust` (Holm by default). The result is `p_adjusted`, and the method is
   recorded in `adjustment`.

As in every function in the package, `p_value` is unadjusted and `p_adjusted`
is adjusted. The intervals are **not** adjusted: each is an ordinary 95%
interval for that one difference. The third row is exactly the Welch t test of
`trt1` against `trt2`:

```{r ttest-check}
trts <- droplevels(subset(PlantGrowth, group != "ctrl"))
t.test(weight ~ group, data = trts)$p.value
```

Reading the table: `trt2` plants were heavier than `trt1` plants by
`r f3(-fit$posthoc$difference[3])`, and that difference survives the Holm
adjustment (adjusted p = `r f3(fit$posthoc$p_adjusted[3])`). The `ctrl` versus
`trt2` difference has an unadjusted p of `r f3(fit$posthoc$p_value[2])` but an
adjusted one of `r f3(fit$posthoc$p_adjusted[2])`: on its own it would pass at
5%, but not once you allow for having made three comparisons.

`"tukey"` is not an accepted `adjust` method here, as it is on the functions
whose comparisons come from **emmeans**. These are separate Welch t tests, not
linear contrasts on a shared error term:

```{r tukey-error, error = TRUE}
anova_welch(PlantGrowth, "weight", "group", adjust = "tukey")
```

## Checking assumptions

Welch's test drops the equal-variance assumption but keeps the others: the
observations are independent, and the mean of each group is approximately
normally distributed, which holds when each group's data are roughly normal or
the groups are reasonably large.

```{r assumptions}
fit$assumptions$normality
fit$assumptions$variance_ratio
```

**Normality within each group.** The Shapiro-Wilk test is run separately for
each group, not on pooled residuals. Pooling residuals from groups with
different spreads produces a mixture that can fail a normality test even when
every group is perfectly normal, which would argue against the very method
being used. A group with fewer than 3 or more than 5,000 observations is not
tested. Its row then has a reason in the `note` column, and `$notes` names it.
None of the three groups here gives cause for concern, but with ten
observations the test has little power, so read the Q-Q plot below as well.

**The variance ratio** is the largest group variance divided by the smallest.
It is information rather than a test: Welch's method does not need it to be
close to 1. Above 4, `$notes` points it out, as it did for `InsectSprays`.

If a group is clearly skewed or has outliers, the mean may not be the summary
you want. Compare ranks with `anova_kw()`, or model the response's
distribution with `anova_glm()` (a Gamma family for a positive, right-skewed
response, for example). The [diagnostics article](diagnostics.html) covers the
checks across the whole package.

## Plots

The fit holds three **ggplot2** objects. None is drawn until you ask for it.

```{r plot-means}
plot(fit, "means")
```

The `means` plot shows each group's mean as a bar, with its 95% interval from
`$emmeans`, and the group size under each label. Each interval is built from
its own group's standard deviation, so different widths reflect different
spreads (and sizes). The intervals here are for the group means, not for the
differences between them: two intervals that overlap do not mean the
difference is non-significant. Read the comparisons from `$posthoc`.

```{r plot-box}
plot(fit, "box")
```

The `box` plot shows the raw distribution of each group, ordered by median and
labelled with its size. This is where unequal spreads, skewness and outliers
show up. Here `trt1` is visibly more spread out than `trt2`, and two `trt1`
plants lie above its upper whisker.

```{r plot-qq}
plot(fit, "qq")
```

```{r qq-values, include = FALSE}
trt1 <- PlantGrowth$weight[PlantGrowth$group == "trt1"]
hi <- sort((trt1 - mean(trt1)) / sd(trt1), decreasing = TRUE)[1:2]
```

The `qq` plot is for normality. Each observation is expressed as its deviation
from its own group's mean, divided by its own group's standard deviation, so
every group contributes on its own scale, as Welch's test allows. If the groups
are roughly normal the points follow the dashed line. A bow away from the line
suggests skewness, and a few isolated points far from it suggest outliers.
Here the points follow the line closely through the middle. The largest few
sit a little above it, among them the two high `trt1` plants, at
`r f2(hi[2])` and `r f2(hi[1])` standard deviations above their group mean.
That is a hint of right skew, but departures of this size are common in 30
points drawn from a normal distribution.

Each plot is an ordinary ggplot object, so you can change it with `+`:

```{r plot-modify}
plot(fit, "box") +
  ggplot2::labs(title = "Dried weight by treatment", y = "Dried weight")
```

The [plots article](visuals.html) covers customisation in more detail.

## Options worth knowing

### `adjust`

`adjust` accepts any `stats::p.adjust()` method, spelled out in full: `"holm"`
(the default), `"hochberg"`, `"hommel"`, `"bonferroni"`, `"BH"`, `"BY"`,
`"fdr"` or `"none"`. Only `p_adjusted` changes. Holm controls the family-wise
error rate and is never less powerful than Bonferroni. Benjamini-Hochberg
(`"BH"`) controls the false discovery rate instead, which suits screening many
comparisons.

```{r adjust}
adj <- sapply(c("holm", "BH", "bonferroni", "none"), function(a)
  anova_welch(PlantGrowth, "weight", "group", adjust = a,
              plots = FALSE)$posthoc$p_adjusted)
cbind(fit$posthoc[, c("group1", "group2")], round(adj, 4))
```

### `conf_level`

`conf_level` sets the level of **every** interval the function returns: the
group means, the pairwise differences and the standardised effect sizes (and
the title of the `means` plot).

```{r conf-level}
f90 <- anova_welch(PlantGrowth, "weight", "group", conf_level = 0.90)
f90$emmeans[, c("group", "mean", "conf_low", "conf_high")]
print(f90$effect_sizes[, c("group1", "group2", "hedges_g", "conf_low",
                           "conf_high")], digits = 3)
```

Because the comparison intervals are per-comparison, one way to get
Bonferroni-simultaneous intervals for all three differences is to ask for
`conf_level = 1 - 0.05 / 3`. That widens the group-mean and effect-size
intervals too.

### `hedges_correction`

With `hedges_correction = FALSE` the effect size is the uncorrected
standardised difference, and the column is named `cohens_d`. The intervals are
unchanged, because they are for the population value either way.

```{r hedges}
d <- anova_welch(PlantGrowth, "weight", "group", hedges_correction = FALSE,
                 plots = FALSE)
print(d$effect_sizes[, c("group1", "group2", "cohens_d", "conf_low",
                         "conf_high")], digits = 3)
```

### `posthoc`

There are `choose(k, 2)` pairs, so with many groups you may want only the
omnibus test. The standardised differences are computed pair by pair alongside
the comparisons, so `posthoc = FALSE` empties `$effect_sizes` too, and `$notes`
says so:

```{r posthoc-false}
np <- anova_welch(PlantGrowth, "weight", "group", posthoc = FALSE,
                  plots = FALSE)
np$notes
is.null(np$effect_sizes)
```

### Several grouping columns

`warpbreaks` counts the warp breaks per loom for two types of wool at three
levels of tension. Given both columns, `anova_welch()` combines them into a
single factor of the level combinations that contain data (unused combinations
are dropped) and runs one Welch test across those cells:

```{r cells}
wb <- anova_welch(warpbreaks, "breaks", c("wool", "tension"))
wb$anova
wb$emmeans[, c("group", "n", "mean", "sd")]
wb$notes
```

The `term` reads `wool x tension cells`, and the cells are labelled `A : L`,
`B : L` and so on. The test asks whether the six cell means are all equal. It
cannot tell a wool effect from a tension effect, or test whether the effect of
tension depends on the wool, and the first note says so. For that factorial
question use `anova_glm()` with `interaction = TRUE` ([walkthrough](glm.html)).
The second note shows that the cell variances differ by a factor of more than
13, which is the situation Welch's test is built for.

### Two groups

With two groups the single pairwise comparison is the omnibus test: Welch's F
is the square of the Welch t, the p-values are identical, and no adjustment is
applied.

```{r two-groups}
f2g <- anova_welch(two, "weight", "group", plots = FALSE)
f2g$anova
f2g$posthoc[, c("statistic", "df", "p_value", "p_adjusted")]
f2g$notes
```

## What the notes say

```{r notes}
fit$notes
```

For the `PlantGrowth` fit, `$notes` is empty. No rows were dropped, every group
was large enough for its Shapiro-Wilk test, the variance ratio was below 4,
there was one grouping column, and there were more than two groups. Each of
those would otherwise have produced a note. You have seen several above: the
variance ratio (`InsectSprays`), combined cells (`warpbreaks`), two groups, and
`posthoc = FALSE`. The function also writes a note if you ask for a low
`conf_level` with very small groups and the bias-corrected estimate falls
outside its own interval. Read `$notes` on every fit, including when you expect
it to be empty.

## Reporting the result

```{r report-values, include = FALSE}
a <- fit$anova
ph <- fit$posthoc
es <- fit$effect_sizes
```

A write-up built from the fit with inline R, so that the numbers cannot drift
from the analysis:

> Dried plant weight differed between the three conditions (Welch's
> F(`r a$num_df`, `r f2(a$den_df)`) = `r f2(a$statistic)`,
> p `r fp(a$p_value)`). In pairwise Welch t tests with Holm's adjustment,
> plants given treatment 2 were heavier than those given treatment 1 by
> `r f3(-ph$difference[3])` (95% CI `r f3(-ph$conf_high[3])` to
> `r f3(-ph$conf_low[3])`; Hedges' g on the average-variance standard
> deviation = `r f2(-es$hedges_g[3])`, 95% CI `r f2(-es$conf_high[3])` to
> `r f2(-es$conf_low[3])`; adjusted p `r fp(ph$p_adjusted[3])`). Neither
> treatment differed clearly from the control after adjustment (treatment 1:
> adjusted p `r fp(ph$p_adjusted[1])`; treatment 2: adjusted
> p `r fp(ph$p_adjusted[2])`).

The signs are flipped from the table so that the sentence reads
"treatment 2 minus treatment 1". State which standardiser you used: a
Hedges' g on the pooled standard deviation is a different quantity.

## See also

- `anova_kw()` and the [Kruskal-Wallis walkthrough](kruskal-wallis.html) for a
  rank-based comparison.
- `anova_glm()` and the [analysis-of-deviance walkthrough](glm.html) for
  factorial designs.
- `anova_ancova()` and the [ANCOVA walkthrough](ancova.html) to adjust for a
  covariate.
- [Working with results](results.html) for the structure of the returned
  object and more on reporting.
- [Diagnostics across the package](diagnostics.html) and
  [Plots and visual customisation](visuals.html).
- The [Get Started vignette](anovakit.html) for a tour of every function.
